---
title: "Charter City"
author: "Andrew Martin"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Existing Charter City Data

```{r}

library(rprojroot)

root <- rprojroot::find_rstudio_root_file()

```



[OLD]

old_ch_city <- readr::read_csv(
  file = file.path(root, 'data-raw', 'charter_city.csv')
)

[OLD] Force zero pad out to 4 digits on charter_city_code

old_ch_city$school_district_code <- sprintf("%04d", old_ch_city$school_district_code)
old_ch_city$charter_city_code <- sprintf("%04d", old_ch_city$charter_city_code)

head(old_ch_city)


## Add new schools

[OLD 2017]

new_fall_2017 <- tibble(
  school = c(
    "PATERSON ARTS AND SCIENCE CS",                      
    "CAMDENS PROMISE CHARTER SCHOOL",             
    "TRENTON STEM-TO-CIVICS CS",                      
    "GREAT FUTURES CHARTER SCHOOL",                         
    "ATLANTIC CITY COMMUNITY CHARTER SCHOOL",               
    "COLLEGE ACHIEVE CENTRAL CHARTER SCHOOL",              
    "INTERNATIONAL ACADEMY OF ATLANTIC CITY CHARTER SCHOOL",
    "HUDSON ARTS AND SCIENCE  CHARTER SCHOOL",      
    "INTERNATIONAL ACADEMY OF TRENTON CHARTER SCHOOL",
    "LINK COMMUNITY CHARTER SCHOOL"
  ),
  school_district_code = c(
    "6096", "6107", "6183", "6184", "6060", "6101", "6104", "6105", "6182", "6099"
  ),
  charter_city = c(
    'Paterson', 'Camden', 'Trenton', 'Jersey City', 'Atlantic City',
    'Plainfield', 'Atlantic City', 'Kearny', 'Trenton', 'Newark'
  ),
  charter_city_code = c(
    '4010', '0680', '5210', '2390', '0110', 
    '4160', '0110', '2410', '5210', '3570'
  )
)

2018

```{r}

more_new_2018 <- tibble(
  district_name = c(
    'Lead Charter School',
    'Achieve Community Charter School',
    'College Achieve Greater Asbury Park Charter School',
    'College Achieve Paterson Charter School',
    'Ocean Academy Charter School'
  ),
  district_id = c(
    '6109',
    '6110',
    '7891',
    '7892',
    '7893'
  ),
  host_district_name = c(
    'Newark',
    'Newark',
    'Asbury Park',
    'Paterson',
    'Lakewood'
  ),
  host_district_id = c(
    '3570',
    '3570',
    '0100',
    '4010',
    '2520'
  )
)

```


Put together

```{r}

charter_city_old_new <- dplyr::bind_rows(
  charter_city,
  more_new_2018
) %>% 
arrange(host_district_name)

head(charter_city_old_new) %>% print.AsIs()
```

[DEPRECATED] Make a numeric column

charter_city_old_new <- charter_city_old_new %>%
  mutate(
    charter_city_code_numeric = as.integer(charter_city_code)
  ) %>% 
  arrange(charter_city, school)


Check for equality

```{r}

identical(
  names(charter_city),
  names(charter_city_old_new)
)

identical(
  map_chr(charter_city, class) %>% unname(),
  map_chr(charter_city_old_new, class) %>% unname()
)
```

Save to rda and data-raw

```{r}

readr::write_csv(
  x = charter_city_old_new,
  path = file.path(root, 'data-raw', 'charter_city.csv')
)

```

```{r}

charter_city <- charter_city_old_new

save(charter_city, file = file.path(root, 'data', 'charter_city.rda'))

```

