---
title: "Charter City"
author: "Andrew Martin"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Existing Charter City Data

```{r}

library(rprojroot)

root <- rprojroot::find_rstudio_root_file()

```



[OLD]

old_ch_city <- readr::read_csv(
  file = file.path(root, 'data-raw', 'charter_city.csv')
)

[OLD] Force zero pad out to 4 digits on charter_city_code

old_ch_city$school_district_code <- sprintf("%04d", old_ch_city$school_district_code)
old_ch_city$charter_city_code <- sprintf("%04d", old_ch_city$charter_city_code)

head(old_ch_city)


## Add new schools

[OLD 2017]

new_fall_2017 <- tibble(
  school = c(
    "PATERSON ARTS AND SCIENCE CS",                      
    "CAMDENS PROMISE CHARTER SCHOOL",             
    "TRENTON STEM-TO-CIVICS CS",                      
    "GREAT FUTURES CHARTER SCHOOL",                         
    "ATLANTIC CITY COMMUNITY CHARTER SCHOOL",               
    "COLLEGE ACHIEVE CENTRAL CHARTER SCHOOL",              
    "INTERNATIONAL ACADEMY OF ATLANTIC CITY CHARTER SCHOOL",
    "HUDSON ARTS AND SCIENCE  CHARTER SCHOOL",      
    "INTERNATIONAL ACADEMY OF TRENTON CHARTER SCHOOL",
    "LINK COMMUNITY CHARTER SCHOOL"
  ),
  school_district_code = c(
    "6096", "6107", "6183", "6184", "6060", "6101", "6104", "6105", "6182", "6099"
  ),
  charter_city = c(
    'Paterson', 'Camden', 'Trenton', 'Jersey City', 'Atlantic City',
    'Plainfield', 'Atlantic City', 'Kearny', 'Trenton', 'Newark'
  ),
  charter_city_code = c(
    '4010', '0680', '5210', '2390', '0110', 
    '4160', '0110', '2410', '5210', '3570'
  )
)

2018


more_new_2018 <- tibble(
  district_name = c(
    'Lead Charter School',
    'Achieve Community Charter School',
    'College Achieve Greater Asbury Park Charter School',
    'College Achieve Paterson Charter School',
    'Ocean Academy Charter School'
  ),
  district_id = c(
    '6109',
    '6110',
    '7891',
    '7892',
    '7893'
  ),
  # these are now needed
  host_county_id = c(
    
  ),
  host_county_name = c(
    
  ),
  host_district_name = c(
    'Newark',
    'Newark',
    'Asbury Park',
    'Paterson',
    'Lakewood'
  ),
  host_district_id = c(
    '3570',
    '3570',
    '0100',
    '4010',
    '2520'
  )
)



historic

   district_id county_name             district_name
1                
2                     
3                                  
4
5                          
6         
7                
8                   
9         
10        
11               
12        
13                  
14        
15        
16                            
17        
18       
19        
20                    
21                     
22       
23                
24        
25                 
26       
27                  
28        
29        
30        
31                       
32              
33              
34              
35                       
36               
37               
38        
39                        
40                    
41              MERCER TRENTON COMMUNITY CHARTER
42        7940      Mercer      Trenton Community CS
43        7940      MERCER      TRENTON COMMUNITY CS

```{r}

historic_pre2010 <- tibble(
  district_name = c(
    'CAPITAL PREP CHARTER SCHO',
    'LEARNING CENTER CS',
    'CALLA CS',
    'C.R.E.A.T.E. CS',
    'FAMILY ALLIANCE CS',
    
    'FRANKLIN CS',
    'GALLOWAY CHARTER SCHOOL',
    'GATEWAY CHARTER SCHOOL',
    'GRANVILLE CS',
    'GRANVILLE CHARTER SCHOOL',
    
    'GRANVILLE CHARTER HS',
    'Granville Middle CS',
    'GREATER TRENTON AREA AC.',
    'Greenville Community CS',
    'Alexander Hamilton CS',
    
    'LEAP ACADEMY CS',
    'PATERSON CS FOR URBAN LEA',
    'PLEASANTVILLE CS FOR AC.',
    'RUSSELL ACADEMY CS',
    'SAMUEL DEWITT PROCTOR ACA',
    
    'Simon Bolivar CS',
    'TRENTON COMMUNITY CS',
    'TRENTON COMMUNITY CS'
  ),
  district_id = c(
    '6023',
    '6040',
    '6210',
    '6220',
    '6510',
    
    '6520',
    '6610',
    '6615',
    '6629',
    '6630',
    
    '6631',
    '6632',
    '6645',
    '6660',
    '6700',
    
    '7110',
    '7505',
    '7525',
    '7750',
    '7815',
    
    '7825',
    '7901',
    '7940'
  ),

  host_county_id = c(
    '21',
    '01',
    '39',
    '17',
    '05',
    
    '35', 
    '01',
    '17',
    '21',
    '21',
    
    '21',
    '21',
    '21',
    '17',
    '31',
    
    '07',
    '31',
    '01',
    '13',
    '21',
    
    '13',
    '21',
    '21'
  ),
  host_county_name = c(
    'Mercer',
    'Atlantic',
    'Union',
    'Hudson',
    'Burlington',
    
    'Somerset',
    'Atlantic',
    'Hudson',
    'Mercer',
    'Mercer',
    
    'Mercer',
    'Mercer',
    'Mercer',
    'Hudson',
    'Passaic',
    
    'Camden',
    'Passaic',
    'Atlantic',
    'Essex',
    'Mercer',
    
    'Essex',
    'Mercer',
    'Mercer'
  ),
  
  host_district_name = c(
    'Trenton',
    'Atlantic City',
    'Plainfield',
    'Jersey City',
    'Willingboro',
    
    'Franklin Township',
    'Galloway Township',
    'Jersey City',
    'Trenton',
    'Trenton',
    
    'Trenton',
    'Trenton',
    'Trenton',
    'Jersey City',
    'Paterson',
    
    'Camden', # LEAP Academy CS
    'Paterson',
    'Pleasantville', # PLEASANTVILLE CS FOR AC.
    'East Orange', # RUSSELL ACADEMY CS
    'Ewing Township', # SAMUEL DEWITT PROCTOR ACA,
    
    'Newark', # Simon Bolivar CS
    'Trenton',
    'Trenton'
  ),
  host_district_id = c(
    '5210',
    '0110',
    '4160',
    '2390',
    '5805',
    
    '1590', 
    '1690',
    '2390',
    '5210', 
    '5210',
    
    '5210',
    '5210',
    '5210',
    '2390',
    '4010',
    
    
    '0680', # LEAP Academy CS
    '4010',
    '4180', # PLEASANTVILLE CS FOR AC.
    '1210', # RUSSELL ACADEMY CS
    '1430', # SAMUEL DEWITT PROCTOR ACA
    
    '3570', # Simon Bolivar CS
    '5210',
    '5210'
  )
)


```

Put together

```{r}

charter_city_old_new <- dplyr::bind_rows(
  charter_city,
  historic_pre2010
) %>% 
arrange(host_district_name)

head(charter_city_old_new) %>% print.AsIs()
```

[DEPRECATED] Make a numeric column

charter_city_old_new <- charter_city_old_new %>%
  mutate(
    charter_city_code_numeric = as.integer(charter_city_code)
  ) %>% 
  arrange(charter_city, school)


Check for equality

```{r}

identical(
  names(charter_city),
  names(charter_city_old_new)
)

identical(
  map_chr(charter_city, class) %>% unname(),
  map_chr(charter_city_old_new, class) %>% unname()
)

identical(
  nrow(charter_city) + nrow(historic_pre2010),
  nrow(charter_city_old_new)
)
```

Save to rda and data-raw

```{r}

readr::write_csv(
  x = charter_city_old_new,
  path = file.path(root, 'data-raw', 'charter_city.csv')
)

```

```{r}

charter_city <- charter_city_old_new

save(charter_city, file = file.path(root, 'data', 'charter_city.rda'))

```

# one time fixes

Put host county onto charter city

```{r eval=FALSE}
county_join_df <- enr_2018[, c('district_id', 'county_id', 'county_name')] %>%
  unique() %>%
  rename(host_district_id = district_id)

head(county_join_df)

charter_city <- charter_city %>%
  left_join(county_join_df, by = 'host_district_id') %>%
  rename(
    host_county_id = county_id,
    host_county_name = county_name
  ) %>%
  select(
    district_id,
    district_name,
    host_county_id,
    host_county_name,
    host_district_id,
    host_district_name
  )

```
