---
title: "Getting Started with njschooldata"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with njschooldata}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

## Introduction

The `njschooldata` package provides a simple interface for accessing New Jersey Department of Education (NJ DOE) school data in R. The NJ DOE publishes raw data files covering enrollment, assessments, graduation rates, and more - but these files use inconsistent formats across years, making longitudinal analysis challenging.

This package solves that problem by:

- Automatically downloading data from NJ DOE servers
- Parsing various file formats (fixed-width, Excel, CSV)
- Standardizing column names across years
- Providing both wide and tidy data formats

## Installation

Install the package from GitHub:

```{r installation, eval = FALSE}
# Using remotes (recommended)
remotes::install_github("almartin82/njschooldata")

# Or using devtools
devtools::install_github("almartin82/njschooldata")
```

Load the package:

```{r setup}
library(njschooldata)
library(dplyr)  # for data manipulation examples
```

## Understanding NJ Education Data

### The `end_year` Convention

All functions use `end_year` to specify the school year. This refers to the **spring semester year**:

| School Year | `end_year` Value |
|-------------|------------------|
| 2023-24 | 2024 |
| 2022-23 | 2023 |
| 2019-20 | 2020 |

### CDS Codes

New Jersey uses a County-District-School (CDS) identifier system:

- **county_id**: 2-digit county code (e.g., "13" = Essex County)
- **district_id**: 4-digit district code
- **school_id**: 3-digit school code ("999" indicates a district-level aggregate)

```{r cds-example}
# A complete CDS code example:
# County 13 (Essex), District 3570 (Newark), School 050 (specific school)
```

### Aggregation Levels

Data is provided at multiple aggregation levels, identified by boolean flags:

| Level | `is_state` | `is_district` | `is_school` | Description |
|-------|------------|---------------|-------------|-------------|
| State | TRUE | FALSE | FALSE | Statewide totals |
| District | FALSE | TRUE | FALSE | District totals (school_id = "999") |
| School | FALSE | FALSE | TRUE | Individual school data |

## Quick Start: Enrollment Data

### Fetching Basic Enrollment

```{r enrollment-basic}
# Get 2024 enrollment data (2023-24 school year)
enr_2024 <- fetch_enr(2024)

# View the structure
glimpse(enr_2024)

# Check dimensions
dim(enr_2024)  # Rows and columns
```

### Wide vs. Tidy Format

The `tidy` parameter transforms data for easier analysis:
```{r enrollment-tidy}
# Wide format (default) - one row per school, many demographic columns
enr_wide <- fetch_enr(2024, tidy = FALSE)

# Tidy format - one row per school-subgroup combination
enr_tidy <- fetch_enr(2024, tidy = TRUE)
```

The tidy format is better for:
- Comparing subgroups within schools
- Longitudinal analysis across years
- Filtering to specific demographics

### Filtering to Specific Levels

```{r enrollment-filtering}
# Get only district-level totals
district_totals <- enr_tidy %>%
  filter(is_district, subgroup == "total_enrollment")

# Get only school-level data
school_data <- enr_tidy %>%
  filter(is_school)

# Get state totals
state_totals <- enr_tidy %>%
  filter(is_state)
```

## Assessment Data

### NJSLA / PARCC (2015-present)

Use `fetch_parcc()` for both NJSLA (2019+) and PARCC (2015-2018) data:

```{r assessment}
# Get 2024 Grade 4 Math results
math_g4_2024 <- fetch_parcc(
  end_year = 2024,
  grade_or_subj = 4,
  subj = "math"
)

# Get 2024 Grade 8 ELA results
ela_g8_2024 <- fetch_parcc(
  end_year = 2024,
  grade_or_subj = 8,
  subj = "ela"
)

# Get Algebra 1 results (high school)
alg1_2024 <- fetch_parcc(
  end_year = 2024,
  grade_or_subj = "ALG1",
  subj = "math"
)
```

**Note**: 2020 assessments were cancelled due to COVID-19.

### Available Grade Levels

| Subject | PARCC (2015-2018) | NJSLA (2019+) |
|---------|-------------------|---------------|
| ELA | Grades 3-11 | Grades 3-10 |
| Math | Grades 3-8 | Grades 3-8 |
| Math Courses | ALG1, GEO, ALG2 | ALG1, GEO, ALG2 |

### Legacy Assessments (2004-2014)

For historical data, use `fetch_old_nj_assess()`:

```{r legacy-assessment}
# Get 2010 Grade 5 NJASK results
njask_2010 <- fetch_old_nj_assess(
  end_year = 2010,
  grade = 5,
  tidy = TRUE
)
```

## Graduation Data

### Graduation Rates

```{r graduation-rates}
# Get 4-year graduation rates for 2024
grad_rate_2024 <- fetch_grad_rate(
  end_year = 2024,
  methodology = "4 year"
)

# Get 5-year graduation rates (available 2012+)
grad_rate_5yr <- fetch_grad_rate(
  end_year = 2024,
  methodology = "5 year"
)
```

### Graduation Counts

```{r graduation-counts}
# Get graduation counts
grad_count_2024 <- fetch_grad_count(end_year = 2024)
```

## School and District Directories

Get metadata about schools and districts:

```{r directories}
# Current school directory with addresses, coordinates, grades served
schools <- get_school_directory()

# Current district directory
districts <- get_district_directory()

# View available columns
names(schools)
```

## Data Coverage Summary

| Data Type | Function | Years Available |
|-----------|----------|-----------------|
| Enrollment | `fetch_enr()` | 1999-2025 |
| NJSLA/PARCC | `fetch_parcc()` | 2015-2024 |
| NJASK | `fetch_old_nj_assess()` | 2004-2014 |
| Graduation Rates | `fetch_grad_rate()` | 2011-2024 |
| Graduation Counts | `fetch_grad_count()` | 2012-2024 |

## Common Subgroups

When working with tidy data, you'll encounter these subgroup codes:

### Race/Ethnicity
- `white`, `black`, `hispanic`, `asian`
- `pacific_islander`, `american_indian`, `other`

### Other Demographics
- `male`, `female`
- `ed` (economically disadvantaged)
- `lep_current` (current English learners)
- `special_education`
- `total_enrollment` or `total_population`

## Tips and Best Practices

### 1. Cache Data Locally

Data is downloaded fresh each call. For repeated analysis, save locally:

```{r caching}
# Download once
enr_2024 <- fetch_enr(2024, tidy = TRUE)

# Save for reuse
saveRDS(enr_2024, "data/enr_2024.rds")

# Load later without re-downloading
enr_2024 <- readRDS("data/enr_2024.rds")
```

### 2. Handle Suppressed Data

Small cell sizes are suppressed with `*` or `NA`:

```{r suppression}
# Filter out suppressed values before calculations
reliable_data <- enr_tidy %>%
  filter(!is.na(n_students), n_students >= 10)
```

### 3. Multi-Year Analysis

Use `purrr::map_df()` for combining multiple years:

```{r multi-year}
library(purrr)

# Fetch 5 years of enrollment data
years <- 2020:2024
multi_year_enr <- map_df(years, ~fetch_enr(.x, tidy = TRUE))

# Now you can analyze trends
enrollment_trends <- multi_year_enr %>%
  filter(is_state, subgroup == "total_enrollment") %>%
  select(end_year, n_students)
```

## Next Steps

- See `vignette("longitudinal-analysis")` for advanced multi-year analysis
- See `vignette("data-dictionary")` for complete column definitions
- Visit the [package website](https://almartin82.github.io/njschooldata/) for full documentation

## Getting Help

- File issues: https://github.com/almartin82/njschooldata/issues
- Email: almartin@gmail.com
