---
title: "Newark on the Efficiency Frontier: Do the Dollars Pay Off?"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Newark on the Efficiency Frontier: Do the Dollars Pay Off?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 6,
  out.width = "100%",
  message = FALSE,
  warning = FALSE,
  # Default OFF: only the expensive fetch chunk caches. Plot chunks must
  # re-evaluate every build or knitr replays stale figures (caschooldata, 2026-03).
  cache = FALSE
)
```

```{r load-packages}
library(njschooldata)
library(ggplot2)
library(dplyr)
library(tidyr)
library(scales)

options(timeout = max(600, getOption("timeout")))
```

```{r theme}
theme_nj <- function() {
  theme_minimal(base_size = 14) +
    theme(
      plot.title = element_text(face = "bold", size = 16),
      plot.subtitle = element_text(color = "gray40"),
      plot.caption = element_text(color = "gray55", size = 9),
      panel.grid.minor = element_blank(),
      legend.position = "bottom"
    )
}

nwk_navy <- "#16356B"
nwk_gold <- "#C8A14B"
nwk_teal <- "#16A085"
nwk_red  <- "#B83227"
peer_gray <- "gray70"

NEWARK <- "3570"
```

The companion brief, [*What Did Newark's Gains Cost?*](newark-fiscal-brief.html),
benchmarks Newark against District Factor Group A one indicator at a time. This
one asks the harder, cross-district question
[MarGrady Research](https://margrady.com/movingup/) left open: outcomes rose, but
were the dollars *efficient*? Answering that needs tools that reason across
districts at once, not one rank at a time. This brief uses the comparative layer
of `njschooldata` built for exactly that:

- `tges_find_peers()` builds Newark's *data-driven* peer set instead of assuming DFG.
- `tges_frontier()` scores every district against the spend-versus-outcome efficiency frontier.
- `tges_gap_cost()` turns a peer gap into real dollars.
- `tges_convergence()` asks whether DFG A spending is converging.
- `tges_composition_drift()` tracks where the budget moved.
- `tges_volatility()` flags fragile funding.
- `tges_compare()` lines the big cities up side by side.

Everything rides on the
[Taxpayers' Guide to Educational Spending](https://www.nj.gov/education/guide/)
(TGES) joined to NJ DOE graduation outcomes. TGES is district-level, so this is a
screen that says where to look, not a verdict.

```{r fetch-data, cache = TRUE}
# One multi-year object drives the brief; the 2024 guide is the latest snapshot.
tgm  <- fetch_many_tges(2019:2024)
tg24 <- tgm[["2024"]]

# Graduation outcomes for the efficiency frontier: 4-year rate, district level,
# total cohort, ranked within DFG A. district_id is the 4-digit code TGES uses.
grate23 <- fetch_grad_rate(2023, methodology = "4 year") %>%
  add_dfg() %>%
  filter(dfg == "A", is_district, subgroup == "total") %>%
  grate_percentile_rank(peer_type = "dfg")
```

## 1. Who are Newark's real peers?

DFG A is a 1990s census construct of 37 districts. But which ones are
*structurally* like Newark today, on enrollment, spending, and how the budget is
funded? `tges_find_peers()` standardizes those features and returns the nearest
districts by distance. The answer is not all 37 DFG A districts, it is the big
urban systems specifically.

```{r peers-compute}
peers <- tges_find_peers(tg24, NEWARK, n = 8)
stopifnot(nrow(peers) > 1)

peers %>%
  select(district_name, dfg, distance, ade, budgetary_pp, local_share)
```

```{r peers-chart}
peer_plot <- peers %>%
  filter(!is_focal) %>%
  mutate(district_name = reorder(district_name, -distance))

ggplot(peer_plot, aes(x = distance, y = district_name)) +
  geom_segment(aes(xend = 0, yend = district_name), color = peer_gray, linewidth = 1) +
  geom_point(color = nwk_navy, size = 4) +
  labs(
    title = "Newark's data-driven fiscal twins are the urban Abbotts",
    subtitle = "Nearest districts to Newark by enrollment, per-pupil cost, composition, and revenue mix",
    x = "Distance from Newark (smaller = more similar)", y = NULL,
    caption = "Source: NJ DOE TGES 2024 via njschooldata::tges_find_peers()"
  ) +
  theme_nj()
```

Paterson, Trenton, Passaic, Elizabeth, and Union City sit closest, every one a
DFG A city. The peer set the data picks is exactly the set a Newark analyst would
defend by hand, which is the point: now it is reproducible.

## 2. The efficiency frontier

Here is the question MarGrady could not answer. `tges_frontier()` plots per-pupil
spend against the graduation-rate percentile within DFG A, then scores each
district against the *free-disposal-hull frontier*: the set of districts that
hit at least a given outcome for the least money. A score of 1.0 means a district
is on the frontier; 0.81 means another district reaches its outcome for 81% of
its spend.

```{r frontier-compute}
spend23 <- tg24$CSG1 %>% filter(calc_type == "Actuals", end_year == 2023)

frontier <- tges_frontier(spend23, grate23,
                          outcome_col = "grad_rate_percentile", peer = "dfg")
stopifnot(nrow(frontier) > 0)

frontier %>%
  arrange(efficiency_score) %>%
  select(district_name, spend, outcome, efficiency_score,
         reference_district_name, excess_spend) %>%
  head(8)
```

```{r frontier-chart}
fr_plot <- frontier %>%
  mutate(is_newark = district_id == NEWARK)

newark_row <- fr_plot %>% filter(is_newark)

ggplot(fr_plot, aes(x = spend, y = outcome)) +
  geom_point(aes(color = on_frontier), size = 3, alpha = 0.85) +
  geom_point(data = newark_row, color = nwk_navy, size = 5) +
  ggrepel::geom_text_repel(
    data = bind_rows(newark_row, filter(fr_plot, on_frontier)),
    aes(label = district_name), size = 3.4, color = "gray25", max.overlaps = 20
  ) +
  scale_color_manual(values = c(`TRUE` = nwk_gold, `FALSE` = peer_gray),
                     labels = c(`TRUE` = "On the frontier", `FALSE` = "Behind the frontier"),
                     name = NULL) +
  scale_x_continuous(labels = label_dollar()) +
  labs(
    title = sprintf("Newark sits at %.2f efficiency within DFG A",
                    newark_row$efficiency_score),
    subtitle = "Budgetary per-pupil spend (2023 actuals) vs. 4-year graduation percentile",
    x = "Per-pupil budgetary cost", y = "Graduation-rate percentile (within DFG A)",
    caption = "Source: NJ DOE TGES + graduation rates via njschooldata::tges_frontier()"
  ) +
  theme_nj()
```

Newark lands at `r sprintf("%.2f", newark_row$efficiency_score)`:
`r newark_row$reference_district_name` reaches at least Newark's graduation
percentile for `r scales::dollar(round(newark_row$excess_spend))` per pupil less.
Union City is the standout, posting near the top of DFG A on graduation while
spending below the middle. The districts in the lower right, spending the most
for middling outcomes, are where a cost-effectiveness review would start.

## 3. What would closing the classroom-spending gap cost?

Newark sends a smaller share of its budget to the classroom than the DFG A
median. `tges_gap_cost()` translates that share gap into dollars, both per pupil
and across Newark's enrollment.

```{r gapcost-compute}
gap <- tges_gap_cost(tg24, NEWARK, metric = "classroom_share",
                     target = "median", peer = "dfg")
stopifnot(nrow(gap) == 1)

gap %>%
  select(focal_value, target_value, gap, budgetary_pp,
         per_pupil_gap_dollars, ade, total_gap_dollars)
```

Newark routes `r scales::percent(gap$focal_value, accuracy = 0.1)` of its
budgetary per-pupil cost to the classroom against a DFG A median of
`r scales::percent(gap$target_value, accuracy = 0.1)`. Reaching that median would
take about `r scales::dollar(gap$per_pupil_gap_dollars)` per pupil, or roughly
`r scales::dollar(gap$total_gap_dollars)` district-wide at current enrollment.
That single number reframes the abstract share gap as a budget line.

## 4. Is DFG A converging on spending?

Are the highest-need districts drifting toward a common per-pupil cost, or are the
gaps widening? `tges_convergence()` regresses each district's spending growth on
its starting level. A negative slope (beta) means the lower-spending districts
grew faster, i.e. the group is converging.

```{r convergence-compute}
conv <- tges_convergence(tgm, peer = "dfg")
conv_a <- conv %>% filter(peer_group == "A")
stopifnot(nrow(conv_a) > 0)

conv %>%
  distinct(peer_group, beta, beta_pvalue, r_squared, converging, n_districts) %>%
  arrange(peer_group)
```

```{r convergence-chart}
beta_a <- conv_a$beta[1]
conv_a <- conv_a %>% mutate(is_newark = district_id == NEWARK)

ggplot(conv_a, aes(x = log_start_value, y = growth)) +
  geom_smooth(method = "lm", se = FALSE, color = nwk_red, linewidth = 1) +
  geom_point(color = peer_gray, size = 3) +
  geom_point(data = filter(conv_a, is_newark), color = nwk_navy, size = 5) +
  ggrepel::geom_text_repel(
    data = filter(conv_a, is_newark), aes(label = district_name),
    color = nwk_navy, size = 3.6
  ) +
  labs(
    title = sprintf("DFG A is converging on per-pupil spend (beta = %.3f)", beta_a),
    subtitle = "Annualized spending growth (2019-2024) vs. starting per-pupil cost; downward slope = convergence",
    x = "Log of 2019 budgetary per-pupil cost", y = "Annualized log growth",
    caption = "Source: NJ DOE TGES via njschooldata::tges_convergence()"
  ) +
  theme_nj()
```

The slope is negative and significant: lower-spending DFG A districts grew their
per-pupil cost faster than the high-spenders, so the band is compressing. Newark,
already near the middle of DFG A spending, grew about in line with where that
trend puts it.

## 5. Where did Newark's budget drift?

`tges_composition_drift()` measures how each spending share moved between two
years and ranks the move against peers. The classroom share is the one that
matters most for the efficiency story.

```{r drift-compute}
drift <- tges_composition_drift(tgm, peer = "dfg")
drift_a <- drift %>% filter(dfg == "A")
stopifnot(nrow(drift_a) > 0)

drift_a %>%
  arrange(classroom_share_drift) %>%
  select(district_name, classroom_share_start, classroom_share_end,
         classroom_share_drift, drift_percentile) %>%
  head(8)
```

```{r drift-chart}
drift_plot <- drift_a %>%
  arrange(classroom_share_drift) %>%
  slice(1:12) %>%
  mutate(district_name = reorder(district_name, classroom_share_drift),
         is_newark = district_id == NEWARK)

ggplot(drift_plot, aes(x = classroom_share_drift, y = district_name, fill = is_newark)) +
  geom_col() +
  scale_fill_manual(values = c(`TRUE` = nwk_navy, `FALSE` = peer_gray), guide = "none") +
  scale_x_continuous(labels = label_percent(accuracy = 1)) +
  labs(
    title = "The largest classroom-share declines in DFG A, 2019-2024",
    subtitle = "Change in the share of budgetary per-pupil cost reaching the classroom (Newark in navy)",
    x = "Classroom-share drift (negative = pulled away from the classroom)", y = NULL,
    caption = "Source: NJ DOE TGES via njschooldata::tges_composition_drift()"
  ) +
  theme_nj()
```

## 6. How fragile is the funding?

A district that funded recurring costs on a one-time federal bump is exposed when
the money expires. `tges_volatility()` ranks how much each district's federal
revenue share whipsawed across the ESSER years.

```{r volatility-compute}
vol <- tges_volatility(tgm, metric = "federal_share", peer = "dfg", min_years = 3)
vol_a <- vol %>% filter(dfg == "A")
stopifnot(nrow(vol_a) > 0)

vol_a %>%
  arrange(desc(vol_percentile)) %>%
  select(district_name, mean_value, cv, max_abs_yoy, vol_percentile) %>%
  head(8)
```

```{r volatility-chart}
vol_plot <- vol_a %>%
  arrange(desc(cv)) %>%
  slice(1:12) %>%
  mutate(district_name = reorder(district_name, cv),
         is_newark = district_id == NEWARK)

ggplot(vol_plot, aes(x = cv, y = district_name, fill = is_newark)) +
  geom_col() +
  scale_fill_manual(values = c(`TRUE` = nwk_navy, `FALSE` = peer_gray), guide = "none") +
  labs(
    title = "Federal-revenue volatility across DFG A, 2019-2024",
    subtitle = "Coefficient of variation of the federal share (higher = bumpier funding; Newark in navy)",
    x = "Coefficient of variation (federal revenue share)", y = NULL,
    caption = "Source: NJ DOE TGES via njschooldata::tges_volatility()"
  ) +
  theme_nj()
```

## 7. The counterfactual cities

Finally, `tges_compare()` lines the big urban systems up on the headline fiscal
metrics, so different reform strategies sit next to each other.

```{r compare-compute}
cities <- c(NEWARK, "0680", "5210", "4010", "2330")  # Newark, Camden, Trenton, Paterson, Jersey City
scorecard <- tges_compare(tg24, district_codes = cities)
stopifnot(nrow(scorecard) >= 1)

scorecard %>%
  select(district_name, total_pp, classroom_share, local_share,
         federal_share, student_admin_ratio, excess_surplus_flag)
```

```{r compare-chart}
sc_plot <- scorecard %>%
  select(district_name, classroom_share, administration_share) %>%
  pivot_longer(-district_name, names_to = "category", values_to = "share") %>%
  mutate(category = recode(category,
                           classroom_share = "Classroom",
                           administration_share = "Administration"))

ggplot(sc_plot, aes(x = share, y = reorder(district_name, share), fill = category)) +
  geom_col(position = "dodge") +
  scale_x_continuous(labels = label_percent(accuracy = 1)) +
  scale_fill_manual(values = c(Classroom = nwk_teal, Administration = nwk_gold), name = NULL) +
  labs(
    title = "Where the dollars go: five urban districts side by side",
    subtitle = "Classroom and administration as a share of budgetary per-pupil cost",
    x = "Share of budgetary per-pupil cost", y = NULL,
    caption = "Source: NJ DOE TGES via njschooldata::tges_compare()"
  ) +
  theme_nj()
```

The scorecard is the assembly point for the whole brief: per-pupil totals, the
classroom and administration shares, how much of each budget is local property
tax versus state aid, the administrative ratio, and the surplus flag, one row per
city. Different strategies, same peer tier, one table.

```{r session-info}
sessionInfo()
```
