---
title: "What Did Newark's Gains Cost? A DFG A Fiscal Brief"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{What Did Newark's Gains Cost? A DFG A Fiscal Brief}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 6,
  out.width = "100%",
  message = FALSE,
  warning = FALSE,
  # Default OFF: only the expensive fetch chunk caches. Plot chunks must
  # re-evaluate every build or knitr replays stale figures (caschooldata, 2026-03).
  cache = FALSE
)
```

```{r load-packages}
library(njschooldata)
library(ggplot2)
library(dplyr)
library(tidyr)
library(scales)

options(timeout = max(600, getOption("timeout")))
```

```{r theme}
theme_nj <- function() {
  theme_minimal(base_size = 14) +
    theme(
      plot.title = element_text(face = "bold", size = 16),
      plot.subtitle = element_text(color = "gray40"),
      plot.caption = element_text(color = "gray55", size = 9),
      panel.grid.minor = element_blank(),
      legend.position = "bottom"
    )
}

nwk_navy <- "#16356B"
nwk_gold <- "#C8A14B"
nwk_teal <- "#16A085"
nwk_red  <- "#B83227"
peer_gray <- "gray70"

NEWARK <- "3570"
```

[MarGrady Research's Newark study](https://margrady.com/movingup/) proved that
Newark's *outcomes* climbed against the right yardstick: not the state as a
whole, but District Factor Group A, the 37 highest-need communities in New
Jersey (Camden, Trenton, Paterson, and the like). Its own list of open questions
ends with the one it could not answer: *what was the per-pupil spending
trajectory, and was it cost-effective?*

This brief points the same peer-benchmarking machinery at dollars. Every chart
below compares Newark to its DFG A peers, using the comparative-fiscal toolkit in
`njschooldata` (`tges_revenue_mix()`, `tges_composition()`,
`tges_real_growth()`, `tges_federal_exposure()`, `tges_staffing()`,
`tges_red_flags()`), all built on the
[Taxpayers' Guide to Educational Spending](https://www.nj.gov/education/guide/)
(TGES). It is a companion, not a verdict: TGES is district-level, and a screen
that flags where to look is not the same as a finding.

```{r fetch-data, cache = TRUE}
# One multi-year object drives the whole brief. fetch_many_tges() returns a
# named list of years, each a list of tidy indicator tables. 2018-2025 spans a
# pre-pandemic baseline, the ESSER window, and enough actuals for growth math.
tgm <- fetch_many_tges(2018:2025)
```

```{r dfg-lookup}
# Attach DFG to any TGES frame by its 4-digit district_code (the same join
# tges_percentile_rank(peer = "dfg") uses internally). DFG A = highest need.
dfg_lk <- fetch_dfg() %>%
  group_by(district_code) %>%
  summarise(dfg = first(dfg), .groups = "drop")

with_dfg <- function(df) left_join(df, dfg_lk, by = "district_code")
```

## 1. Whose money is it? Newark runs on state aid, not property tax

For a suburban homeowner the school budget is mostly their own property tax. For
Newark it is mostly Trenton's. `tges_revenue_mix()` decomposes VITSTAT into both
the revenue shares and the per-pupil dollars each source buys.

```{r rev-compute}
rev <- tges_revenue_mix(tgm) %>% with_dfg()

# Newark's latest-year mix, in cents-on-the-dollar and per-pupil dollars
newark_rev_latest <- rev %>%
  filter(district_code == NEWARK, end_year == max(end_year)) %>%
  select(end_year, total_pp, local_share, state_share, federal_share,
         local_pp, state_pp, federal_pp)
stopifnot(nrow(newark_rev_latest) == 1)
newark_rev_latest

# How does Newark's local (property-tax) share compare to DFG A peers?
dfg_a_local <- rev %>%
  filter(dfg == "A", end_year == max(end_year)) %>%
  summarise(
    n = n(),
    newark_local = local_share[district_code == NEWARK][1],
    dfg_a_median_local = median(local_share, na.rm = TRUE)
  )
dfg_a_local
```

```{r rev-chart}
rev_long <- rev %>%
  filter(district_code == NEWARK) %>%
  select(end_year, Local = local_share, State = state_share,
         Federal = federal_share) %>%
  pivot_longer(-end_year, names_to = "source", values_to = "share")
stopifnot(nrow(rev_long) > 0)

ggplot(rev_long, aes(end_year, share, fill = source)) +
  geom_area(alpha = 0.9) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_fill_manual(values = c(Local = nwk_gold, State = nwk_navy,
                               Federal = nwk_red)) +
  labs(
    title = "Newark's school dollars are overwhelmingly state aid",
    subtitle = "Revenue mix, share of total. Local property tax is a thin slice.",
    x = NULL, y = "Share of revenue", fill = NULL,
    caption = "Source: NJ DOE Taxpayers' Guide to Educational Spending (VITSTAT)"
  ) +
  theme_nj()
```

The headline a taxpayer understands instantly: of Newark's roughly
\$`r format(round(newark_rev_latest$total_pp), big.mark = ",")` per pupil, only
\$`r format(round(newark_rev_latest$local_pp), big.mark = ",")` comes from local
property tax. The local share sits far below even its high-need DFG A peers,
which is the whole logic of *Abbott v. Burke*: the state, not the city, writes
the checks.

## 2. Newark spends near the top of DFG A, but a low share reaches the classroom

The most-cited efficiency proxy in school finance is "dollars to the classroom."
`tges_composition()` builds the classroom share; `tges_percentile_rank()` ranks
it inside DFG A.

```{r classroom-compute}
comp <- tges_composition(tgm, calc_type = "Budgeted")

# Rank classroom share within DFG A in the most recent budgeted year
latest_yr <- max(comp$end_year)
classroom_ranked <- comp %>%
  filter(end_year == latest_yr) %>%
  tges_percentile_rank("classroom_share", peer = "dfg") %>%
  filter(dfg == "A")
stopifnot(nrow(classroom_ranked) > 0)

newark_classroom <- classroom_ranked %>%
  filter(district_code == NEWARK) %>%
  select(end_year, classroom_share, peer_percentile, peer_n)
newark_classroom

# For contrast: where does Newark rank on total budgetary spend in DFG A?
spend_ranked <- tges_percentile_rank(
  comp %>% filter(end_year == latest_yr),
  "budgetary_pp", peer = "dfg"
) %>%
  filter(dfg == "A", district_code == NEWARK) %>%
  select(budgetary_pp, peer_percentile, peer_n)
spend_ranked
```

```{r classroom-chart}
plot_df <- classroom_ranked %>%
  mutate(is_newark = district_code == NEWARK) %>%
  arrange(classroom_share) %>%
  mutate(rank_order = row_number())
stopifnot(sum(plot_df$is_newark) == 1)

ggplot(plot_df, aes(reorder(district_name, classroom_share), classroom_share,
                    fill = is_newark)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_fill_manual(values = c(`TRUE` = nwk_red, `FALSE` = peer_gray),
                    guide = "none") +
  labs(
    title = "Cents on the dollar reaching the classroom, DFG A districts",
    subtitle = paste0("Newark in red. Classroom share = classroom instruction / ",
                      "budgetary per-pupil cost, ", latest_yr, " budget."),
    x = NULL, y = "Classroom share of budgetary per-pupil cost",
    caption = "Source: NJ DOE TGES (CSG1, CSG2)"
  ) +
  theme_nj()
```

Newark spends at the
`r round(spend_ranked$peer_percentile)`th percentile of DFG A on total budgetary
cost per pupil, but only the
`r round(newark_classroom$peer_percentile)`th percentile of that dollar reaches a
classroom (`r percent(newark_classroom$classroom_share, accuracy = 0.1)`). High
spend, low classroom share is exactly the pattern a board should be able to
explain.

## 3. Is spending exploding, or is enrollment shrinking?

Per-pupil cost rises mechanically when enrollment falls, because fixed costs
spread over fewer students. `tges_real_growth()` splits each year's per-pupil
growth into a real-cost component and an enrollment (denominator) component that
sum to the total.

```{r growth-compute}
rg <- tges_real_growth(tgm) %>%
  filter(district_code == NEWARK) %>%
  arrange(end_year)
stopifnot(nrow(rg) > 1)

rg %>%
  select(end_year, per_pupil, per_pupil_growth,
         real_cost_component, enrollment_component, enrollment_effect_share)
```

```{r growth-chart}
growth_long <- rg %>%
  filter(is.finite(real_cost_component)) %>%
  select(end_year,
         `Real cost growth` = real_cost_component,
         `Enrollment effect` = enrollment_component) %>%
  pivot_longer(-end_year, names_to = "component", values_to = "log_change")
stopifnot(nrow(growth_long) > 0)

ggplot(growth_long, aes(factor(end_year), log_change, fill = component)) +
  geom_col() +
  geom_hline(yintercept = 0, color = "gray40") +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_fill_manual(values = c(`Real cost growth` = nwk_navy,
                               `Enrollment effect` = nwk_gold)) +
  labs(
    title = "Decomposing Newark's per-pupil cost growth",
    subtitle = paste0("Bars sum to per-pupil growth. Gold above zero = cost per ",
                      "pupil rising because enrollment fell, not because Newark ",
                      "spent more."),
    x = NULL, y = "Contribution to per-pupil growth (log points)", fill = NULL,
    caption = "Source: NJ DOE TGES (CSG1AA: total expenditure and ADE)"
  ) +
  theme_nj()
```

In the pandemic-enrollment years the gold bars are large: a meaningful slice of
Newark's per-pupil increase is the denominator shrinking, not the budget growing.
Read per-pupil trends with enrollment in view, always.

## 4. The ESSER cliff: did spending ride one-time federal money?

Federal revenue share spiked nationwide in 2021-2024 as ESSER relief flowed in.
`tges_federal_exposure()` compares each district's pre-pandemic baseline federal
share to its ESSER-window peak, and flags districts whose per-pupil spending grew
during the surge, the structural set-up for a funding cliff.

```{r esser-compute}
fe <- tges_federal_exposure(tgm) %>% with_dfg()

fe %>%
  filter(district_code == NEWARK) %>%
  select(baseline_federal_share, peak_federal_share, federal_bump,
         pp_growth, cliff_exposure)

# How many DFG A peers carry the same exposure?
fe %>%
  filter(dfg == "A") %>%
  summarise(n = n(), flagged = sum(cliff_exposure, na.rm = TRUE))
```

```{r esser-chart}
fed_series <- rev %>%
  mutate(grp = case_when(
    district_code == NEWARK ~ "Newark",
    dfg == "A" ~ "DFG A peers",
    TRUE ~ NA_character_
  )) %>%
  filter(!is.na(grp)) %>%
  group_by(end_year, grp) %>%
  summarise(federal_share = median(federal_share, na.rm = TRUE), .groups = "drop")
stopifnot(nrow(fed_series) > 0)

ggplot(fed_series, aes(end_year, federal_share, color = grp)) +
  annotate("rect", xmin = 2020.5, xmax = 2024.5, ymin = -Inf, ymax = Inf,
           fill = "gray85", alpha = 0.5) +
  annotate("text", x = 2022.5, y = Inf, label = "ESSER window", vjust = 1.5,
           color = "gray45", size = 3.5) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_color_manual(values = c(Newark = nwk_red, `DFG A peers` = nwk_navy)) +
  labs(
    title = "Federal share of revenue: the ESSER bump",
    subtitle = "Newark vs the DFG A median. The shaded band is the relief window.",
    x = NULL, y = "Federal share of revenue", color = NULL,
    caption = "Source: NJ DOE TGES (VITSTAT federal revenue %)"
  ) +
  theme_nj()
```

## 5. The negotiation dashboard: ratios, salaries, and the benefits squeeze

Boards approve contracts against two numbers: are we competitive enough to retain
talent, and how much of every payroll dollar buys benefits instead of
instruction? `tges_staffing()` puts both in one frame.

```{r staffing-compute}
staff <- tges_staffing(tgm) %>% with_dfg()

# Newark's latest staffing line, ranked within DFG A on the bloat axis
staff_latest_yr <- max(staff$end_year[is.finite(staff$student_admin_ratio)])
newark_staff <- staff %>%
  filter(district_code == NEWARK, end_year == staff_latest_yr) %>%
  select(end_year, student_teacher_ratio, student_admin_ratio,
         faculty_admin_ratio, teacher_salary, admin_salary, benefits_pct_salary)
newark_staff

# Higher students-per-administrator = leaner administration. Rank within DFG A.
admin_rank <- staff %>%
  filter(end_year == staff_latest_yr) %>%
  tges_percentile_rank("student_admin_ratio", peer = "dfg") %>%
  filter(dfg == "A", district_code == NEWARK) %>%
  select(student_admin_ratio, peer_percentile, peer_n)
admin_rank
```

```{r staffing-chart}
benefits_series <- staff %>%
  filter(is.finite(benefits_pct_salary)) %>%
  mutate(grp = case_when(
    district_code == NEWARK ~ "Newark",
    dfg == "A" ~ "DFG A peers",
    TRUE ~ NA_character_
  )) %>%
  filter(!is.na(grp)) %>%
  group_by(end_year, grp) %>%
  summarise(benefits = median(benefits_pct_salary, na.rm = TRUE), .groups = "drop")
stopifnot(nrow(benefits_series) > 0)

ggplot(benefits_series, aes(end_year, benefits, color = grp)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_color_manual(values = c(Newark = nwk_red, `DFG A peers` = nwk_navy)) +
  labs(
    title = "The benefits squeeze: employee benefits as a share of salaries",
    subtitle = "Every point here is a payroll dollar buying pension and health, not a raise.",
    x = NULL, y = "Benefits as % of total salaries", color = NULL,
    caption = "Source: NJ DOE TGES (CSG14)"
  ) +
  theme_nj()
```

Newark holds roughly `r round(admin_rank$student_admin_ratio)` students per
administrator, ranking at the
`r round(admin_rank$peer_percentile)`th percentile of DFG A on that lean-vs-heavy
axis, evidence a board can use to rebut or confirm the "too many administrators"
charge.

## 6. The pre-meeting red-flag scan

Finally, the product a board member actually wants before a budget vote.
`tges_red_flags()` runs the peer rank across every major spending indicator and
surfaces where Newark lands in the top or bottom decile of DFG A.

```{r redflags-compute}
flags <- tges_red_flags(tgm, district_code = NEWARK, peer = "dfg",
                        only_flagged = FALSE)
flags
```

```{r redflags-chart}
flag_plot <- flags %>%
  mutate(
    band = case_when(
      flag == "top decile" ~ "Top decile",
      flag == "bottom decile" ~ "Bottom decile",
      TRUE ~ "Mid-pack"
    )
  )
stopifnot(nrow(flag_plot) > 0)

ggplot(flag_plot, aes(reorder(indicator, peer_percentile), peer_percentile,
                      fill = band)) +
  geom_col() +
  geom_hline(yintercept = c(10, 90), linetype = "dashed", color = "gray50") +
  coord_flip() +
  scale_fill_manual(values = c(`Top decile` = nwk_red, `Bottom decile` = nwk_teal,
                               `Mid-pack` = peer_gray)) +
  labs(
    title = "Newark's fiscal fingerprint within DFG A",
    subtitle = "Percentile rank by indicator. Dashed lines mark the top and bottom deciles.",
    x = NULL, y = "Percentile within DFG A", fill = NULL,
    caption = "Source: NJ DOE TGES. Check higher_means for each indicator's direction."
  ) +
  theme_nj()
```

The fingerprint reads in one glance: where Newark is top-decile on cost (and
whether that cost reaches a classroom or an administrative office), and where it
is bottom-decile. Pair this brief with the outcome percentiles from the MarGrady
study via `tges_efficiency()` to close the loop the original study left open:
the gains were real, and now the cost structure behind them is on the table too.

```{r session-info}
sessionInfo()
```
