---
title: "Following the Money: 25 Years of Spending in Newark"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Following the Money: 25 Years of Spending in Newark}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 6,
  out.width = "100%",
  message = FALSE,
  warning = FALSE,
  # Default OFF: only the expensive fetch chunks cache. Plot chunks must
  # re-evaluate every build, or knitr replays stale figures (caschooldata, 2026-03).
  cache = FALSE
)
```

```{r load-packages}
library(njschooldata)
library(ggplot2)
library(dplyr)
library(scales)
library(purrr)
library(tibble)

# The arts section reads a School Performance Report workbook (~100 MB), so give
# the download more than R's stingy 60-second default.
options(timeout = max(600, getOption("timeout")))
```

```{r theme}
theme_nj <- function() {
  theme_minimal(base_size = 14) +
    theme(
      plot.title = element_text(face = "bold", size = 16),
      plot.subtitle = element_text(color = "gray40"),
      plot.caption = element_text(color = "gray55", size = 9),
      panel.grid.minor = element_blank(),
      legend.position = "bottom"
    )
}

nwk_navy   <- "#16356B"
nwk_gold   <- "#C8A14B"
nwk_teal   <- "#16A085"
nwk_red    <- "#B83227"
nwk_purple <- "#6C3483"
```

The [Taxpayers' Guide to Educational Spending](https://www.nj.gov/education/guide/)
(TGES) is the New Jersey Department of Education's annual report on what every
district spends, broken down per pupil into classroom instruction, support
services, administration, plant operations, and more. `njschooldata` pulls it
straight from the state for every year back to 2001 (the pre-2011 editions were
branded the "Comparative Spending Guide").

This article follows one district, **Newark**, over a quarter century. Newark is
the state's largest school district, roughly 43,000 students and a budget near
\$1.5 billion. It is also one of the original *Abbott v. Burke* districts, where
a decades-long school-funding lawsuit forced the state to pour money into New
Jersey's poorest cities, and it spent more than 20 years under state control
before local control returned in 2018. The result is a spending story almost the
mirror image of a wealthy suburb like
[South Orange-Maplewood](somsd-school-spending.html): in Newark the *state*, not
the local taxpayer, writes most of the checks.

`fetch_tges(year)` returns one tidy data frame per spending indicator:

```{r intro-example, cache = TRUE}
spending_2025 <- fetch_tges(2025)

# one tidy table per TGES indicator (CSG1 = total budgetary cost per pupil)
names(spending_2025)
```

Each table is district-level and long, with a three-year window per report:
two finalized `Actuals` years plus the current `Budgeted` year. Newark is
district code `3570`.

```{r intro-glimpse}
spending_2025[["CSG1"]] %>%
  filter(district_code == "3570") %>%
  select(district_name, end_year, calc_type, `Per Pupil costs`, `District rank`)
```

```{r fetch-data, cache = TRUE}
# Pull every published guide, 2001-2025. fetch_tges(2025) was already fetched
# above, so reuse it and download the rest. Drop (with a visible warning) any
# year that fails rather than letting one network hiccup kill the whole build.
reports <- list("2025" = spending_2025)
for (y in 2001:2024) {
  reports[[as.character(y)]] <- tryCatch(
    fetch_tges(y),
    error = function(e) {
      warning(paste("TGES", y, "failed:", conditionMessage(e)))
      NULL
    }
  )
}
reports <- reports[!vapply(reports, is.null, logical(1))]
reports <- reports[order(as.integer(names(reports)))]
```

```{r helpers}
NEWARK <- "3570"

# Pull one Newark value column across every report and keep one row per school
# year. Each actual recurs in consecutive guides; we keep the most recent
# (most finalized) report's figure. VITSTAT/CSG16 carry no calc_type, so the
# calc filter is skipped for them.
nwk_series <- function(table, value_col, calc = "Actuals") {
  map_dfr(names(reports), function(y) {
    d <- reports[[y]][[table]]
    if (is.null(d) || !value_col %in% names(d)) return(NULL)
    r <- filter(d, !is.na(district_code), district_code == NEWARK)
    if ("calc_type" %in% names(r)) r <- filter(r, calc_type == calc)
    if (nrow(r) == 0) return(NULL)
    tibble(report_year = as.integer(y), end_year = r$end_year, value = r[[value_col]])
  }) %>%
    group_by(end_year) %>%
    slice_max(report_year, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    arrange(end_year)
}
```

## 1. The Abbott surge, then a post-2020 climb

Newark's actual budgetary cost per pupil sat near \$10,000 at the turn of the
century. Then it jumped: from \$13,687 (2004) to \$16,758 (2005), a 22% rise in
a single year. That surge lines up with the *Abbott v. Burke* parity remedy, in
which the state Supreme Court ordered New Jersey to fund its poorest districts
at the level of its wealthiest. Spending then plateaued in the \$16,000-\$18,000
range for the entire 2010s before climbing again after 2020, to \$21,202 in 2024.
The 2025 budget pushes it to **\$27,050**.

```{r pp-data}
pp <- nwk_series("CSG1", "Per Pupil costs", "Actuals")
pp_budget_2025 <- nwk_series("CSG1", "Per Pupil costs", "Budgeted") %>%
  filter(end_year == 2025)

stopifnot(nrow(pp) > 0, nrow(pp_budget_2025) == 1)
pp
```

The comparative "budgetary cost per pupil" above is the figure TGES uses to line
districts up against each other; it excludes things like capital construction
and debt service. Newark's *total* expenditures run far higher, about \$1.49
billion in 2024, or roughly \$34,000 per pupil once every fund is counted.

```{r total-spend}
reports[["2025"]][["CSG1AA_AVGS"]] %>%
  filter(district_code == NEWARK, calc_type == "Actuals", end_year == 2024) %>%
  transmute(end_year,
            total_expenditures = `Total Expenditures, actual costs`,
            enrollment = `Average Daily Enrollment plus Sent Pupils`,
            per_pupil_all_funds = `Per Pupil Total Expenditures`)
```

```{r pp-chart}
ggplot(pp, aes(end_year, value)) +
  annotate("rect", xmin = 2004, xmax = 2005, ymin = -Inf, ymax = Inf,
           fill = nwk_gold, alpha = 0.30) +
  annotate("text", x = 2004.5, y = 25500, label = "Abbott\nparity",
           color = "gray35", fontface = "italic", size = 3.3) +
  geom_line(color = nwk_navy, linewidth = 1.2) +
  geom_point(color = nwk_navy, size = 2) +
  geom_point(data = pp_budget_2025, aes(end_year, value),
             color = nwk_red, size = 3, shape = 17) +
  annotate("text", x = 2023.2, y = pp_budget_2025$value + 1100,
           label = "2025\nbudgeted", color = nwk_red, size = 3.3) +
  scale_y_continuous(labels = dollar) +
  scale_x_continuous(breaks = seq(2000, 2025, 5)) +
  labs(title = "Newark per-pupil spending: the Abbott jump, then a post-2020 climb",
       subtitle = "Budgetary cost per pupil, actuals (2025 is budgeted)",
       x = NULL, y = "Cost per pupil",
       caption = "Source: NJ DOE Taxpayers' Guide to Educational Spending") +
  theme_nj()
```

## 2. The state pays the bill

This is the heart of Newark's finances and the cleanest contrast with a
property-tax suburb. In South Orange-Maplewood, local taxpayers fund about 70-85%
of the budget. In Newark it is the reverse: **state aid covers roughly 80%** and
the local property-tax share has never topped 16%, drifting down to under 10% by
2024. The federal line spikes in a crisis (the 2010 stimulus, the 2023 pandemic
ESSER money) and recedes.

```{r revenue-data}
revenue <- bind_rows(
  nwk_series("VITSTAT_TOTAL", "Revenue: Local %")   %>% mutate(source = "Local taxes"),
  nwk_series("VITSTAT_TOTAL", "Revenue: State %")   %>% mutate(source = "State aid"),
  nwk_series("VITSTAT_TOTAL", "Revenue: Federal %") %>% mutate(source = "Federal")
)
stopifnot(nrow(revenue) > 0)

revenue %>%
  tidyr::pivot_wider(id_cols = end_year, names_from = source, values_from = value)
```

```{r revenue-chart}
ggplot(revenue, aes(end_year, value, color = source)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_color_manual(values = c("Local taxes" = nwk_navy,
                                "State aid" = nwk_teal,
                                "Federal" = nwk_gold)) +
  scale_y_continuous(labels = percent_format(accuracy = 1), limits = c(0, NA)) +
  scale_x_continuous(breaks = seq(2010, 2024, 4)) +
  labs(title = "In Newark, the state writes most of the checks",
       subtitle = "Share of revenue by source",
       x = NULL, y = "Share of revenue", color = NULL,
       caption = "Source: NJ DOE TGES Vital Statistics") +
  theme_nj()
```

## 3. Newark Public Schools outspends its charter rivals

Newark is one of the country's most-watched charter battlegrounds, the city of
Mark Zuckerberg's \$100 million gift and Dale Russakoff's book *The Prize*. About
a third of Newark's public-school students now attend charters. TGES reports
those charters as their own districts, so we can line them up against the
district. The two largest networks, **North Star Academy** (Uncommon Schools)
and **TEAM Academy** (KIPP), each spend close to \$18,000 per pupil, below the
district's \$21,202.

A caveat: the district and the charters do not carry identical cost burdens.
Newark Public Schools provides district-wide transportation and absorbs many of
the highest-cost special-education placements, expenses that never appear on a
charter's per-pupil line. The comparison is real, but it is not apples to apples.

```{r charter-data}
# major Newark charter networks, selected by district code so the comparison is
# explicit and reproducible
nwk_charters <- tribble(
  ~code,   ~name,
  "7320",  "North Star (Uncommon)",
  "7325",  "TEAM (KIPP)",
  "6053",  "Great Oaks Legacy",
  "7210",  "Marion P. Thomas",
  "6099",  "Link Community",
  "7730",  "Robert Treat Academy"
)

csg1_2024 <- reports[["2025"]][["CSG1"]] %>%
  filter(calc_type == "Actuals", end_year == 2024)

charter_cmp <- bind_rows(
  csg1_2024 %>% filter(district_code == NEWARK) %>%
    transmute(name = "Newark Public Schools", per_pupil = `Per Pupil costs`,
              kind = "District"),
  csg1_2024 %>% filter(district_code %in% nwk_charters$code) %>%
    left_join(nwk_charters, by = c("district_code" = "code")) %>%
    transmute(name, per_pupil = `Per Pupil costs`, kind = "Charter")
)
stopifnot(nrow(charter_cmp) > 0, all(!is.na(charter_cmp$per_pupil)))
charter_cmp %>% arrange(desc(per_pupil))
```

```{r charter-chart}
ggplot(charter_cmp, aes(reorder(name, per_pupil), per_pupil, fill = kind)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = dollar(per_pupil)), hjust = -0.1, size = 3.5) +
  coord_flip() +
  scale_fill_manual(values = c("District" = nwk_navy, "Charter" = nwk_gold)) +
  scale_y_continuous(labels = dollar, expand = expansion(mult = c(0, 0.18))) +
  labs(title = "The district spends more per pupil than Newark's big charters",
       subtitle = "Budgetary cost per pupil, 2024 actuals",
       x = NULL, y = "Cost per pupil", fill = NULL,
       caption = "Source: NJ DOE TGES (CSG1). Charters and the district do not carry identical cost burdens.") +
  theme_nj()
```

## 4. Benefits ate the salary line

Here is the trend that drives every contentious school budget meeting: employee
benefits as a share of salaries. In Newark it **roughly tripled**, from 10.9% in
2000 to 30.6% in 2024. Health-plan premiums are the main driver, and they keep
climbing faster than pay, the same pressure squeezing districts statewide.

```{r benefits-data}
# the 1999 figure is a known decimal artifact in the early files; start at 2000
benefits <- nwk_series("CSG14", "% of Total Salaries", "Actuals") %>%
  filter(end_year >= 2000)
stopifnot(nrow(benefits) > 0)
benefits
```

```{r benefits-chart}
ggplot(benefits, aes(end_year, value)) +
  geom_area(fill = nwk_gold, alpha = 0.25) +
  geom_line(color = nwk_gold, linewidth = 1.2) +
  geom_point(color = nwk_gold, size = 2) +
  scale_y_continuous(labels = percent_format(accuracy = 1),
                     limits = c(0, NA)) +
  scale_x_continuous(breaks = seq(2000, 2025, 5)) +
  labs(title = "Employee benefits as a share of salaries tripled",
       subtitle = "Newark benefit costs divided by total salaries, 2000-2024",
       x = NULL, y = "Benefits / salaries",
       caption = "Source: NJ DOE TGES (CSG14)") +
  theme_nj()
```

## 5. Where each dollar goes

TGES breaks the budgetary cost per pupil into its components. In the 2025 budget,
classroom instruction is the biggest slice, but the striking number is plant
operations: Newark budgets nearly as much to run and maintain its buildings
(\$4,818 per pupil) as it spends on all support services combined (\$4,999),
a footprint you rarely see in a younger suburban district.

```{r category-data}
category_tables <- tribble(
  ~category,                  ~table,
  "Classroom instruction",    "CSG2",
  "Support services",         "CSG6",
  "Plant ops & maintenance",  "CSG10",
  "Administration",           "CSG8",
  "Extracurricular",          "CSG13"
)

categories <- map_dfr(seq_len(nrow(category_tables)), function(i) {
  d <- reports[["2025"]][[category_tables$table[i]]]
  r <- filter(d, district_code == NEWARK, end_year == 2025)
  tibble(category = category_tables$category[i],
         per_pupil = r[["Per Pupil costs"]][1])
})

stopifnot(nrow(categories) > 0, all(!is.na(categories$per_pupil)))
categories %>% arrange(desc(per_pupil))
```

```{r category-chart}
ggplot(categories, aes(reorder(category, per_pupil), per_pupil)) +
  geom_col(fill = nwk_navy) +
  geom_text(aes(label = dollar(per_pupil)), hjust = -0.1, size = 3.5) +
  coord_flip() +
  scale_y_continuous(labels = dollar, expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Where Newark spends, per pupil (2025 budget)",
       subtitle = "Plant operations nearly matches all support services",
       x = NULL, y = "Cost per pupil",
       caption = "Source: NJ DOE TGES (CSG2, CSG6, CSG8, CSG10, CSG13)") +
  theme_nj()
```

## 6. Facilities: a heavy, growing line on aging buildings

In the wealthy suburbs, plant operations are a frozen line whose share of the
budget *shrinks* over time. Newark is the opposite. Heating, cleaning, and
repairing some of the oldest school buildings in the state cost \$1,528 per pupil
in 1999 and **\$3,899 by 2024**, and the facilities share of the budget held
between 15% and 19% the whole time, drifting *up* after 2020. (TGES tracks
operating cost only; the separate state-funded SDA capital program that rebuilds
Abbott-district schools is not in these numbers.)

```{r facilities-data}
facilities <- nwk_series("CSG10", "Per Pupil costs") %>%
  select(end_year, facilities = value) %>%
  inner_join(nwk_series("CSG1", "Per Pupil costs") %>% select(end_year, total = value),
             by = "end_year") %>%
  mutate(share = facilities / total)
stopifnot(nrow(facilities) > 0)
facilities
```

```{r facilities-share-chart}
ggplot(facilities, aes(end_year, share)) +
  geom_line(color = nwk_red, linewidth = 1.2) +
  geom_point(color = nwk_red, size = 2) +
  scale_y_continuous(labels = percent_format(accuracy = 1), limits = c(0, NA)) +
  scale_x_continuous(breaks = seq(2000, 2024, 6)) +
  labs(title = "Facilities keep a heavy, steady claim on Newark's budget",
       subtitle = "Plant operations & maintenance as a share of budgetary cost per pupil",
       x = NULL, y = "Facilities share of budget",
       caption = "Source: NJ DOE TGES (CSG10 / CSG1)") +
  theme_nj()
```

## 7. A closer look at the classroom dollar

TGES splits classroom instruction into three pieces: salaries and benefits,
supplies and textbooks, and "purchased services and other." Salaries and benefits
dominate and nearly doubled, from \$5,095 per pupil in 2000 to \$10,038 in 2024.
Supplies and textbooks stayed frozen near \$200-\$400 the entire period, a deep
real-terms cut. And purchased services, contracted and out-of-district
instruction, tell the reverse of the suburban story: they ballooned to \$1,589
per pupil in 2008, then collapsed back to \$154 by 2024 as Newark brought
services back in-house.

```{r classroom-data}
classroom_parts <- bind_rows(
  nwk_series("CSG3", "Per Pupil costs") %>% mutate(part = "Salaries & benefits"),
  nwk_series("CSG4", "Per Pupil costs") %>% mutate(part = "Supplies & textbooks"),
  nwk_series("CSG5", "Per Pupil costs") %>% mutate(part = "Purchased services & other")
) %>%
  mutate(part = factor(part, levels = c("Salaries & benefits",
                                        "Supplies & textbooks",
                                        "Purchased services & other")))
stopifnot(nrow(classroom_parts) > 0)

classroom_parts %>%
  filter(end_year %in% c(2000, 2008, 2016, 2024)) %>%
  select(end_year, part, value) %>%
  arrange(end_year, part)
```

```{r classroom-dollar-chart}
ggplot(classroom_parts, aes(end_year, value, fill = part)) +
  geom_area(alpha = 0.9) +
  scale_fill_manual(values = c("Salaries & benefits" = nwk_navy,
                               "Supplies & textbooks" = nwk_teal,
                               "Purchased services & other" = nwk_gold)) +
  scale_y_continuous(labels = dollar) +
  scale_x_continuous(breaks = seq(2000, 2024, 6)) +
  labs(title = "The classroom dollar: salaries up, contracted services in and out",
       subtitle = "Newark classroom cost per pupil by component",
       x = NULL, y = "Cost per pupil", fill = NULL,
       caption = "Source: NJ DOE TGES (CSG3, CSG4, CSG5)") +
  theme_nj()
```

## 8. From the top of the pack toward the middle

TGES ranks each district against its peer group. Newark sits with the large K-12
districts (3,500+ students). In this ranking, **1 = the lowest spender**, so a
higher number means spending more than peers. Through the Abbott years Newark was
ranked 104th or 105th out of about 105 districts, essentially the highest-spending
big district in the state. It has since slid toward the top quartile (rank in the
70s and 80s out of about 95) as its budget flattened in the recession years while
other districts caught up and the 2018 funding formula redistributed aid.

```{r rank-data}
rank <- nwk_series("CSG1", "District rank", "Actuals")
stopifnot(nrow(rank) > 0)

# size of the peer group in the most recent guide, for context
n_peers <- reports[["2025"]][["CSG1"]] %>%
  filter(group == "G. K-12 / 3501 +", end_year == 2024,
         !is.na(district_code), district_code != "00NA") %>%
  distinct(district_code) %>%
  nrow()
n_peers

rank
```

```{r rank-chart}
ggplot(rank, aes(end_year, value)) +
  geom_line(color = nwk_purple, linewidth = 1.2) +
  geom_point(color = nwk_purple, size = 2) +
  scale_x_continuous(breaks = seq(2000, 2024, 4)) +
  labs(title = "Newark's spending rank among large K-12 districts",
       subtitle = paste0("Rank within the roughly ", n_peers,
                          "-district peer group (1 = lowest spender)"),
       x = NULL, y = "Spending rank",
       caption = "Source: NJ DOE TGES (CSG1). Peer-group composition shifts year to year.") +
  theme_nj()
```

## 9. What the spending buys: the arts

TGES tells you what a district spends; the School Performance Report tells you
what students actually take. Newark has a deep arts-education tradition, home to
the nation's first public arts high school (Arts High School, founded 1931) and
the New Jersey Performing Arts Center. `fetch_arts_enrollment()` pulls visual and
performing arts participation, and Newark runs ahead of the state at every grade
band: arts enrollment is effectively universal in its elementary schools.

```{r fetch-arts, cache = TRUE}
arts_raw <- fetch_arts_enrollment(2024, level = "district") %>%
  filter(district_id == "3570")
stopifnot(nrow(arts_raw) > 0)
```

```{r arts-any-data}
arts_any <- arts_raw %>%
  transmute(grades,
            Newark = as.numeric(any_visual_perf_art_district),
            `New Jersey` = as.numeric(any_visual_perf_art_state)) %>%
  tidyr::pivot_longer(c(Newark, `New Jersey`), names_to = "geo", values_to = "pct")
stopifnot(nrow(arts_any) > 0)
arts_any
```

```{r arts-grade-chart}
ggplot(arts_any, aes(grades, pct, fill = geo)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.7) +
  geom_text(aes(label = paste0(round(pct), "%")),
            position = position_dodge(width = 0.75), vjust = -0.4, size = 3.3) +
  scale_fill_manual(values = c("Newark" = nwk_navy, "New Jersey" = "gray65")) +
  scale_y_continuous(labels = percent_format(scale = 1), limits = c(0, 112)) +
  labs(title = "Newark students take the arts at above-state rates",
       subtitle = "Share taking any visual or performing arts course, 2023-24",
       x = NULL, y = "Students taking an arts course", fill = NULL,
       caption = "Source: NJ DOE School Performance Report (Visual & Performing Arts)") +
  theme_nj()
```

The standout is dance. Across every grade band Newark enrolls students in dance
at many times the state rate, peaking at **38% of elementary students against a
statewide 4%**, a legacy of the city's performing-arts programs.

```{r arts-dance-data}
arts_dance <- arts_raw %>%
  transmute(grades,
            Newark = as.numeric(dance_district),
            `New Jersey` = as.numeric(dance_state)) %>%
  tidyr::pivot_longer(c(Newark, `New Jersey`), names_to = "geo", values_to = "pct")
stopifnot(nrow(arts_dance) > 0)
arts_dance
```

```{r arts-dance-chart}
ggplot(arts_dance, aes(grades, pct, fill = geo)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.7) +
  geom_text(aes(label = paste0(round(pct), "%")),
            position = position_dodge(width = 0.75), vjust = -0.4, size = 3.3) +
  scale_fill_manual(values = c("Newark" = nwk_red, "New Jersey" = "gray65")) +
  scale_y_continuous(labels = percent_format(scale = 1), limits = c(0, 45)) +
  labs(title = "Newark dance enrollment dwarfs the state rate",
       subtitle = "Share of students taking a dance course, 2023-24",
       x = NULL, y = "Students taking dance", fill = NULL,
       caption = "Source: NJ DOE School Performance Report (Visual & Performing Arts)") +
  theme_nj()
```

## Reproduce this yourself

Every chart above comes from `fetch_tges()`. Once you have a year's tables
(`spending_2025 <- fetch_tges(2025)`), pulling Newark's revenue mix is two lines:

```{r reproduce}
spending_2025[["VITSTAT_TOTAL"]] %>%
  filter(district_code == "3570") %>%
  select(district_name, end_year, `Revenue: State %`, `Revenue: Local %`, `Revenue: Federal %`)
```

Swap the district code for any NJ district, or loop `fetch_many_tges(2001:2025)`
for the full history.

```{r session-info}
sessionInfo()
```
</content>
</invoke>
