---
title: "Assessment & Graduation Detail: Behind the Headline Rates"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Assessment & Graduation Detail: Behind the Headline Rates}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 6,
  out.width = "100%",
  message = FALSE,
  warning = FALSE,
  cache = FALSE
)
```

```{r load-packages}
library(njschooldata)
library(dplyr)
library(ggplot2)
```

The 2024-25 School Performance Reports added several sheets that slice
assessment and graduation results more finely than the headline proficiency and
graduation rates. This vignette uses three of the new fetchers to show how the
detail changes the story:

- `fetch_spr_proficiency_by_test()` splits NJSLA math by the *test* a student
  sat, separating the grade-level tests from the high-school end-of-course
  exams (Algebra I, Geometry, Algebra II) that `fetch_parcc()` folds together.
- `fetch_spr_grad_cohort()` returns the 4-, 5-, and 6-year graduation cohorts in
  one frame, so the recovery between them is visible at a glance.
- `fetch_spr_fed_grad()` reads the federally reported Adjusted Cohort Graduation
  Rate (ACGR) by student group, the rate used for ESSA accountability.

Two companion fetchers round out the set: `fetch_spr_science_grade()` (NJSLA
science by grade) and `fetch_spr_elp_progress()` (progress toward English
language proficiency). Every value comes straight from the published workbook;
suppressed cells stay `NA` rather than being guessed.

## Grade-8 math looks like a crisis until you see who skipped the test

The single lowest NJSLA math proficiency rate in New Jersey is **Grade 8, at
20.7%**. Taken alone that reads like a collapse. The by-test breakdown explains
it: the strongest eighth graders do not sit the Grade 8 NJSLA at all. They take
**Algebra I** (38.1% proficient) or **Geometry** (58.4%) as accelerated math.
What is left in the Grade 8 pool is the lower-performing remainder, so the
grade-level rate is depressed by who *left* it, not only by who is in it. The
end-of-course exams, which `fetch_parcc()` does not separate, are the missing
context.

```{r math-by-test-data}
math_state <- fetch_spr_proficiency_by_test(2025, subject = "math",
                                            level = "district") %>%
  filter(is_state, subgroup == "total population")

stopifnot(nrow(math_state) > 0)

math_state <- math_state %>%
  mutate(kind = if_else(grade_test %in% c("Algebra I", "Geometry", "Algebra II"),
                        "End-of-course", "Grade-level NJSLA")) %>%
  arrange(proficiency_rate)

# Print-before-plot: the data feeding the chart.
math_state %>% select(grade_test, kind, valid_scores, proficiency_rate)
```

```{r math-by-test-chart}
ggplot(math_state, aes(x = reorder(grade_test, proficiency_rate),
                       y = proficiency_rate, fill = kind)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = paste0(proficiency_rate, "%")), hjust = -0.15,
            size = 3.6) +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  scale_fill_manual(values = c("End-of-course" = "#1b7837",
                               "Grade-level NJSLA" = "#762a83")) +
  labs(
    title = "NJ math proficiency by test, 2024-25",
    subtitle = "Grade 8 is the lowest grade-level rate; strong students sit Algebra I / Geometry instead",
    x = NULL,
    y = "Percent proficient (NJSLA level 4 or 5)",
    fill = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top")
```

## The 6-year cohort quietly recovers students the 4-year rate misses

The headline graduation rate is the 4-year cohort: 91.8% statewide for the class
that should have finished in 2024-25. Following the same students another two
years lifts the rate to **92.9% at five years and 93.2% at six**, while the
share still enrolled ("continuing") falls from 3.4% to 1.0%. The 6-year cohort
also reports a high-school **persistence** rate of 94.2% (graduated plus still
enrolled). Combining all three cohort lengths in one frame, which
`fetch_spr_grad_cohort()` does, makes the slow recovery legible.

```{r grad-cohort-data}
cohort_state <- fetch_spr_grad_cohort(2025, level = "district") %>%
  filter(is_state, subgroup == "total population") %>%
  select(cohort_type, graduated, continuing, non_continuing, persisting)

stopifnot(nrow(cohort_state) == 3)

# Print-before-plot.
cohort_state
```

```{r grad-cohort-chart}
ggplot(cohort_state, aes(x = cohort_type, y = graduated, group = 1)) +
  geom_line(color = "#2166ac", linewidth = 1.2) +
  geom_point(color = "#2166ac", size = 3.5) +
  geom_text(aes(label = paste0(graduated, "%")), vjust = -1.1, size = 4) +
  scale_y_continuous(limits = c(90, 94)) +
  labs(
    title = "NJ statewide graduation rate by cohort length, class of 2024-25",
    subtitle = "Following the same students longer recovers about 1.4 points",
    x = NULL,
    y = "Percent graduated"
  ) +
  theme_minimal(base_size = 12)
```

## The federal graduation rate hides a 43-point equity gap

The federally reported ACGR is a single statewide 4-year number, 88.9%. By
student group it spans an enormous range. Students in foster care graduate at
**53.4%** and students with disabilities at **65.6%**, against **96.5%** for
Asian/Native Hawaiian/Pacific Islander students and **92.8%** for white students.
`fetch_spr_fed_grad()` returns the rate by subgroup and cohort length, so the gap
behind the headline is one filter away.

```{r fed-grad-data}
fed_state <- fetch_spr_fed_grad(2025, level = "district") %>%
  filter(is_state, cohort_years == 4,
         !is.na(graduation_rate_federal)) %>%
  select(subgroup, graduation_rate_federal) %>%
  arrange(graduation_rate_federal)

stopifnot(nrow(fed_state) > 0)

total_rate <- fed_state$graduation_rate_federal[fed_state$subgroup == "total population"]

# Print-before-plot.
fed_state
```

```{r fed-grad-chart}
fed_plot <- fed_state %>%
  mutate(highlight = subgroup == "total population")

ggplot(fed_plot, aes(x = reorder(subgroup, graduation_rate_federal),
                     y = graduation_rate_federal, fill = highlight)) +
  geom_col(width = 0.75) +
  geom_hline(yintercept = total_rate, linetype = "dashed", color = "grey30") +
  geom_text(aes(label = paste0(graduation_rate_federal, "%")), hjust = -0.15,
            size = 3.2) +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  scale_fill_manual(values = c("TRUE" = "#b2182b", "FALSE" = "#4393c3"),
                    guide = "none") +
  labs(
    title = "Federal 4-year ACGR by student group, NJ 2024-25",
    subtitle = "Dashed line = statewide 88.9%; foster-care and disability cohorts trail by 20-35 points",
    x = NULL,
    y = "Federal adjusted cohort graduation rate"
  ) +
  theme_minimal(base_size = 12)
```

## Notes

- **Coverage.** The proficiency-by-test, science-by-grade, ELP-progress, and
  graduation-cohort sheets are new in the 2024-25 redesign and are available for
  `end_year` 2025 only. `fetch_spr_fed_grad()` covers 2021-2025 (the 6-year
  cohort from 2024).
- **Suppression.** Cells for fewer than ten students are masked in the source and
  return `NA`; the 4- and 5-year cohorts report no persistence rate ("n/a"), so
  `persisting` is `NA` there.
- **Entity values.** Each sheet repeats its values as a school/district/state
  triple; these fetchers return the entity-appropriate value, and the statewide
  figure is carried on the `is_state` row of district-level output.

```{r session-info}
sessionInfo()
```
