---
title: "Advanced-Coursework Access & Equity"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced-Coursework Access & Equity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 6,
  out.width = "100%",
  message = FALSE,
  warning = FALSE,
  cache = FALSE
)
```

```{r load-packages}
library(njschooldata)
library(dplyr)
library(ggplot2)
```

The usual AP/IB question is "how many students took an exam." The harder,
more newsworthy question is about **access**: does a school even offer advanced
courses, and once offered, who actually gets in. `fetch_advanced_course_access()`
answers both. A single front door over three NJ School Performance Report sheet
families:

- `type = "courses_offered"` -- one row per school per advanced course
  (enrolled / tested), 2017-2025.
- `type = "participation_by_group"` -- AP/IB and dual-enrollment participation
  rates broken out by student group, 2021-2025.
- `type = "sle"` -- Structured Learning Experience participation, 2017-2025.

Every rate and count comes straight from the published SPR workbooks; suppression
and no-data text is kept as `NA`, never a guessed number.

## The AP/IB access gap by student group

`type = "participation_by_group"` reports, for each student group, the percent of
students enrolled in one or more AP/IB courses. The statewide rate per group is
carried in the `apib_pct_state` column. Lining the groups up shows an access gap
that is anything but subtle: in 2023-24 Asian students participated at **70%**
while students with disabilities sat at **under 5%**, with English learners,
Black, Hispanic, and economically disadvantaged students all well below the
statewide average.

```{r access-gap-data}
groups <- c(
  "asian, native hawaiian, or pacific islander", "white", "multiracial",
  "total population", "hispanic", "black", "economically disadvantaged",
  "limited english proficiency", "students with disabilities"
)

apib_state <- fetch_advanced_course_access(
  2024, type = "participation_by_group", level = "district"
) %>%
  filter(is_state, subgroup %in% groups) %>%
  group_by(subgroup) %>%
  summarise(apib_pct = first(apib_pct_state), .groups = "drop") %>%
  filter(!is.na(apib_pct)) %>%
  arrange(apib_pct)

# Print-before-plot: confirm the data feeding the chart.
stopifnot(nrow(apib_state) > 0)
apib_state
```

```{r access-gap-chart}
ggplot(apib_state, aes(x = reorder(subgroup, apib_pct), y = apib_pct)) +
  geom_col(fill = "#2c7fb8") +
  geom_text(aes(label = paste0(apib_pct, "%")), hjust = -0.15, size = 3.6) +
  coord_flip() +
  scale_y_continuous(limits = c(0, 80), expand = expansion(mult = c(0, 0.08))) +
  labs(
    title = "Who is enrolled in AP/IB coursework in New Jersey, 2023-24",
    subtitle = "Statewide share of each student group enrolled in one or more AP/IB courses",
    x = NULL,
    y = "Percent enrolled in one or more AP/IB courses"
  ) +
  theme_minimal(base_size = 13)
```

## How many advanced courses does a school offer at all?

Access starts with supply. `type = "courses_offered"` lists every advanced course
a school runs, so counting distinct courses per school measures the **breadth** of
the advanced-course menu. In 2023-24 the typical NJ high school offering any
advanced course listed about 19 of them, but the range is enormous - from a single
course to nearly 50.

```{r breadth-data}
breadth <- fetch_advanced_course_access(2024, type = "courses_offered") %>%
  filter(is_school) %>%
  group_by(district_id, school_id, school_name) %>%
  summarise(n_courses = n_distinct(course_name), .groups = "drop")

# Print-before-plot.
stopifnot(nrow(breadth) > 0)
summary(breadth$n_courses)
```

```{r breadth-chart}
ggplot(breadth, aes(x = n_courses)) +
  geom_histogram(binwidth = 2, fill = "#2c7fb8", colour = "white") +
  geom_vline(xintercept = median(breadth$n_courses),
             linetype = "dashed", colour = "#d95f0e", linewidth = 0.9) +
  labs(
    title = "Breadth of the advanced-course menu, 2023-24",
    subtitle = paste0("Distinct AP/IB courses offered per school; dashed line = median (",
                      median(breadth$n_courses), " courses)"),
    x = "Number of distinct advanced courses offered",
    y = "Number of schools"
  ) +
  theme_minimal(base_size = 13)
```

The schools with the widest menus:

```{r breadth-top}
breadth %>%
  arrange(desc(n_courses)) %>%
  slice_head(n = 10) %>%
  select(school_name, n_courses)
```

## Notes

- `fetch_advanced_course_access()` accepts `level = "school"` (default) or
  `level = "district"`; the statewide `is_state` row lives in the District
  workbook.
- Coverage: `courses_offered` and `sle` span 2017-2025; `participation_by_group`
  spans 2021-2025 (the sheet is absent 2017-2020 and those years error). The 2025
  participation sheet is a multi-year trend table filtered to the requested year.
- Suppression / no-data strings (e.g. "There is no data available for this school
  year.") are coerced to `NA`, never a fabricated number.

```{r session-info}
sessionInfo()
```
