---
title: "Seal of Biliteracy: Statewide Trend & the Seal-Earning Equity Gap"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Seal of Biliteracy: Statewide Trend & the Seal-Earning Equity Gap}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 6,
  out.width = "100%",
  message = FALSE,
  warning = FALSE,
  cache = FALSE
)
```

```{r load-packages}
library(njschooldata)
library(dplyr)
library(ggplot2)
```

The New Jersey State Seal of Biliteracy recognizes graduating students who reach
proficiency in English and at least one world language. The 2024-25 School
Performance Reports split this into a richer set of views: a school/district/state
**summary** (`fetch_biliteracy_summary()`), a five-year **trend**
(`fetch_biliteracy_trends()`), and a seal-earning rate **by student group**
(`fetch_biliteracy_by_group()`). All three are new in the 2024-25 redesign and
cover end_year 2025 only; the per-language detail (2018-2025) stays in
`fetch_biliteracy_seal()`.

Throughout, suppressed cells (`"Fewer than 5 seals"`, `"Enrollment for the group
is <10 students."`) are returned as `NA`, never a guessed number. A real
published `0` stays `0`.

## Statewide seals more than doubled in four years

The trend sheet is unusual: it carries five school years inside the single 2025
workbook. The statewide row shows seals earned climbing from about 4,950 in
2020-21 to 12,644 in 2024-25.

```{r trend-data, cache = TRUE}
state_trend <- fetch_biliteracy_trends(2025, level = "district") %>%
  filter(is_state) %>%
  select(school_year, total_seals_earned) %>%
  arrange(school_year)
```

```{r trend-print}
# Print-before-plot: confirm the data feeding the chart.
state_trend
stopifnot(nrow(state_trend) == 5)
```

```{r trend-chart}
ggplot(state_trend, aes(x = school_year, y = total_seals_earned, group = 1)) +
  geom_line(colour = "#2c7fb8", linewidth = 1.1) +
  geom_point(size = 3, colour = "#2c7fb8") +
  geom_text(aes(label = format(total_seals_earned, big.mark = ",")),
            vjust = -1, size = 3.6) +
  scale_y_continuous(limits = c(0, 15000),
                     labels = function(x) format(x, big.mark = ",")) +
  labs(
    title = "New Jersey Seal of Biliteracy: statewide seals earned",
    subtitle = "Total seals earned per school year (SealofBiliteracy_Trends, 2024-25 SPR)",
    x = "School year",
    y = "Total seals earned"
  ) +
  theme_minimal(base_size = 13)
```

## English learners earn seals at the highest rate

The student-group view reframes the seal as an equity measure. Because the seal
rewards proficiency in a language other than English, the groups earning it at
the highest rates are exactly the students who arrive multilingual: current and
former English learners lead at about 25%, with Hispanic/Latino students next.
Students with disabilities and Black students earn seals at the lowest rates.
The statewide rate for each group is carried in the
`students_earning_seal_pct_state` column.

```{r group-data, cache = TRUE}
group_rates <- fetch_biliteracy_by_group(2025, level = "district") %>%
  filter(!is.na(students_earning_seal_pct_state)) %>%
  distinct(subgroup, students_earning_seal_pct_state)
```

```{r group-print}
# Print-before-plot: confirm the values exist and are sane.
group_rates <- group_rates %>%
  arrange(desc(students_earning_seal_pct_state))
group_rates
stopifnot(nrow(group_rates) > 0)
stopifnot("limited english proficiency" %in% group_rates$subgroup)
```

```{r group-chart}
ggplot(
  group_rates,
  aes(x = reorder(subgroup, students_earning_seal_pct_state),
      y = students_earning_seal_pct_state)
) +
  geom_col(fill = "#2c7fb8") +
  geom_text(aes(label = paste0(students_earning_seal_pct_state, "%")),
            hjust = -0.15, size = 3.4) +
  coord_flip() +
  scale_y_continuous(limits = c(0, 30)) +
  labs(
    title = "Seal-earning rate by student group, 2024-25 (statewide)",
    subtitle = "Percent of each group earning a Seal of Biliteracy (SealofBiliteracy_StudentGroup)",
    x = NULL,
    y = "Percent earning a seal"
  ) +
  theme_minimal(base_size = 13)
```

## Summary view: seals, languages, and unique earners

The summary sheet pairs each entity's seal count with the number of distinct
languages and the count and share of unique students earning a seal. Here are the
districts producing the most seals statewide.

```{r summary-data, cache = TRUE}
summary25 <- fetch_biliteracy_summary(2025, level = "district") %>%
  filter(is_district, !is.na(total_seals_earned))
```

```{r summary-print}
nrow(summary25)
stopifnot(nrow(summary25) > 0)

summary25 %>%
  arrange(desc(total_seals_earned)) %>%
  slice(1:10) %>%
  select(district_name, total_seals_earned, numberof_languages,
         unique_students_earning_seals, unique_students_earning_seals_pct)
```

## Notes

- All three fetchers cover **end_year 2025 only** (the 2024-25 SPR redesign sheets
  `SealofBiliteracy_Summary`, `_Trends`, `_StudentGroup`). For the per-language
  detail across 2018-2025 use `fetch_biliteracy_seal()`.
- Suppression and text-bleed strings become `NA`, never a fabricated number. A
  genuine published `0` is preserved.
- A handful of small high schools publish a seal-earning rate above 100% for a
  group (the rate is seals earned over a 12th-grade-style denominator, not over
  the group's own enrollment); those real cells pass through unclipped.

```{r session-info}
sessionInfo()
```
