---
title: "New Jersey Enrollment Insights"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{New Jersey Enrollment Insights}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 6,
  out.width = "100%",
  message = FALSE,
  warning = FALSE,
  cache = TRUE
)
```

```{r load-packages}
library(njschooldata)
library(ggplot2)
library(dplyr)
library(scales)
```

```{r theme}
theme_nj <- function() {
  theme_minimal(base_size = 14) +
    theme(
      plot.title = element_text(face = "bold", size = 16),
      plot.subtitle = element_text(color = "gray40"),
      panel.grid.minor = element_blank(),
      legend.position = "bottom"
    )
}

nj_colors <- c("total" = "#2C3E50", "white" = "#3498DB", "black" = "#E74C3C",
               "hispanic" = "#F39C12", "asian" = "#9B59B6", "charter" = "#1ABC9C",
               "multiracial" = "#27AE60", "prek" = "#E67E22")
```

```{r fetch-data, cache = TRUE}
# Fetch 2020-2025 enrollment data (post-format-change for consistent structure)
years <- 2020:2025
enr_all <- purrr::map_df(years, ~{
  tryCatch(
    fetch_enr(.x, tidy = TRUE),
    error = function(e) {
      warning(paste("Year", .x, "failed:", conditionMessage(e)))
      NULL
    }
  )
})

enr_current <- fetch_enr(2025, tidy = TRUE)

# State-level summary aggregated from district totals for time-series consistency
state_summary <- enr_all %>%
  filter(is_district) %>%
  group_by(end_year, subgroup, grade_level) %>%
  summarize(n_students = sum(n_students, na.rm = TRUE), .groups = "drop")

state_totals <- state_summary %>%
  filter(subgroup == "total_enrollment") %>%
  select(end_year, grade_level, total = n_students)

state_summary <- state_summary %>%
  left_join(state_totals, by = c("end_year", "grade_level")) %>%
  mutate(pct = n_students / total)
```

## 1. New Jersey educates 1.4 million students

New Jersey has one of the largest public school systems in the country, with enrollment holding steady through COVID and beyond.

```{r statewide-enrollment}
state_total <- state_summary %>%
  filter(subgroup == "total_enrollment", grade_level == "TOTAL")

stopifnot(nrow(state_total) > 0)
state_total

ggplot(state_total, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.5, color = nj_colors["total"]) +
  geom_point(size = 3, color = nj_colors["total"]) +
  scale_y_continuous(labels = comma, limits = c(0, NA)) +
  labs(title = "New Jersey Educates 1.4 Million Students",
       subtitle = "Statewide public school enrollment has held steady since 2020",
       x = "School Year", y = "Students") +
  theme_nj()
```

## 2. Charter enrollment grew 15% in five years

New Jersey's charter sector added 8,000+ students from 2020 to 2025, growing from 55,600 to over 63,800.

```{r charter-growth}
charter_trend <- enr_all %>%
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL") %>%
  mutate(sector = ifelse(is_charter, "Charter", "Traditional")) %>%
  group_by(end_year, sector) %>%
  summarize(n_students = sum(n_students, na.rm = TRUE), .groups = "drop")

stopifnot(nrow(charter_trend) > 0)
charter_trend

ggplot(charter_trend, aes(x = end_year, y = n_students, fill = sector)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = comma) +
  scale_fill_manual(values = c("Charter" = nj_colors["charter"],
                               "Traditional" = nj_colors["total"])) +
  labs(title = "Charter Enrollment Grew 15% in Five Years",
       subtitle = "From 55,600 to 63,800 students statewide",
       x = "School Year", y = "Students", fill = "") +
  theme_nj()
```

## 3. Hispanic students hit 35% and rising

Hispanic enrollment surged from 30% to 35% of all NJ students in just five years, making it the fastest demographic shift in state history.

```{r hispanic-growth}
hispanic <- state_summary %>%
  filter(subgroup == "hispanic", grade_level == "TOTAL")

stopifnot(nrow(hispanic) > 0)
hispanic %>% select(end_year, n_students, pct)

ggplot(hispanic, aes(x = end_year, y = pct * 100)) +
  geom_line(linewidth = 1.5, color = nj_colors["hispanic"]) +
  geom_point(size = 3, color = nj_colors["hispanic"]) +
  labs(title = "Hispanic Students Hit 35% and Rising",
       subtitle = "From 30% to 35% of all NJ students in five years",
       x = "School Year", y = "Percent Hispanic") +
  theme_nj()
```

## 4. The Big Three: Newark, Jersey City, and Paterson

New Jersey's three largest districts educate over 93,000 students combined - nearly 7% of the state.

```{r big-three}
big_three_names <- c("Newark Public School District",
                     "Jersey City Public Schools",
                     "Paterson Public School District")
big_three_trend <- enr_all %>%
  filter(is_district, !is_charter,
         district_name %in% big_three_names,
         subgroup == "total_enrollment", grade_level == "TOTAL")

stopifnot(nrow(big_three_trend) > 0)
big_three_trend %>% select(end_year, district_name, n_students)

ggplot(big_three_trend, aes(x = end_year, y = n_students, color = district_name)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_y_continuous(labels = comma) +
  labs(title = "The Big Three: Newark, Jersey City, and Paterson",
       subtitle = "Combined enrollment of over 93,000 students",
       x = "School Year", y = "Students", color = "") +
  theme_nj()
```

## 5. Kindergarten rebounded from COVID

New Jersey lost 5% of kindergartners during COVID - but by 2025, K enrollment nearly recovered while Pre-K surged past pre-pandemic levels.

```{r covid-kindergarten}
k_trend <- state_summary %>%
  filter(subgroup == "total_enrollment",
         grade_level %in% c("PK", "K", "01", "06", "12")) %>%
  mutate(grade_label = case_when(
    grade_level == "PK" ~ "Pre-K",
    grade_level == "K" ~ "Kindergarten",
    grade_level == "01" ~ "Grade 1",
    grade_level == "06" ~ "Grade 6",
    grade_level == "12" ~ "Grade 12",
    TRUE ~ grade_level
  ))

stopifnot(nrow(k_trend) > 0)
k_trend %>%
  filter(grade_level %in% c("K", "PK")) %>%
  select(end_year, grade_label, n_students)

ggplot(k_trend, aes(x = end_year, y = n_students, color = grade_label)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_y_continuous(labels = comma) +
  labs(title = "Kindergarten Rebounded from COVID",
       subtitle = "K enrollment dipped but Pre-K surged past pre-pandemic levels",
       x = "School Year", y = "Students", color = "") +
  theme_nj()
```

## 6. Free/reduced lunch ranges from 98% to under 5%

Passaic City has 98% of students on free/reduced lunch while affluent suburbs like Millburn have under 5% - a stark measure of NJ's wealth divide.

```{r frl-distribution}
frl <- enr_current %>%
  filter(is_district, !is_charter,
         subgroup == "free_reduced_lunch", grade_level == "TOTAL",
         !is.na(pct), n_students >= 100) %>%
  arrange(desc(pct)) %>%
  head(15) %>%
  mutate(district_label = reorder(district_name, pct))

stopifnot(nrow(frl) > 0)
frl %>% select(district_name, n_students, pct)

ggplot(frl, aes(x = district_label, y = pct * 100)) +
  geom_col(fill = nj_colors["total"]) +
  coord_flip() +
  labs(title = "Free/Reduced Lunch Varies Dramatically",
       subtitle = "Some districts approach 100% while affluent suburbs are under 5%",
       x = "", y = "Percent Free/Reduced Lunch") +
  theme_nj()
```

## 7. White students dropped below 37%

White students went from 42% to under 37% of NJ public school enrollment in just five years. NJ public schools are now decisively majority-minority.

```{r demographic-shift}
demo <- state_summary %>%
  filter(subgroup %in% c("white", "hispanic", "black", "asian"),
         grade_level == "TOTAL") %>%
  mutate(subgroup = factor(subgroup, levels = c("white", "hispanic", "black", "asian")))

stopifnot(nrow(demo) > 0)
demo %>% select(end_year, subgroup, pct) %>%
  mutate(pct = round(pct * 100, 1)) %>%
  tidyr::pivot_wider(names_from = subgroup, values_from = pct)

ggplot(demo, aes(x = end_year, y = pct * 100, color = subgroup)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_color_manual(values = nj_colors, labels = c("White", "Hispanic", "Black", "Asian")) +
  labs(title = "White Students Dropped Below 37%",
       subtitle = "NJ public schools are now decisively majority-minority",
       x = "School Year", y = "Percent of Students", color = "") +
  theme_nj()
```

## 8. English learners top 45% in some districts

ELL students make up over 45% in Plainfield and Lakewood but under 1% in most suburban districts - a concentration driven by immigration patterns.

```{r ell-concentration}
ell <- enr_current %>%
  filter(is_district, !is_charter,
         subgroup == "lep", grade_level == "TOTAL",
         !is.na(pct), n_students >= 50) %>%
  arrange(desc(pct)) %>%
  head(15) %>%
  mutate(district_label = reorder(district_name, pct))

stopifnot(nrow(ell) > 0)
ell %>% select(district_name, n_students, pct)

ggplot(ell, aes(x = district_label, y = pct * 100)) +
  geom_col(fill = nj_colors["hispanic"]) +
  coord_flip() +
  labs(title = "English Learners Top 45% in Some Districts",
       subtitle = "Concentrated in urban areas; under 1% in most suburbs",
       x = "", y = "Percent English Language Learners") +
  theme_nj()
```

## 9. Top 10 districts serve 15% of all students

Just 10 out of 580+ districts educate nearly 1 in 6 NJ students. Newark alone has 44,000.

```{r top-10-districts}
top_10 <- enr_current %>%
  filter(is_district, !is_charter,
         subgroup == "total_enrollment", grade_level == "TOTAL") %>%
  arrange(desc(n_students)) %>%
  head(10) %>%
  mutate(district_label = reorder(district_name, n_students))

stopifnot(nrow(top_10) > 0)
top_10 %>% select(district_name, n_students)

ggplot(top_10, aes(x = district_label, y = n_students)) +
  geom_col(fill = nj_colors["total"]) +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  labs(title = "Top 10 Districts Serve 15% of All Students",
       subtitle = "Just 10 out of 580+ districts educate nearly 1 in 6 NJ students",
       x = "", y = "Students") +
  theme_nj()
```

## 10. Multiracial students: fastest-growing category

Multiracial students grew 39% in five years - from 2.4% to 3.3% of enrollment - making it the fastest-growing racial category in NJ.

```{r multiracial-growth}
multi <- state_summary %>%
  filter(subgroup == "multiracial", grade_level == "TOTAL")

stopifnot(nrow(multi) > 0)
multi %>% select(end_year, n_students, pct)

ggplot(multi, aes(x = end_year, y = pct * 100)) +
  geom_line(linewidth = 1.5, color = nj_colors["multiracial"]) +
  geom_point(size = 3, color = nj_colors["multiracial"]) +
  labs(title = "Multiracial Students: Fastest-Growing Category",
       subtitle = "From 2.4% to 3.3% of enrollment in five years (39% growth)",
       x = "School Year", y = "Percent Multiracial") +
  theme_nj()
```

## 11. Pre-K nearly doubled since 2020

NJ's Pre-K enrollment surged from 45,000 to 87,000 in five years - fueled by the state's expanding universal pre-K program.

```{r prek-surge}
prek <- state_summary %>%
  filter(subgroup == "total_enrollment", grade_level == "PK")

stopifnot(nrow(prek) > 0)
prek %>% select(end_year, n_students)

ggplot(prek, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.5, color = nj_colors["prek"]) +
  geom_point(size = 3, color = nj_colors["prek"]) +
  scale_y_continuous(labels = comma, limits = c(0, NA)) +
  labs(title = "Pre-K Nearly Doubled Since 2020",
       subtitle = "From 45,000 to 87,000 students - NJ's universal pre-K expansion",
       x = "School Year", y = "Pre-K Students") +
  theme_nj()
```

## 12. Bergen County has more students than 12 US states

With 132,000+ students, Bergen County alone has a larger public school system than Wyoming, Vermont, North Dakota, and nine other states.

```{r county-enrollment}
county_enr <- enr_current %>%
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL") %>%
  group_by(county_name) %>%
  summarize(n_students = sum(n_students, na.rm = TRUE),
            n_districts = n(), .groups = "drop") %>%
  filter(county_name != "Charters") %>%
  arrange(desc(n_students)) %>%
  head(15) %>%
  mutate(county_label = reorder(county_name, n_students))

stopifnot(nrow(county_enr) > 0)
county_enr

ggplot(county_enr, aes(x = county_label, y = n_students)) +
  geom_col(fill = nj_colors["total"]) +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  labs(title = "Bergen County Has More Students Than 12 US States",
       subtitle = "Top 15 NJ counties by enrollment",
       x = "", y = "Students") +
  theme_nj()
```

## 13. Most NJ districts are tiny

Half of NJ's 580 districts have fewer than 1,200 students. The median district is smaller than a single large high school.

```{r district-size-distribution}
dist_sizes <- enr_current %>%
  filter(is_district, !is_charter,
         subgroup == "total_enrollment", grade_level == "TOTAL")

stopifnot(nrow(dist_sizes) > 0)
cat("Districts:", nrow(dist_sizes), "\n")
cat("Median:", median(dist_sizes$n_students, na.rm = TRUE), "\n")
cat("Under 1000:", sum(dist_sizes$n_students < 1000, na.rm = TRUE), "\n")
cat("Over 10000:", sum(dist_sizes$n_students > 10000, na.rm = TRUE), "\n")

ggplot(dist_sizes, aes(x = n_students)) +
  geom_histogram(binwidth = 500, fill = nj_colors["total"], color = "white") +
  geom_vline(xintercept = median(dist_sizes$n_students, na.rm = TRUE),
             linetype = "dashed", color = "red", linewidth = 1) +
  scale_x_continuous(labels = comma) +
  labs(title = "Most NJ Districts Are Tiny",
       subtitle = "Half have fewer than 1,200 students (red line = median)",
       x = "Students", y = "Number of Districts") +
  theme_nj()
```

## 14. NJ's enrollment pyramid shows the pre-K boom

The 2025 grade-level distribution reveals the pre-K surge: PK enrollment (87K) is approaching K (89K), reflecting NJ's universal pre-K push.

```{r grade-pyramid}
grade_enr <- state_summary %>%
  filter(end_year == 2025, subgroup == "total_enrollment",
         grade_level != "TOTAL", !is.na(grade_level)) %>%
  mutate(grade_label = case_when(
    grade_level == "PK" ~ "Pre-K",
    grade_level == "K" ~ "K",
    TRUE ~ paste("Grade", grade_level)
  ),
  grade_order = case_when(
    grade_level == "PK" ~ 0,
    grade_level == "K" ~ 1,
    TRUE ~ as.numeric(grade_level) + 1
  )) %>%
  arrange(grade_order) %>%
  mutate(grade_label = factor(grade_label, levels = grade_label))

stopifnot(nrow(grade_enr) > 0)
grade_enr %>% select(grade_label, n_students)

ggplot(grade_enr, aes(x = grade_label, y = n_students)) +
  geom_col(fill = nj_colors["total"]) +
  scale_y_continuous(labels = comma) +
  labs(title = "NJ's Enrollment Pyramid Shows the Pre-K Boom",
       subtitle = "Pre-K (87K) nearly matches Kindergarten (89K) in 2025",
       x = "Grade Level", y = "Students") +
  theme_nj() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## 15. NJ's poverty gap: Passaic vs Westfield

Passaic City has 98% of students on free/reduced lunch. Nearby Westfield has under 2%. This 96-point gap captures NJ's extreme wealth inequality.

```{r poverty-gap}
# Compare highest and lowest FRL districts
frl_all <- enr_current %>%
  filter(is_district, !is_charter,
         subgroup == "free_reduced_lunch", grade_level == "TOTAL",
         !is.na(pct), n_students >= 100) %>%
  arrange(desc(pct))

top_5 <- frl_all %>% head(5) %>% mutate(group = "Highest FRL")
bottom_5 <- frl_all %>% tail(5) %>% mutate(group = "Lowest FRL")
frl_extremes <- bind_rows(top_5, bottom_5) %>%
  mutate(district_label = reorder(district_name, pct))

stopifnot(nrow(frl_extremes) > 0)
frl_extremes %>% select(district_name, n_students, pct, group)

ggplot(frl_extremes, aes(x = district_label, y = pct * 100, fill = group)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("Highest FRL" = nj_colors["black"],
                               "Lowest FRL" = nj_colors["asian"])) +
  labs(title = "NJ's Poverty Gap: 96 Points Between Districts",
       subtitle = "Passaic City (98% FRL) vs Westfield (under 2%)",
       x = "", y = "Percent Free/Reduced Lunch", fill = "") +
  theme_nj()
```

```{r session-info}
sessionInfo()
```
