---
title: "New Jersey Enrollment Insights"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{New Jersey Enrollment Insights}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 6,
  out.width = "100%",
  message = FALSE,
  warning = FALSE,
  eval = FALSE
)
```

```{r load-packages}
library(njschooldata)
library(ggplot2)
library(dplyr)
library(scales)
```

```{r theme}
theme_nj <- function() {
  theme_minimal(base_size = 14) +
    theme(
      plot.title = element_text(face = "bold", size = 16),
      plot.subtitle = element_text(color = "gray40"),
      panel.grid.minor = element_blank(),
      legend.position = "bottom"
    )
}

colors <- c("total" = "#2C3E50", "white" = "#3498DB", "black" = "#E74C3C",
            "hispanic" = "#F39C12", "asian" = "#9B59B6", "charter" = "#1ABC9C")
```

```{r fetch-data, cache = TRUE}
# Fetch recent years of enrollment data
years <- 2015:2025
enr_all <- purrr::map_df(years, ~{
  tryCatch(fetch_enr(.x, tidy = TRUE, use_cache = TRUE), error = function(e) {
    warning("Failed to fetch enrollment for year ", .x, ": ", conditionMessage(e))
    NULL
  })
})

enr_current <- fetch_enr(2025, tidy = TRUE, use_cache = TRUE)
enr_2015 <- fetch_enr(2015, tidy = TRUE, use_cache = TRUE)
```

## 1. New Jersey educates 1.4 million students

New Jersey has one of the largest public school systems in the country, with enrollment holding relatively steady over the past decade.

```{r statewide-enrollment}
state_total <- enr_all %>%
  filter(is_state, subgroup == "total_enrollment", grade_level == "TOTAL")
stopifnot(nrow(state_total) > 0)

ggplot(state_total, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.5, color = colors["total"]) +
  geom_point(size = 3, color = colors["total"]) +
  scale_y_continuous(labels = comma, limits = c(0, NA)) +
  labs(title = "New Jersey Educates 1.4 Million Students",
       subtitle = "Statewide public school enrollment has held steady",
       x = "School Year", y = "Students") +
  theme_nj()
```

## 2. Newark leads the charter school revolution

Newark has one of the highest charter school enrollment rates in the nation - over 30% of students attend charter schools.

```{r newark-charter}
newark <- enr_all %>%
  filter(grepl("Newark", district_name, ignore.case = TRUE),
         is_district | is_charter,
         subgroup == "total_enrollment", grade_level == "TOTAL") %>%
  mutate(sector = ifelse(is_charter, "Charter", "Traditional"))

newark_summary <- newark %>%
  group_by(end_year, sector) %>%
  summarize(n_students = sum(n_students, na.rm = TRUE), .groups = "drop")
stopifnot(nrow(newark_summary) > 0)

ggplot(newark_summary, aes(x = end_year, y = n_students, fill = sector)) +
  geom_area(alpha = 0.8) +
  scale_y_continuous(labels = comma) +
  scale_fill_manual(values = c("Charter" = colors["charter"], "Traditional" = colors["total"])) +
  labs(title = "Newark Leads the Charter School Revolution",
       subtitle = "Over 30% of Newark students now attend charter schools",
       x = "School Year", y = "Students", fill = "") +
  theme_nj()
```

## 3. Hispanic students are the fastest-growing group

Hispanic enrollment has grown from 20% to nearly 30% of all NJ students in two decades.

```{r hispanic-growth}
hispanic <- enr_all %>%
  filter(is_state, subgroup == "hispanic", grade_level == "TOTAL")
stopifnot(nrow(hispanic) > 0)

ggplot(hispanic, aes(x = end_year, y = pct * 100)) +
  geom_line(linewidth = 1.5, color = colors["hispanic"]) +
  geom_point(size = 3, color = colors["hispanic"]) +
  labs(title = "Hispanic Students are the Fastest-Growing Group",
       subtitle = "From 20% to nearly 30% of all NJ students",
       x = "School Year", y = "Percent Hispanic") +
  theme_nj()
```

## 4. The Big Three: Newark, Jersey City, and Paterson

New Jersey's three largest districts educate over 100,000 students combined - nearly 8% of the state.

```{r big-three}
big_three <- c("Newark", "Jersey City", "Paterson")
big_three_trend <- enr_all %>%
  filter(is_district,
         grepl(paste(big_three, collapse = "|"), district_name, ignore.case = TRUE),
         subgroup == "total_enrollment", grade_level == "TOTAL")
stopifnot(nrow(big_three_trend) > 0)

ggplot(big_three_trend, aes(x = end_year, y = n_students, color = district_name)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_y_continuous(labels = comma) +
  labs(title = "The Big Three: Newark, Jersey City, and Paterson",
       subtitle = "Combined enrollment of over 100,000 students",
       x = "School Year", y = "Students", color = "") +
  theme_nj()
```

## 5. COVID hit kindergarten hard

New Jersey lost nearly 10% of kindergartners in 2021 - and enrollment hasn't fully recovered.

```{r covid-kindergarten}
k_trend <- enr_all %>%
  filter(is_state, subgroup == "total_enrollment",
         grade_level %in% c("PK", "K", "01", "06", "12")) %>%
  mutate(grade_label = case_when(
    grade_level == "PK" ~ "Pre-K",
    grade_level == "K" ~ "Kindergarten",
    grade_level == "01" ~ "Grade 1",
    grade_level == "06" ~ "Grade 6",
    grade_level == "12" ~ "Grade 12",
    TRUE ~ grade_level
  ))
stopifnot(nrow(k_trend) > 0)

ggplot(k_trend, aes(x = end_year, y = n_students, color = grade_label)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  geom_vline(xintercept = 2021, linetype = "dashed", color = "red", alpha = 0.5) +
  scale_y_continuous(labels = comma) +
  labs(title = "COVID Hit New Jersey Kindergarten Hard",
       subtitle = "Lost nearly 10% of kindergartners in 2021 - still recovering",
       x = "School Year", y = "Students", color = "") +
  theme_nj()
```

## 6. Free and reduced lunch eligibility reveals deep inequality

Some districts have nearly all students qualifying for free or reduced lunch, while affluent suburbs have under 5%.

```{r free-reduced-lunch}
frl <- enr_current %>%
  filter(is_district, subgroup == "free_reduced_lunch", grade_level == "TOTAL",
         !is.na(pct), n_students >= 100) %>%
  arrange(desc(pct)) %>%
  head(15) %>%
  mutate(district_label = reorder(district_name, pct))
stopifnot(nrow(frl) > 0)

ggplot(frl, aes(x = district_label, y = pct * 100)) +
  geom_col(fill = colors["total"]) +
  coord_flip() +
  labs(title = "Free/Reduced Lunch Eligibility Reveals Deep Inequality",
       subtitle = "Top 15 districts by share of students qualifying for free/reduced lunch",
       x = "", y = "Percent Free/Reduced Lunch") +
  theme_nj()
```

## 7. White student share has declined dramatically

White students went from majority to minority status in NJ public schools over the past two decades.

```{r demographic-shift}
demo <- enr_all %>%
  filter(is_state, subgroup %in% c("white", "hispanic", "black", "asian"),
         grade_level == "TOTAL") %>%
  mutate(subgroup = factor(subgroup, levels = c("white", "hispanic", "black", "asian")))
stopifnot(nrow(demo) > 0)

ggplot(demo, aes(x = end_year, y = pct * 100, color = subgroup)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_color_manual(values = colors, labels = c("White", "Hispanic", "Black", "Asian")) +
  labs(title = "White Student Share Has Declined Dramatically",
       subtitle = "NJ public schools are now majority-minority",
       x = "School Year", y = "Percent of Students", color = "") +
  theme_nj()
```

## 8. English Language Learners concentrated in urban areas

ELL students make up over 20% in some districts but under 1% in most suburban districts.

```{r ell-concentration}
ell <- enr_current %>%
  filter(is_district, subgroup == "lep", grade_level == "TOTAL",
         !is.na(pct), n_students >= 50) %>%
  arrange(desc(pct)) %>%
  head(15) %>%
  mutate(district_label = reorder(district_name, pct))
stopifnot(nrow(ell) > 0)

ggplot(ell, aes(x = district_label, y = pct * 100)) +
  geom_col(fill = colors["hispanic"]) +
  coord_flip() +
  labs(title = "English Learners Concentrated in Urban Areas",
       subtitle = "Some districts have over 20% ELL students",
       x = "", y = "Percent English Language Learners") +
  theme_nj()
```

## 9. Top 10 districts educate 20% of all students

Concentration at the top: just 10 districts out of 600+ serve one-fifth of all NJ students.

```{r top-10-districts}
top_10 <- enr_current %>%
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL") %>%
  arrange(desc(n_students)) %>%
  head(10) %>%
  mutate(district_label = reorder(district_name, n_students))
stopifnot(nrow(top_10) > 0)

ggplot(top_10, aes(x = district_label, y = n_students)) +
  geom_col(fill = colors["total"]) +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  labs(title = "Top 10 Districts Educate 20% of All Students",
       subtitle = "Just 10 out of 600+ districts serve one-fifth of NJ students",
       x = "", y = "Students") +
  theme_nj()
```

## 10. Multiracial students are NJ's fastest-growing identity

Multiracial enrollment has surged as more families identify with multiple races — a nationwide trend reshaping school demographics.

```{r multiracial-growth}
multiracial <- enr_all %>%
  filter(is_state, subgroup == "multiracial", grade_level == "TOTAL")
stopifnot(nrow(multiracial) > 0)

ggplot(multiracial, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.5, color = colors["charter"]) +
  geom_point(size = 3, color = colors["charter"]) +
  scale_y_continuous(labels = comma) +
  labs(title = "Multiracial Students Are NJ's Fastest-Growing Identity",
       subtitle = "Multiracial enrollment has surged across the state",
       x = "School Year", y = "Multiracial Students") +
  theme_nj()
```
