---
title: "New Jersey Enrollment Insights"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{New Jersey Enrollment Insights}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 6,
  out.width = "100%",
  message = FALSE,
  warning = FALSE,
  cache = TRUE
)
```

```{r load-packages}
library(njschooldata)
library(ggplot2)
library(dplyr)
library(scales)
```

```{r theme}
theme_nj <- function() {
  theme_minimal(base_size = 14) +
    theme(
      plot.title = element_text(face = "bold", size = 16),
      plot.subtitle = element_text(color = "gray40"),
      panel.grid.minor = element_blank(),
      legend.position = "bottom"
    )
}

colors <- c("total" = "#2C3E50", "white" = "#3498DB", "black" = "#E74C3C",
            "hispanic" = "#F39C12", "asian" = "#9B59B6", "charter" = "#1ABC9C",
            "frl" = "#E67E22")
```

```{r fetch-data, cache = TRUE}
# Fetch 2015-2024 enrollment data
enr_all <- purrr::map_df(2015:2024, ~{
  tryCatch(
    fetch_enr(.x, tidy = TRUE, use_cache = TRUE),
    error = function(e) { warning(paste("Failed year", .x, ":", e$message)); NULL }
  )
})

enr_current <- fetch_enr(2024, tidy = TRUE, use_cache = TRUE)
```

## 1. New Jersey educates 1.38 million students

Statewide public school enrollment has held remarkably steady over the past decade, hovering between 1.36 and 1.38 million.

```{r statewide-enrollment}
state_total <- enr_all %>%
  filter(is_state, subgroup == "total_enrollment", grade_level == "TOTAL")

stopifnot(nrow(state_total) > 0)

print(state_total %>% select(end_year, n_students))

ggplot(state_total, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.5, color = colors["total"]) +
  geom_point(size = 3, color = colors["total"]) +
  scale_y_continuous(labels = comma, limits = c(0, NA)) +
  labs(title = "New Jersey Educates 1.38 Million Students",
       subtitle = "Statewide public school enrollment has held steady since 2015",
       x = "School Year", y = "Students") +
  theme_nj()
```

## 2. Charter schools have grown to serve 61,000+ students

New Jersey's 84 charter school districts educated over 61,000 students in 2024, up from the early 2010s. Charter schools operate as independent districts under county code 80.

```{r charter-growth}
charter_trend <- enr_all %>%
  filter(is_charter, is_district, subgroup == "total_enrollment", grade_level == "TOTAL") %>%
  group_by(end_year) %>%
  summarize(charter_students = sum(n_students, na.rm = TRUE),
            n_charters = n(), .groups = "drop")

stopifnot(nrow(charter_trend) > 0)

print(charter_trend)

ggplot(charter_trend, aes(x = end_year, y = charter_students)) +
  geom_line(linewidth = 1.5, color = colors["charter"]) +
  geom_point(size = 3, color = colors["charter"]) +
  scale_y_continuous(labels = comma) +
  labs(title = "Charter Schools Serve 61,000+ Students",
       subtitle = "84 charter districts operate across New Jersey",
       x = "School Year", y = "Charter Students") +
  theme_nj()
```

## 3. Hispanic students are the fastest-growing group

Hispanic enrollment grew from 25.3% to 34.1% of all NJ students between 2015 and 2024 -- a gain of nearly 125,000 students.

```{r hispanic-growth}
hispanic <- enr_all %>%
  filter(is_state, subgroup == "hispanic", grade_level == "TOTAL")

stopifnot(nrow(hispanic) > 0)

print(hispanic %>% select(end_year, n_students, pct) %>% mutate(pct = round(pct * 100, 1)))

ggplot(hispanic, aes(x = end_year, y = pct * 100)) +
  geom_line(linewidth = 1.5, color = colors["hispanic"]) +
  geom_point(size = 3, color = colors["hispanic"]) +
  labs(title = "Hispanic Students are the Fastest-Growing Group",
       subtitle = "From 25% to 34% of all NJ students in a decade",
       x = "School Year", y = "Percent Hispanic") +
  theme_nj()
```

## 4. The Big Three: Newark, Jersey City, and Paterson

Newark Public School District leads with 42,600 students, followed by Jersey City (26,023) and Paterson (24,090). Combined, these three districts educate nearly 93,000 students.

```{r big-three}
big_three_trend <- enr_all %>%
  filter(is_district, !is_charter,
         grepl("^Newark Public|^Jersey City Public|^Paterson Public",
               district_name, ignore.case = TRUE),
         subgroup == "total_enrollment", grade_level == "TOTAL")

stopifnot(nrow(big_three_trend) > 0)
stopifnot(big_three_trend %>% filter(end_year == 2024) %>% nrow() == 3)

print(big_three_trend %>% filter(end_year == 2024) %>%
        select(district_name, n_students) %>% arrange(desc(n_students)))

ggplot(big_three_trend, aes(x = end_year, y = n_students, color = district_name)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_y_continuous(labels = comma) +
  labs(title = "The Big Three: Newark, Jersey City, and Paterson",
       subtitle = "Combined enrollment of nearly 93,000 students",
       x = "School Year", y = "Students", color = "") +
  theme_nj()
```

## 5. COVID hit kindergarten hard

New Jersey lost over 8,000 kindergartners in 2021 (a 9% drop), and enrollment only partially recovered by 2024.

```{r covid-kindergarten}
k_trend <- enr_all %>%
  filter(is_state, subgroup == "total_enrollment", grade_level == "K") %>%
  group_by(end_year) %>%
  summarize(n_students = sum(n_students, na.rm = TRUE), .groups = "drop")

stopifnot(nrow(k_trend) > 0)

print(k_trend)

ggplot(k_trend, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.5, color = colors["total"]) +
  geom_point(size = 3, color = colors["total"]) +
  geom_vline(xintercept = 2021, linetype = "dashed", color = "red", alpha = 0.5) +
  scale_y_continuous(labels = comma) +
  labs(title = "COVID Hit Kindergarten Hard",
       subtitle = "Lost 9% of kindergartners in 2021 -- partial recovery by 2024",
       x = "School Year", y = "Kindergarten Students") +
  theme_nj()
```

## 6. Free/reduced lunch enrollment tracks economic cycles

About 38% of NJ students qualify for free or reduced-price lunch. The rate dipped during COVID reporting disruptions, then rebounded.

```{r frl-trend}
frl_state <- enr_all %>%
  filter(is_state, subgroup == "free_reduced_lunch", grade_level == "TOTAL")

stopifnot(nrow(frl_state) > 0)

print(frl_state %>% select(end_year, n_students, pct) %>% mutate(pct = round(pct * 100, 1)))

ggplot(frl_state, aes(x = end_year, y = pct * 100)) +
  geom_line(linewidth = 1.5, color = colors["frl"]) +
  geom_point(size = 3, color = colors["frl"]) +
  labs(title = "Free/Reduced Lunch Rate Tracks Economic Cycles",
       subtitle = "About 38% of students qualify -- dipped during COVID, now rebounding",
       x = "School Year", y = "Percent Free/Reduced Lunch") +
  theme_nj()
```

## 7. White student share has declined sharply

White students went from 44.2% in 2015 to 37.6% in 2024. Meanwhile, Hispanic students surpassed Black students to become the second-largest group.

```{r demographic-shift}
demo <- enr_all %>%
  filter(is_state, subgroup %in% c("white", "hispanic", "black", "asian"),
         grade_level == "TOTAL") %>%
  mutate(subgroup = factor(subgroup, levels = c("white", "hispanic", "black", "asian")))

stopifnot(nrow(demo) > 0)

print(demo %>% filter(end_year == 2024) %>%
        select(subgroup, n_students, pct) %>%
        mutate(pct = round(pct * 100, 1)) %>% arrange(desc(pct)))

ggplot(demo, aes(x = end_year, y = pct * 100, color = subgroup)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_color_manual(values = colors, labels = c("White", "Hispanic", "Black", "Asian")) +
  labs(title = "White Student Share Has Declined Sharply",
       subtitle = "NJ demographics shifting: Hispanic enrollment now nearly equals White",
       x = "School Year", y = "Percent of Students", color = "") +
  theme_nj()
```

## 8. English learners concentrated in specific districts

Some districts have over 40% LEP (Limited English Proficient) students, while most districts have under 5%. The disparities are stark.

```{r ell-concentration}
ell <- enr_current %>%
  filter(is_district, subgroup == "lep", grade_level == "TOTAL",
         !is.na(pct), n_students >= 50) %>%
  arrange(desc(pct)) %>%
  head(15) %>%
  mutate(district_label = reorder(district_name, pct))

stopifnot(nrow(ell) > 0)

print(ell %>% select(district_name, n_students, pct) %>%
        mutate(pct = round(pct * 100, 1)))

ggplot(ell, aes(x = district_label, y = pct * 100)) +
  geom_col(fill = colors["hispanic"]) +
  coord_flip() +
  labs(title = "English Learners Concentrated in Specific Districts",
       subtitle = "Some districts have over 40% LEP students",
       x = "", y = "Percent English Language Learners") +
  theme_nj()
```

## 9. Top 10 districts educate 15% of all students

Just 10 out of 667 districts serve about one in seven NJ students. Newark alone educates 42,600.

```{r top-10-districts}
top_10 <- enr_current %>%
  filter(is_district, !is_charter,
         subgroup == "total_enrollment", grade_level == "TOTAL") %>%
  arrange(desc(n_students)) %>%
  head(10) %>%
  mutate(district_label = reorder(district_name, n_students))

stopifnot(nrow(top_10) == 10)

print(top_10 %>% select(district_name, n_students))

ggplot(top_10, aes(x = district_label, y = n_students)) +
  geom_col(fill = colors["total"]) +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  labs(title = "Top 10 Districts Educate 15% of All Students",
       subtitle = "Just 10 out of 667 districts serve 1 in 7 NJ students",
       x = "", y = "Students") +
  theme_nj()
```

## 10. Free/reduced lunch varies from 98% to under 1%

The gap between the highest-poverty and lowest-poverty districts is enormous. Some districts approach 100% free/reduced lunch.

```{r frl-districts}
frl_districts <- enr_current %>%
  filter(is_district, subgroup == "free_reduced_lunch", grade_level == "TOTAL",
         !is.na(pct), n_students >= 100) %>%
  arrange(desc(pct)) %>%
  head(15) %>%
  mutate(district_label = reorder(district_name, pct))

stopifnot(nrow(frl_districts) > 0)

print(frl_districts %>% select(district_name, n_students, pct) %>%
        mutate(pct = round(pct * 100, 1)))

ggplot(frl_districts, aes(x = district_label, y = pct * 100)) +
  geom_col(fill = colors["frl"]) +
  coord_flip() +
  labs(title = "Free/Reduced Lunch Varies from 98% to Under 1%",
       subtitle = "Highest-poverty districts approach 100%",
       x = "", y = "Percent Free/Reduced Lunch") +
  theme_nj()
```

## 11. Pre-K enrollment has more than doubled since 2015

New Jersey's universal pre-K expansion grew from 35,583 students in 2015 to 83,463 in 2024 -- a 135% increase.

```{r prek-growth}
prek <- enr_all %>%
  filter(is_state, subgroup == "total_enrollment", grade_level == "PK") %>%
  group_by(end_year) %>%
  summarize(n_students = sum(n_students, na.rm = TRUE), .groups = "drop")

stopifnot(nrow(prek) > 0)

print(prek)

ggplot(prek, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.5, color = colors["hispanic"]) +
  geom_point(size = 3, color = colors["hispanic"]) +
  scale_y_continuous(labels = comma) +
  labs(title = "Pre-K Enrollment Has More Than Doubled Since 2015",
       subtitle = "From 35,583 to 83,463 students -- a 135% increase",
       x = "School Year", y = "Pre-K Students") +
  theme_nj()
```

## 12. Boys outnumber girls in NJ public schools

A consistent 51.4-48.6 split favoring male students across all years.

```{r gender-balance}
gender <- enr_all %>%
  filter(is_state, subgroup %in% c("male", "female"), grade_level == "TOTAL")

stopifnot(nrow(gender) > 0)

print(gender %>% filter(end_year == 2024) %>%
        select(subgroup, n_students, pct) %>%
        mutate(pct = round(pct * 100, 1)))

ggplot(gender, aes(x = end_year, y = pct * 100, color = subgroup)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_color_manual(values = c("male" = colors["total"], "female" = "#E91E63"),
                     labels = c("Female", "Male")) +
  labs(title = "Boys Outnumber Girls in NJ Public Schools",
       subtitle = "Consistent 51-49 split favoring male students",
       x = "School Year", y = "Percent of Students", color = "") +
  theme_nj()
```

## 13. Black student enrollment declined 8% since 2015

Black student enrollment dropped from 217,179 in 2015 to 199,088 in 2024, a decline of about 18,000 students.

```{r black-decline}
black <- enr_all %>%
  filter(is_state, subgroup == "black", grade_level == "TOTAL")

stopifnot(nrow(black) > 0)

print(black %>% select(end_year, n_students, pct) %>%
        mutate(pct = round(pct * 100, 1)))

ggplot(black, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.5, color = colors["black"]) +
  geom_point(size = 3, color = colors["black"]) +
  scale_y_continuous(labels = comma) +
  labs(title = "Black Student Enrollment Declined 8% Since 2015",
       subtitle = "From 217,179 to 199,088 -- a loss of ~18,000 students",
       x = "School Year", y = "Black Students") +
  theme_nj()
```

## 14. Asian and Black enrollment converging but have not crossed

Asian enrollment grew from 130K to 143K while Black enrollment fell from 217K to 199K. The gap is narrowing but Asian students have not yet overtaken Black students.

```{r asian-black-crossover}
asian_black <- enr_all %>%
  filter(is_state, subgroup %in% c("asian", "black"), grade_level == "TOTAL")

stopifnot(nrow(asian_black) > 0)

print(asian_black %>%
        select(end_year, subgroup, n_students) %>%
        tidyr::pivot_wider(names_from = subgroup, values_from = n_students))

ggplot(asian_black, aes(x = end_year, y = n_students, color = subgroup)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_y_continuous(labels = comma) +
  scale_color_manual(values = c("asian" = colors["asian"], "black" = colors["black"]),
                     labels = c("Asian", "Black")) +
  labs(title = "Asian and Black Enrollment Converging",
       subtitle = "Gap narrowing: 143K Asian vs 199K Black in 2024",
       x = "School Year", y = "Students", color = "") +
  theme_nj()
```

## 15. 335 districts have fewer than 1,000 students

Small districts dominate NJ's fragmented school system. Over half of all districts serve fewer than 1,000 students each.

```{r small-districts}
small_districts <- enr_current %>%
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL") %>%
  mutate(size_category = case_when(
    n_students < 500 ~ "Under 500",
    n_students < 1000 ~ "500-999",
    n_students < 2500 ~ "1,000-2,499",
    n_students < 5000 ~ "2,500-4,999",
    n_students < 10000 ~ "5,000-9,999",
    TRUE ~ "10,000+"
  )) %>%
  count(size_category) %>%
  mutate(size_category = factor(size_category,
         levels = c("Under 500", "500-999", "1,000-2,499", "2,500-4,999", "5,000-9,999", "10,000+")))

stopifnot(nrow(small_districts) > 0)

print(small_districts)

ggplot(small_districts, aes(x = size_category, y = n)) +
  geom_col(fill = colors["total"]) +
  labs(title = "335 Districts Have Fewer Than 1,000 Students",
       subtitle = "Small districts dominate NJ's fragmented school system",
       x = "District Size", y = "Number of Districts") +
  theme_nj()
```

## Session Info

```{r session-info}
sessionInfo()
```
