---
title: "New Jersey English Learners"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{New Jersey English Learners}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 6,
  out.width = "100%",
  message = FALSE,
  warning = FALSE,
  # Default OFF: only the expensive fetch chunks cache. Plot chunks must
  # re-evaluate every build so knitr never replays a stale cached figure.
  cache = FALSE
)
```

`fetch_ell()` returns the **English Learner population** — how many students are
identified as English Learners (formerly "Limited English Proficient", now
"Multilingual Learners" in NJ DOE parlance) and what share of enrollment they
make up — at the state, district, and school level. The numbers come straight
from the NJ DOE Fall Enrollment files; no count is ever derived from a
percentage.

This is the EL *population* feature. EL *proficiency* (the WIDA ACCESS for ELLs
assessment) is a separate measure, available through `fetch_access()`.

```{r load-packages}
library(njschooldata)
library(ggplot2)
library(dplyr)
library(scales)
```

```{r theme}
theme_nj <- function() {
  theme_minimal(base_size = 14) +
    theme(
      plot.title = element_text(face = "bold", size = 16),
      plot.subtitle = element_text(color = "gray40"),
      panel.grid.minor = element_blank(),
      legend.position = "bottom"
    )
}
el_blue <- "#2C7FB8"
el_orange <- "#E67E22"
```

## 1. New Jersey's English Learner population has tripled since 2006

In the 2005-06 school year, New Jersey enrolled about 50,000 English Learners —
3.6% of all students. By 2024-25 that had grown to more than 155,000, **11.2% of
enrollment**. The EL count roughly tripled while total enrollment barely moved,
so the EL *share* of New Jersey's classrooms has more than doubled in two
decades.

```{r fetch-state, cache = TRUE}
# Statewide EL population, every year NJ publishes a count (2006-2025).
state_el <- fetch_ell_multi(2006:2025, use_cache = TRUE) %>%
  filter(is_state) %>%
  select(end_year, n_students, total_enrollment, pct_of_enrollment) %>%
  arrange(end_year)

stopifnot(nrow(state_el) > 0)
print(state_el)
```

```{r state-trend-chart}
stopifnot(nrow(state_el) > 0)

ggplot(state_el, aes(x = end_year, y = pct_of_enrollment)) +
  geom_line(color = el_blue, linewidth = 1.3) +
  geom_point(color = el_blue, size = 2.5) +
  scale_y_continuous(labels = label_percent(scale = 1), limits = c(0, NA)) +
  scale_x_continuous(breaks = seq(2006, 2025, 3)) +
  labs(
    title = "English Learners are a rising share of NJ enrollment",
    subtitle = "Statewide EL students as a percent of total enrollment, 2006-2025",
    x = NULL, y = "EL share of enrollment",
    caption = "Source: NJ DOE Fall Enrollment"
  ) +
  theme_nj()
```

## 2. In a dozen districts, two of every five students are English Learners

EL students are far from evenly spread. Statewide the share is about 11%, but in
the highest-EL districts more than 40% of students are English Learners.
Plainfield tops the list at roughly 45%, followed by a cluster of Central and
North Jersey cities.

```{r fetch-2025, cache = TRUE}
ell_2025 <- fetch_ell(2025, use_cache = TRUE)
stopifnot(nrow(ell_2025) > 0)

top_share <- ell_2025 %>%
  filter(is_district, total_enrollment > 2000, !is.na(n_students)) %>%
  arrange(desc(pct_of_enrollment)) %>%
  select(district_name, n_students, total_enrollment, pct_of_enrollment) %>%
  head(12)

print(top_share)
```

```{r el-hotspots-chart}
stopifnot(nrow(top_share) > 0)

ggplot(top_share, aes(x = reorder(district_name, pct_of_enrollment),
                      y = pct_of_enrollment)) +
  geom_col(fill = el_orange) +
  geom_text(aes(label = label_percent(scale = 1, accuracy = 0.1)(pct_of_enrollment)),
            hjust = -0.1, size = 3.5) +
  coord_flip() +
  scale_y_continuous(labels = label_percent(scale = 1),
                     expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "Where English Learners are concentrated",
    subtitle = "EL share of enrollment, districts over 2,000 students, 2024-25",
    x = NULL, y = "EL share of enrollment",
    caption = "Source: NJ DOE Fall Enrollment"
  ) +
  theme_nj()
```

## 3. The biggest EL populations are in New Jersey's largest cities

Share and headcount tell different stories. The districts with the *most* EL
students are the big urban systems: Newark enrolls more than 12,000 English
Learners, Elizabeth nearly 11,000, and Paterson over 8,000. Together a handful
of cities account for a large slice of the state's EL students.

```{r largest-el}
largest_el <- ell_2025 %>%
  filter(is_district, !is.na(n_students)) %>%
  arrange(desc(n_students)) %>%
  select(district_name, n_students, total_enrollment, pct_of_enrollment) %>%
  head(10)

stopifnot(nrow(largest_el) > 0)
print(largest_el)
```

```{r largest-el-chart}
stopifnot(nrow(largest_el) > 0)

ggplot(largest_el, aes(x = reorder(district_name, n_students), y = n_students)) +
  geom_col(fill = el_blue) +
  geom_text(aes(label = comma(round(n_students))), hjust = -0.1, size = 3.5) +
  coord_flip() +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "Largest English Learner populations in New Jersey",
    subtitle = "EL student headcount by district, 2024-25",
    x = NULL, y = "English Learners",
    caption = "Source: NJ DOE Fall Enrollment"
  ) +
  theme_nj()
```

## Data notes

- **Source:** NJ DOE Fall Enrollment files
  (`nj.gov/education/doedata/enr/`). The EL measure appears as `LEP` through
  2018-19, `English Learners` for 2019-20 through 2022-23, and
  `Multilingual Learners` from 2023-24 onward — `fetch_ell()` reconciles the
  renames.
- **Years:** 2006-2026 (`get_available_ell_years()`). Earlier enrollment files
  do not report an EL measure.
- **Counts vs. percent (COVID-era gap):** for 2020, 2021, and 2022 the NJ DOE
  *district* and *school* worksheets published only an EL **percent**, not a
  headcount. For those entity-years `n_students` is `NA` and only
  `pct_of_enrollment` is populated — the count is never back-derived from the
  percent. The statewide count is published every year. From 2023-24 on, real
  counts return at every level.
- **No suppression:** NJ does not suppress EL counts, so `n_students_lower` and
  `n_students_upper` equal `n_students` wherever a count is published.
- **Fractional values:** some counts and enrollment totals carry a `.5` — these
  are real shared-time / county-vocational FTE students, preserved exactly as
  published.
- **EL status:** NJ publishes a single current-EL headcount, so `el_status` is
  always `"current"` and `subgroup` is always `"total"`.

```{r session-info}
sessionInfo()
```
