---
title: "New Jersey School Facilities Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{New Jersey School Facilities Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 6,
  out.width = "100%",
  message = FALSE,
  warning = FALSE,
  cache = FALSE
)
```

```{r load-packages}
library(njschooldata)
library(ggplot2)
library(dplyr)
library(tidyr)
library(scales)

options(timeout = max(600, getOption("timeout")))
```

```{r theme}
theme_nj_facilities <- function() {
  theme_minimal(base_size = 13) +
    theme(
      plot.title = element_text(face = "bold", size = 16),
      plot.subtitle = element_text(color = "gray40"),
      plot.caption = element_text(color = "gray55", size = 9),
      panel.grid.minor = element_blank(),
      legend.position = "bottom"
    )
}

nj_green <- "#2F7D62"
nj_blue <- "#235789"
nj_gold <- "#B08B2E"
```

New Jersey facilities data is spread across public NJDOE, NJGIN, and NJSDA
sources instead of one statewide facilities warehouse. `fetch_facilities()` keeps
that fragmentation visible: every row carries a category, source agency, source
URL, source type, and vintage. Values stay character until you cast them for a
specific analysis, and missing source cells are dropped rather than filled.

```{r coverage}
available_facilities <- get_available_facilities()
stopifnot(nrow(available_facilities) > 0)
print(available_facilities)
```

## 1. The FY2026 SDA allocation table publishes 31 district grants

The NJDOE FY2026 SDA Emergent/Capital Maintenance allocation workbook lists 31
district grants totaling $50 million. Newark, Elizabeth, Jersey City, and
Paterson receive the largest allocations in this source.

```{r fetch-finance, cache = TRUE}
facilities_finance <- fetch_facilities("finance", use_cache = TRUE)
stopifnot(nrow(facilities_finance) > 0)

sda_allocations <- facilities_finance %>%
  filter(metric == "sda_grant_allocation") %>%
  mutate(amount = as.numeric(value)) %>%
  arrange(desc(amount))

sda_summary <- sda_allocations %>%
  summarize(
    districts = n_distinct(entity_id),
    total_allocation = sum(amount, na.rm = TRUE),
    largest_allocation = max(amount, na.rm = TRUE)
  )

sda_top <- sda_allocations %>%
  select(entity_id, entity_name, amount) %>%
  head(10)

stopifnot(nrow(sda_top) > 0)
print(sda_summary)
print(sda_top)
```

```{r sda-allocation-chart}
stopifnot(nrow(sda_top) > 0)

ggplot(sda_top, aes(x = reorder(entity_name, amount), y = amount)) +
  geom_col(fill = nj_blue) +
  geom_text(aes(label = dollar(amount, accuracy = 1)),
            hjust = -0.08, size = 3.4, color = nj_blue) +
  coord_flip() +
  scale_y_continuous(labels = label_dollar(scale = 1e-6, suffix = "M"),
                     expand = expansion(mult = c(0, 0.22))) +
  labs(
    title = "Largest FY2026 SDA emergent/capital maintenance allocations",
    subtitle = "Top 10 districts in the NJDOE allocation workbook",
    x = NULL,
    y = "Allocation",
    caption = "Source: NJDOE DistrictAllocationTable.xlsx via fetch_facilities('finance')"
  ) +
  theme_nj_facilities()
```

## 2. The 2024-25 Lead SOA workbook reports 67,086 district outlets tested

The district sheet in NJDOE's 2024-25 lead Statement of Assurance workbook
reports 67,086 outlets tested and 1,842 outlets exceeding the action level. The
source also contains one district row where exceeded outlets are greater than
tested outlets; the package preserves that source value and documents the caveat
instead of guessing a correction.

```{r fetch-lead, cache = TRUE}
facilities_lead <- fetch_facilities("environmental", use_cache = TRUE)
stopifnot(nrow(facilities_lead) > 0)

lead_wide <- facilities_lead %>%
  filter(entity_level == "district",
         metric %in% c("n_outlets_tested", "n_outlets_exceeded")) %>%
  select(entity_id, entity_name, metric, value) %>%
  pivot_wider(names_from = metric, values_from = value) %>%
  mutate(
    n_outlets_tested = as.numeric(n_outlets_tested),
    n_outlets_exceeded = as.numeric(n_outlets_exceeded)
  )

lead_summary <- lead_wide %>%
  summarize(
    districts = n(),
    outlets_tested = sum(n_outlets_tested, na.rm = TRUE),
    outlets_exceeded = sum(n_outlets_exceeded, na.rm = TRUE),
    pct_exceeded = outlets_exceeded / outlets_tested
  )

lead_top_tested <- lead_wide %>%
  filter(!is.na(n_outlets_tested)) %>%
  arrange(desc(n_outlets_tested)) %>%
  select(entity_id, entity_name, n_outlets_tested, n_outlets_exceeded) %>%
  head(10)

lead_source_anomalies <- lead_wide %>%
  filter(!is.na(n_outlets_tested),
         !is.na(n_outlets_exceeded),
         n_outlets_exceeded > n_outlets_tested) %>%
  select(entity_id, entity_name, n_outlets_tested, n_outlets_exceeded)

stopifnot(nrow(lead_top_tested) > 0)
print(lead_summary)
print(lead_top_tested)
print(lead_source_anomalies)
```

```{r lead-outlets-chart}
stopifnot(nrow(lead_top_tested) > 0)

ggplot(lead_top_tested,
       aes(x = reorder(entity_name, n_outlets_tested), y = n_outlets_tested)) +
  geom_col(fill = nj_green) +
  geom_text(aes(label = comma(n_outlets_tested)),
            hjust = -0.08, size = 3.4, color = nj_green) +
  coord_flip() +
  scale_y_continuous(labels = comma,
                     expand = expansion(mult = c(0, 0.2))) +
  labs(
    title = "Districts with the most drinking-water outlets tested",
    subtitle = "2024-25 NJDOE lead Statement of Assurance district sheet",
    x = NULL,
    y = "Outlets tested",
    caption = "Source: NJDOE 2024-25 Lead SOA workbook via fetch_facilities('environmental')"
  ) +
  theme_nj_facilities()
```

## 3. NJSDA active project pages expose a $1.89 billion capital portfolio

The NJSDA active capital program pages currently expose 14 project pages with
parsed cost fields totaling about $1.89 billion. Project attributes are reported
only where the public page states them explicitly; for example, added capacity
is populated for Bridgeton Senior High School and left missing elsewhere.

```{r fetch-projects, cache = TRUE}
facility_projects <- fetch_facilities("projects", use_cache = TRUE)
facility_capacity <- fetch_facilities("capacity", use_cache = TRUE)
facility_attributes <- fetch_facilities("attributes", use_cache = TRUE)

project_costs <- facility_projects %>%
  filter(metric == "total_estimated_project_cost") %>%
  mutate(cost = as.numeric(value)) %>%
  select(entity_id, entity_name, cost)

project_capacity <- facility_capacity %>%
  filter(metric == "added_capacity") %>%
  mutate(added_capacity = as.numeric(value)) %>%
  select(entity_id, added_capacity)

project_attributes <- facility_attributes %>%
  filter(metric %in% c("new_construction_sq_ft",
                       "renovation_sq_ft",
                       "year_constructed")) %>%
  select(entity_id, metric, value) %>%
  pivot_wider(names_from = metric, values_from = value) %>%
  mutate(
    new_construction_sq_ft = as.numeric(new_construction_sq_ft),
    renovation_sq_ft = as.numeric(renovation_sq_ft),
    year_constructed = as.numeric(year_constructed)
  )

project_profile <- project_costs %>%
  left_join(project_capacity, by = "entity_id") %>%
  left_join(project_attributes, by = "entity_id") %>%
  arrange(desc(cost))

project_summary <- project_profile %>%
  summarize(
    active_projects = n(),
    total_estimated_cost = sum(cost, na.rm = TRUE),
    projects_with_added_capacity = sum(!is.na(added_capacity)),
    added_capacity = sum(added_capacity, na.rm = TRUE)
  )

project_top_costs <- project_profile %>%
  select(entity_id, entity_name, cost, added_capacity, year_constructed) %>%
  head(10)

stopifnot(nrow(project_top_costs) > 0)
print(project_summary)
print(project_top_costs)
```

```{r active-project-cost-chart}
stopifnot(nrow(project_top_costs) > 0)

ggplot(project_top_costs, aes(x = reorder(entity_name, cost), y = cost)) +
  geom_col(fill = nj_gold) +
  geom_text(aes(label = dollar(cost, scale = 1e-6, suffix = "M", accuracy = 1)),
            hjust = -0.08, size = 3.4, color = nj_gold) +
  coord_flip() +
  scale_y_continuous(labels = label_dollar(scale = 1e-6, suffix = "M"),
                     expand = expansion(mult = c(0, 0.24))) +
  labs(
    title = "Largest active NJSDA projects by estimated cost",
    subtitle = "Parsed from public active project detail pages",
    x = NULL,
    y = "Estimated project cost",
    caption = "Source: NJSDA Capital Program pages via fetch_facilities('projects')"
  ) +
  theme_nj_facilities()
```

## 4. NJGIN publishes 3,736 school point locations

`fetch_facility_gis()` is the spatial companion to `fetch_facilities()`. It
returns NJGIN school points as an `sf` object when `sf` is installed, or a plain
data frame with latitude, longitude, and WKT when `sf = FALSE`.

```{r fetch-gis, cache = TRUE}
facility_points <- fetch_facility_gis("school_points", sf = FALSE, use_cache = TRUE)
stopifnot(nrow(facility_points) > 0)

point_summary <- facility_points %>%
  summarize(
    points = n(),
    points_with_coordinates = sum(!is.na(latitude) & !is.na(longitude)),
    school_types = n_distinct(school_type, na.rm = TRUE)
  )

point_types <- facility_points %>%
  count(school_type, sort = TRUE) %>%
  head(8)

print(point_summary)
print(point_types)
```

```{r school-points-map}
stopifnot(nrow(facility_points) > 0)

ggplot(facility_points, aes(x = longitude, y = latitude)) +
  geom_point(color = nj_green, alpha = 0.35, size = 0.7) +
  coord_quickmap() +
  labs(
    title = "NJGIN school point locations",
    subtitle = "Public school point layer, WGS84 coordinates",
    x = "Longitude",
    y = "Latitude",
    caption = "Source: NJGIN School Point Locations of NJ via fetch_facility_gis()"
  ) +
  theme_nj_facilities()
```

## Data notes

- **Inventory and closures:** NJDOE County/District/School Code workbook, current
  as of June 15, 2026. Closures come from the workbook's `Updates` sheet where
  `Action Taken` is `DELETED`.
- **GIS:** NJGIN School Point Locations of NJ ArcGIS FeatureServer. Geometry is
  returned separately by `fetch_facility_gis()`; latitude and longitude are also
  available as inventory metrics.
- **Projects, capacity, and attributes:** NJSDA public active capital program
  pages, accessed June 22, 2026. Fields are populated only when the page text
  explicitly states the value.
- **Finance:** NJDOE FY2026 SDA Emergent/Capital Maintenance allocation workbook.
- **Environmental:** NJDOE 2024-25 Lead Statement of Assurance workbook. The
  source contains one district row where exceeded outlets are greater than
  tested outlets; the package preserves the reported exceeded value and drops
  only impossible negative count cells.
- **Not yet shipped:** `condition` and `capital_needs` are part of the standard
  facilities vocabulary, but New Jersey does not yet have a verified populated
  public statewide bulk source in this package for those categories.

```{r session-info}
sessionInfo()
```
