---
title: "Restraint & Seclusion: A Special-Education Liability Metric"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Restraint & Seclusion: A Special-Education Liability Metric}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 6,
  out.width = "100%",
  message = FALSE,
  warning = FALSE,
  cache = FALSE
)
```

```{r load-packages}
library(njschooldata)
library(dplyr)
library(ggplot2)
```

Physical restraint and seclusion are among the most legally sensitive events a
school can record. New Jersey publishes them school-by-school in standalone DARS
(Discipline & Restraint) workbooks - separate from the violence/vandalism/HIB
data in `fetch_violence_vandalism_hib()`. `fetch_restraint_seclusion()` reads
them for **end_year 2023 (SY2022-23) and 2024 (SY2023-24)**.

The source is **school-level only** (no state/district aggregate) and is crossed
by student group, so every row is one school for one group. Small cells are
masked: `"*"` hides a value entirely and `"<5"` is a published range for 1-4
students. Both become `NA` here, never a guessed number - so the statewide sums
below are **floors** computed over unsuppressed cells, not exact totals.

## Which schools report the most restraint events?

The schoolwide totals (`subgroup == "total population"`, `grade_level ==
"TOTAL"`) show that reported restraint is overwhelmingly concentrated in
county special-services districts and out-of-district placement schools that
serve students with the most intensive needs.

```{r top-schools-data}
sw24 <- fetch_restraint_seclusion(2024) %>%
  filter(subgroup == "total population", grade_level == "TOTAL",
         !is.na(restraint_count))

stopifnot(nrow(sw24) > 0)

top10 <- sw24 %>%
  arrange(desc(restraint_count)) %>%
  slice(1:10) %>%
  mutate(label = paste0(school_name, " (", district_name, ")"))

# Print-before-plot: confirm the data feeding the chart.
top10 %>% select(district_name, school_name, restraint_count, seclusion_count)
```

```{r top-schools-chart}
ggplot(top10, aes(x = reorder(label, restraint_count), y = restraint_count)) +
  geom_col(fill = "#762a83") +
  geom_text(aes(label = restraint_count), hjust = -0.2, size = 3.6) +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(
    title = "New Jersey schools reporting the most restraint events, 2023-24",
    subtitle = "Schoolwide restraint counts; suppressed cells (<5, *) excluded",
    x = NULL,
    y = "Restraint events reported"
  ) +
  theme_minimal(base_size = 12)
```

## Restraint is concentrated among students with disabilities

The student-group breakdown lets us compare the schoolwide restraint-and-seclusion
total against the students-with-disabilities subgroup. Summing the unsuppressed
school cells statewide, students with disabilities account for roughly **half**
of all reported restraint-or-seclusion events - far above their share of
enrollment (about one in six students). This is the disproportionality that makes
the metric a board-level concern.

```{r swd-share-data}
rs24 <- fetch_restraint_seclusion(2024)

group_totals <- rs24 %>%
  filter(grade_level == "TOTAL",
         subgroup %in% c("total population", "students with disabilities")) %>%
  group_by(subgroup) %>%
  summarise(
    any_events = sum(any_restraint_seclusion_count, na.rm = TRUE),
    .groups = "drop"
  )

stopifnot(nrow(group_totals) == 2)

schoolwide_total <- group_totals$any_events[group_totals$subgroup == "total population"]
swd_total <- group_totals$any_events[group_totals$subgroup == "students with disabilities"]

share_tbl <- group_totals %>%
  mutate(pct_of_schoolwide = round(100 * any_events / schoolwide_total, 1))

# Print-before-plot.
share_tbl
```

```{r swd-share-chart}
ggplot(share_tbl, aes(x = reorder(subgroup, any_events), y = any_events,
                      fill = subgroup)) +
  geom_col(width = 0.6, show.legend = FALSE) +
  geom_text(aes(label = paste0(format(any_events, big.mark = ","),
                               "  (", pct_of_schoolwide, "%)")),
            hjust = -0.05, size = 4) +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.18))) +
  scale_fill_manual(values = c("students with disabilities" = "#d95f0e",
                               "total population" = "#2c7fb8")) +
  labs(
    title = "Restraint or seclusion events: students with disabilities vs. all students",
    subtitle = paste0("NJ 2023-24, summed over unsuppressed school cells; ",
                      "students with disabilities = ",
                      share_tbl$pct_of_schoolwide[
                        share_tbl$subgroup == "students with disabilities"],
                      "% of the schoolwide total"),
    x = NULL,
    y = "Reported restraint-or-seclusion events"
  ) +
  theme_minimal(base_size = 12)
```

## Notes

- **School-level only.** The DARS workbook has no district/state/county
  aggregate; `level` must be `"school"`. Statewide figures here are sums of the
  per-school cells.
- **Suppression -> `NA`.** `"*"` and `"<5"` both become `NA`; `"<5"` is never
  read as `5`. Statewide sums are therefore **floors** over unsuppressed cells -
  the true totals are higher, but the masked counts are 1-4 students each.
- **Student group.** The raw `Student Group` label is kept as `student_group`
  and split into normalized `subgroup` + `grade_level`. The schoolwide total is
  labelled `"Schoolwide"` in 2022-23 and `"School Total"` in 2023-24; both map to
  `subgroup == "total population"`.

```{r session-info}
sessionInfo()
```
