---
title: "Staff History: Educator Evaluations & the Certificated Workforce"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Staff History: Educator Evaluations & the Certificated Workforce}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 6,
  out.width = "100%",
  message = FALSE,
  warning = FALSE,
  cache = FALSE
)
```

```{r load-packages}
library(njschooldata)
library(dplyr)
library(tidyr)
library(ggplot2)
```

Two NJ DOE "doedata" staff sources sit outside the School Performance Reports and
reach back well before them. `fetch_staff_evaluations()` returns the summative
educator **evaluation rating distributions** NJ published for the three years it
ran the file (2014-2016), and `fetch_certificated_staff()` returns the deep
**certificated-staff FTE** series by position, race, and gender (2000-2008 and
2020-2026; the 2009-2019 intermediate files use a non-uniform layout and are not
supported). Every number below traces to a downloaded NJ DOE file: a `"*"`
suppression mask is `NA`, and a race column an era did not report is `NA`, never
zero.

## Where do teacher evaluation ratings land?

NJ's summative evaluation places each educator in one of four tiers. The
statewide teacher distribution (the published `is_state` row, available 2014 and
2015) clusters heavily at the top two tiers, and the **highly effective** share
jumped sharply between the two years.

```{r eval-data, cache=TRUE}
eval_state <- bind_rows(
  fetch_staff_evaluations(2014, level = "district"),
  fetch_staff_evaluations(2015, level = "district")
) %>%
  filter(is_state, staff_category == "teachers")
```

```{r eval-prep}
tiers <- c("ineffective", "partially_effective", "effective", "highly_effective")

eval_long <- eval_state %>%
  select(end_year, all_of(tiers)) %>%
  pivot_longer(all_of(tiers), names_to = "rating", values_to = "n") %>%
  group_by(end_year) %>%
  mutate(share = 100 * n / sum(n, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(rating = factor(rating, levels = tiers))

# Print-before-plot: confirm the data feeding the chart.
stopifnot(nrow(eval_long) > 0)
eval_long %>% arrange(end_year, rating) %>% as.data.frame()
```

```{r eval-chart}
ggplot(eval_long, aes(x = factor(end_year), y = share, fill = rating)) +
  geom_col(width = 0.65) +
  geom_text(aes(label = ifelse(share >= 3, sprintf("%.1f%%", share), "")),
            position = position_stack(vjust = 0.5), colour = "white", size = 3.4) +
  scale_fill_manual(
    values = c(ineffective = "#b2182b", partially_effective = "#ef8a62",
               effective = "#67a9cf", highly_effective = "#2166ac"),
    labels = c("Ineffective", "Partially effective", "Effective",
               "Highly effective")
  ) +
  labs(
    title = "Statewide teacher evaluation rating distribution",
    subtitle = "Share of evaluated NJ teachers in each summative tier",
    x = NULL, y = "Share of teachers (%)", fill = "Rating"
  ) +
  theme_minimal(base_size = 13)
```

Effective and highly effective together account for more than 97% of teachers in
both years, and highly effective alone climbs from about 24% to about 34% in a
single year. (The 2016 workbook publishes no statewide aggregate row, so the
clean statewide comparison runs 2014-2015.)

## The certificated teacher workforce, across two eras

`fetch_certificated_staff()` reports staff full-time-equivalents (FTE), so counts
are fractional. Pulling the statewide teacher total across both covered eras
shows a workforce that grew through the 2000s and has held roughly flat in the
2020s.

```{r workforce-data, cache=TRUE}
years <- c(2000:2008, 2020:2026)
teachers_state <- bind_rows(lapply(years, function(y) {
  fetch_certificated_staff(y, level = "state") %>%
    filter(position == "teachers", gender == "total") %>%
    transmute(end_year, white, total)
}))
```

```{r workforce-prep}
stopifnot(nrow(teachers_state) > 0)
teachers_state %>% as.data.frame()
```

```{r workforce-chart}
ggplot(teachers_state, aes(x = end_year, y = total)) +
  geom_col(fill = "#2c7fb8") +
  scale_x_continuous(breaks = years) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Statewide certificated teacher FTE",
    subtitle = "NJ DOE certificated-staff files (2009-2019 not published in a supported layout)",
    x = NULL, y = "Teacher FTE"
  ) +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The gap from 2009 to 2019 is a coverage gap, not a workforce collapse: those
years exist but in a drifting layout the fetcher refuses to parse rather than risk
misaligned values. The single low bar at 2006 is the value NJ published in that
year's file and is passed through as-is.

## The teaching workforce got less white

The same series carries race, so the white share of the teacher workforce is
directly computable across both eras. It has drifted down from roughly 88% at the
start of the 2000s toward about 81% in the mid-2020s.

```{r diversity-prep}
white_share <- teachers_state %>%
  mutate(pct_white = 100 * white / total)

stopifnot(nrow(white_share) > 0)
white_share %>%
  filter(end_year %in% c(2000, 2008, 2020, 2026)) %>%
  as.data.frame()
```

```{r diversity-chart}
ggplot(white_share, aes(x = end_year, y = pct_white)) +
  geom_line(colour = "#d95f0e", linewidth = 1) +
  geom_point(colour = "#d95f0e", size = 2) +
  scale_x_continuous(breaks = years) +
  labs(
    title = "White share of NJ certificated teachers",
    subtitle = "white FTE / total teacher FTE, statewide",
    x = NULL, y = "White share (%)"
  ) +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Note that the legacy era (2000-2008) reported a single combined Asian/Pacific
Islander group, so `fetch_certificated_staff()` carries that combined count in
`asian` and leaves `pacific_islander` and `two_or_more` as `NA` for those years;
the white share above is unaffected by that distinction.

```{r session-info}
sessionInfo()
```
