---
title: "School Environment: Instructional Time & Device Access"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{School Environment: Instructional Time & Device Access}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 6,
  out.width = "100%",
  message = FALSE,
  warning = FALSE,
  cache = FALSE
)
```

```{r load-packages}
library(njschooldata)
library(dplyr)
library(ggplot2)
```

Two levers a building leader directly controls - **how much time** students spend
in school and **whether every student has a device** - are reported school-by-school
in the NJ School Performance Reports but are rarely surfaced in a comparable way.
`fetch_school_day()` and `fetch_device_ratios()` expose them. Both are school-level
sheets (an attribute of a building, with no district/state aggregate), so every
example below filters to `is_school`.

## How long is the New Jersey school day?

`fetch_school_day()` returns the published `length_of_day` string (e.g.
`"6 Hrs. 25 Mins."`) alongside a derived numeric `length_of_day_minutes`. The
typical NJ school day in 2023-24 runs about six and a half hours, but the spread
is wide - from half-day programs near four hours to schools topping nine.

```{r school-day-data}
sd24 <- fetch_school_day(2024) %>%
  filter(is_school, !is.na(length_of_day_minutes))

# Print-before-plot: confirm the data feeding the chart.
nrow(sd24)
summary(sd24$length_of_day_minutes)
```

```{r school-day-hist}
ggplot(sd24, aes(x = length_of_day_minutes)) +
  geom_histogram(binwidth = 15, fill = "#2c7fb8", colour = "white") +
  geom_vline(xintercept = median(sd24$length_of_day_minutes),
             linetype = "dashed", colour = "#d95f0e", linewidth = 0.9) +
  scale_x_continuous(breaks = seq(240, 600, 60)) +
  labs(
    title = "Length of the New Jersey school day, 2023-24",
    subtitle = paste0("Each school's reported day length; dashed line = median (",
                      median(sd24$length_of_day_minutes), " minutes)"),
    x = "Length of school day (minutes)",
    y = "Number of schools"
  ) +
  theme_minimal(base_size = 13)
```

The longest and shortest reported days:

```{r school-day-extremes}
sd24 %>%
  arrange(desc(length_of_day_minutes)) %>%
  slice(c(1:3, (n() - 2):n())) %>%
  select(district_name, school_name, length_of_day, length_of_day_minutes)
```

## Device scarcity peaked in the first pandemic year

`fetch_device_ratios()` reports the student-to-computing-device ratio per school
(`students_per_device`; a value of 1 means 1:1). Tracking the share of schools at
1:1-or-better across the years the sheet exists (it is absent for SY2016-17 and
SY2019-20) shows that access was actually **tightest in 2020-21**, the first full
remote-learning year, before recovering - not the steady march to 1:1 one might
expect.

```{r device-trend-data}
device_years <- c(2018, 2019, 2021, 2022, 2023, 2024, 2025)

device_trend <- purrr::map_dfr(device_years, function(yr) {
  fetch_device_ratios(yr) %>%
    filter(is_school, !is.na(students_per_device)) %>%
    summarise(
      end_year   = yr,
      n_schools  = n(),
      pct_1to1   = round(100 * mean(students_per_device <= 1), 1),
      median_ratio = median(students_per_device)
    )
})

# Print-before-plot.
device_trend
```

```{r device-trend-chart}
ggplot(device_trend, aes(x = end_year, y = pct_1to1)) +
  geom_line(colour = "#2c7fb8", linewidth = 1.1) +
  geom_point(size = 3, colour = "#2c7fb8") +
  geom_text(aes(label = paste0(pct_1to1, "%")), vjust = -1, size = 3.6) +
  scale_x_continuous(breaks = device_years) +
  scale_y_continuous(limits = c(40, 85)) +
  labs(
    title = "Share of NJ schools at 1:1 student-to-device or better",
    subtitle = "Sheet absent for SY2019-20 (2020); access dips in 2020-21 then recovers",
    x = "School year (end)",
    y = "Percent of schools at 1:1 or better"
  ) +
  theme_minimal(base_size = 13)
```

## Notes

- Both fetchers are **school-level only**. The `SchoolDay` sheet covers 2017-2025
  (the 2017 sheet omits entity names; CDS ids are present). The `DeviceRatios`
  sheet covers 2018-2025 **except SY2016-17 and SY2019-20**.
- Durations and ratios are published as strings; the numeric `*_minutes` and
  `students_per_device` columns are a deterministic re-expression of the published
  value. Non-values (`"n/a"`, `"No devices reported"`) become `NA`, never a
  fabricated number.

```{r session-info}
sessionInfo()
```
