---
title: "Following the Money: 25 Years of Spending in South Orange-Maplewood"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Following the Money: 25 Years of Spending in South Orange-Maplewood}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 6,
  out.width = "100%",
  message = FALSE,
  warning = FALSE,
  # Default OFF: only the expensive fetch chunks cache. Plot chunks must
  # re-evaluate every build, or knitr replays stale figures (caschooldata, 2026-03).
  cache = FALSE
)
```

```{r load-packages}
library(njschooldata)
library(ggplot2)
library(dplyr)
library(scales)
library(purrr)
library(tibble)

# The arts section reads a School Performance Report workbook (~100 MB), so give
# the download more than R's stingy 60-second default.
options(timeout = max(600, getOption("timeout")))
```

```{r theme}
theme_nj <- function() {
  theme_minimal(base_size = 14) +
    theme(
      plot.title = element_text(face = "bold", size = 16),
      plot.subtitle = element_text(color = "gray40"),
      plot.caption = element_text(color = "gray55", size = 9),
      panel.grid.minor = element_blank(),
      legend.position = "bottom"
    )
}

somsd_navy   <- "#1F3A5F"
somsd_orange <- "#E67E22"
somsd_teal   <- "#16A085"
somsd_red    <- "#C0392B"
somsd_purple <- "#8E44AD"
```

The [Taxpayers' Guide to Educational Spending](https://www.nj.gov/education/guide/)
(TGES) is the New Jersey Department of Education's annual report on what every
district spends, broken down per pupil into classroom instruction, support
services, administration, plant operations, and more. `njschooldata` pulls it
straight from the state for every year back to 2001 (the pre-2011 editions were
branded the "Comparative Spending Guide").

This article follows one district, **South Orange-Maplewood (SOMSD)**, over a
quarter century. It's a single mid-sized K-12 district in Essex County, but the
patterns are the kind every NJ taxpayer eventually asks about: where the money
goes, why the benefits line keeps climbing, and who actually pays the bill.

`fetch_tges(year)` returns one tidy data frame per spending indicator:

```{r intro-example, cache = TRUE}
spending_2025 <- fetch_tges(2025)

# one tidy table per TGES indicator (CSG1 = total budgetary cost per pupil)
names(spending_2025)
```

Each table is district-level and long, with a three-year window per report:
two finalized `Actuals` years plus the current `Budgeted` year. SOMSD is
district code `4900`.

```{r intro-glimpse}
spending_2025[["CSG1"]] %>%
  filter(district_id == "4900") %>%
  select(district_name, end_year, calc_type, `Per Pupil costs`, `District rank`)
```

```{r fetch-data, cache = TRUE}
# Pull every published guide, 2001-2025. fetch_tges(2025) was already fetched
# above, so reuse it and download the rest. Drop (with a visible warning) any
# year that fails rather than letting one network hiccup kill the whole build.
reports <- list("2025" = spending_2025)
for (y in 2001:2024) {
  reports[[as.character(y)]] <- tryCatch(
    fetch_tges(y),
    error = function(e) {
      warning(paste("TGES", y, "failed:", conditionMessage(e)))
      NULL
    }
  )
}
reports <- reports[!vapply(reports, is.null, logical(1))]
reports <- reports[order(as.integer(names(reports)))]
```

```{r helpers}
SOMSD <- "4900"

# Pull one SOMSD value column across every report and keep one row per school
# year. Each actual recurs in consecutive guides; we keep the most recent
# (most finalized) report's figure. VITSTAT/CSG16 carry no calc_type, so the
# calc filter is skipped for them.
somsd_series <- function(table, value_col, calc = "Actuals") {
  map_dfr(names(reports), function(y) {
    d <- reports[[y]][[table]]
    if (is.null(d) || !value_col %in% names(d)) return(NULL)
    r <- filter(d, !is.na(district_id), district_id == SOMSD)
    if ("calc_type" %in% names(r)) r <- filter(r, calc_type == calc)
    if (nrow(r) == 0) return(NULL)
    tibble(report_year = as.integer(y), end_year = r$end_year, value = r[[value_col]])
  }) %>%
    group_by(end_year) %>%
    slice_max(report_year, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    arrange(end_year)
}
```

## 1. A decade of flat spending, then a sharp surge

For most of the 2010s, SOMSD's actual budgetary cost per pupil barely moved:
about \$14,300 in 2010 and \$14,700 in 2019, a real-terms decline once you
account for inflation. Then it broke loose. Actuals jumped from \$14,660 (2020)
to \$18,952 (2024), and the 2025 budget sets it at \$20,066.

```{r pp-data}
pp <- somsd_series("CSG1", "Per Pupil costs", "Actuals")
pp_budget_2025 <- somsd_series("CSG1", "Per Pupil costs", "Budgeted") %>%
  filter(end_year == 2025)

stopifnot(nrow(pp) > 0, nrow(pp_budget_2025) == 1)
pp
```

```{r pp-chart}
ggplot(pp, aes(end_year, value)) +
  annotate("rect", xmin = 2010, xmax = 2019, ymin = -Inf, ymax = Inf,
           fill = "gray85", alpha = 0.5) +
  annotate("text", x = 2014.5, y = 19500, label = "flat decade",
           color = "gray45", fontface = "italic") +
  geom_line(color = somsd_navy, linewidth = 1.2) +
  geom_point(color = somsd_navy, size = 2) +
  geom_point(data = pp_budget_2025, aes(end_year, value),
             color = somsd_orange, size = 3, shape = 17) +
  annotate("text", x = 2024.7, y = pp_budget_2025$value + 900,
           label = "2025\nbudgeted", color = somsd_orange, size = 3.3) +
  scale_y_continuous(labels = dollar) +
  scale_x_continuous(breaks = seq(2000, 2025, 5)) +
  labs(title = "SOMSD per-pupil spending: flat, then a post-2020 surge",
       subtitle = "Budgetary cost per pupil, actuals (2025 is budgeted)",
       x = NULL, y = "Cost per pupil",
       caption = "Source: NJ DOE Taxpayers' Guide to Educational Spending") +
  theme_nj()
```

## 2. Benefits ate the salary line

Here's the trend that drives every contentious school budget meeting: employee
benefits as a share of salaries. In SOMSD it has **more than doubled**, from
11.8% in 1999 to 24.4% in 2024. Health-plan premiums (SEHBP) are the main
driver, and they keep climbing faster than pay.

```{r benefits-data}
benefits <- somsd_series("CSG14", "% of Total Salaries", "Actuals")
stopifnot(nrow(benefits) > 0)
benefits
```

```{r benefits-chart}
ggplot(benefits, aes(end_year, value)) +
  geom_area(fill = somsd_orange, alpha = 0.25) +
  geom_line(color = somsd_orange, linewidth = 1.2) +
  geom_point(color = somsd_orange, size = 2) +
  scale_y_continuous(labels = percent_format(accuracy = 1),
                     limits = c(0, NA)) +
  scale_x_continuous(breaks = seq(2000, 2025, 5)) +
  labs(title = "Employee benefits as a share of salaries doubled",
       subtitle = "SOMSD benefit costs divided by total salaries, 1999-2024",
       x = NULL, y = "Benefits / salaries",
       caption = "Source: NJ DOE TGES (CSG14)") +
  theme_nj()
```

## 3. Who pays the bill is shifting

SOMSD is a property-tax town: local taxpayers funded about **85%** of the budget
in 2010. But the state's share has roughly doubled since then (from 12% to 26%),
while the local share fell to 70%. Federal money never tops a few percent.

```{r revenue-data}
revenue <- bind_rows(
  somsd_series("VITSTAT_TOTAL", "Revenue: Local %") %>% mutate(source = "Local taxes"),
  somsd_series("VITSTAT_TOTAL", "Revenue: State %")  %>% mutate(source = "State aid")
)
stopifnot(nrow(revenue) > 0)

revenue %>%
  tidyr::pivot_wider(id_cols = end_year, names_from = source, values_from = value)
```

```{r revenue-chart}
ggplot(revenue, aes(end_year, value, color = source)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_color_manual(values = c("Local taxes" = somsd_navy, "State aid" = somsd_teal)) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_x_continuous(breaks = seq(2010, 2024, 4)) +
  labs(title = "Local taxpayers still carry SOMSD, but state aid is rising",
       subtitle = "Share of revenue by source",
       x = NULL, y = "Share of revenue", color = NULL,
       caption = "Source: NJ DOE TGES Vital Statistics") +
  theme_nj()
```

## 4. Teacher pay up, class sizes down

The median teacher salary nearly doubled over the period, from about \$45,000 to
roughly \$84,000. Over the same stretch the student-to-teacher ratio drifted down
from 14:1 toward 11:1, so the district is paying more teachers, more.

```{r salary-data}
salary <- somsd_series("CSG16", "Teacher Salary")
ratio  <- somsd_series("CSG16", "Student/Teacher ratio")
stopifnot(nrow(salary) > 0, nrow(ratio) > 0)
salary %>%
  inner_join(ratio, by = "end_year", suffix = c("_salary", "_ratio")) %>%
  transmute(end_year, median_salary = value_salary, students_per_teacher = value_ratio)
```

```{r salary-chart}
ggplot(salary, aes(end_year, value)) +
  geom_line(color = somsd_purple, linewidth = 1.2) +
  geom_point(color = somsd_purple, size = 2) +
  scale_y_continuous(labels = dollar) +
  scale_x_continuous(breaks = seq(2000, 2025, 5)) +
  labs(title = "Median teacher salary nearly doubled",
       subtitle = "SOMSD median classroom-teacher salary",
       x = NULL, y = "Median salary",
       caption = "Source: NJ DOE TGES (CSG16)") +
  theme_nj()
```

```{r ratio-chart}
ggplot(ratio, aes(end_year, value)) +
  geom_line(color = somsd_teal, linewidth = 1.2) +
  geom_point(color = somsd_teal, size = 2) +
  scale_x_continuous(breaks = seq(2000, 2025, 5)) +
  labs(title = "...while the student-to-teacher ratio fell",
       subtitle = "Students per teacher",
       x = NULL, y = "Students per teacher",
       caption = "Source: NJ DOE TGES (CSG16)") +
  theme_nj()
```

## 5. Where each dollar goes

TGES breaks the budgetary cost per pupil into its components. In the 2025 budget,
classroom instruction is by far the biggest slice (about 61 cents of every
dollar), followed by support services, plant operations, and administration.

```{r category-data}
category_tables <- tribble(
  ~category,                  ~table,
  "Classroom instruction",    "CSG2",
  "Support services",         "CSG6",
  "Plant ops & maintenance",  "CSG10",
  "Administration",           "CSG8",
  "Extracurricular",          "CSG13"
)

categories <- map_dfr(seq_len(nrow(category_tables)), function(i) {
  d <- reports[["2025"]][[category_tables$table[i]]]
  r <- filter(d, district_id == SOMSD, end_year == 2025)
  tibble(category = category_tables$category[i],
         per_pupil = r[["Per Pupil costs"]][1])
})

stopifnot(nrow(categories) > 0, all(!is.na(categories$per_pupil)))
categories %>% arrange(desc(per_pupil))
```

```{r category-chart}
ggplot(categories, aes(reorder(category, per_pupil), per_pupil)) +
  geom_col(fill = somsd_navy) +
  geom_text(aes(label = dollar(per_pupil)), hjust = -0.1, size = 3.5) +
  coord_flip() +
  scale_y_continuous(labels = dollar, expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Where SOMSD spends, per pupil (2025 budget)",
       subtitle = "Classroom instruction dominates the budgetary cost per pupil",
       x = NULL, y = "Cost per pupil",
       caption = "Source: NJ DOE TGES (CSG2, CSG6, CSG8, CSG10, CSG13)") +
  theme_nj()
```

## 6. Drifting through the pack of big districts

TGES ranks each district against its peer group. SOMSD sits with the large K-12
districts (3,500+ students). In this ranking, **1 = the lowest spender**, so a
higher number means spending more than peers. SOMSD climbed toward the top of
the pack by 2010, slid to near the bottom during its flat decade as peers kept
growing, then recovered to mid-pack after the post-2020 surge.

```{r rank-data}
rank <- somsd_series("CSG1", "District rank", "Actuals")
stopifnot(nrow(rank) > 0)

# size of the peer group in the most recent guide, for context
n_peers <- reports[["2025"]][["CSG1"]] %>%
  filter(group == "G. K-12 / 3501 +", end_year == 2025,
         !is.na(district_id), district_id != "00NA") %>%
  nrow()
n_peers

rank
```

```{r rank-chart}
ggplot(rank, aes(end_year, value)) +
  geom_line(color = somsd_red, linewidth = 1.2) +
  geom_point(color = somsd_red, size = 2) +
  scale_x_continuous(breaks = seq(2000, 2024, 4)) +
  labs(title = "SOMSD's spending rank among large K-12 districts",
       subtitle = paste0("Rank within the ", n_peers,
                          "-district peer group (1 = lowest spender)"),
       x = NULL, y = "Spending rank",
       caption = "Source: NJ DOE TGES (CSG1). Peer-group composition shifts modestly year to year.") +
  theme_nj()
```

## 7. A closer look at the classroom dollar

TGES splits classroom instruction into three pieces: salaries and benefits,
supplies and textbooks, and "purchased services and other." Two things jump out.
Spending on books and supplies has been frozen near \$200 per pupil for 25 years,
a steep real-terms cut. Meanwhile purchased services, contracted and
out-of-district instruction, went from a rounding error (\$7 per pupil in 1999)
to \$1,804 in 2024, about 15 cents of every classroom dollar.

```{r classroom-data}
classroom_parts <- bind_rows(
  somsd_series("CSG3", "Per Pupil costs") %>% mutate(part = "Salaries & benefits"),
  somsd_series("CSG4", "Per Pupil costs") %>% mutate(part = "Supplies & textbooks"),
  somsd_series("CSG5", "Per Pupil costs") %>% mutate(part = "Purchased services & other")
) %>%
  mutate(part = factor(part, levels = c("Salaries & benefits",
                                        "Supplies & textbooks",
                                        "Purchased services & other")))
stopifnot(nrow(classroom_parts) > 0)

classroom_parts %>%
  filter(end_year %in% c(1999, 2008, 2016, 2024)) %>%
  select(end_year, part, value) %>%
  arrange(end_year, part)
```

```{r classroom-dollar-chart}
ggplot(classroom_parts, aes(end_year, value, fill = part)) +
  geom_area(alpha = 0.9) +
  scale_fill_manual(values = c("Salaries & benefits" = somsd_navy,
                               "Supplies & textbooks" = somsd_teal,
                               "Purchased services & other" = somsd_orange)) +
  scale_y_continuous(labels = dollar) +
  scale_x_continuous(breaks = seq(2000, 2024, 6)) +
  labs(title = "The classroom dollar: contracted services moved in",
       subtitle = "SOMSD classroom cost per pupil by component",
       x = NULL, y = "Cost per pupil", fill = NULL,
       caption = "Source: NJ DOE TGES (CSG3, CSG4, CSG5)") +
  theme_nj()
```

## 8. Facilities: a frozen line on aging buildings

Plant operations and maintenance, the cost of heating, cleaning, and repairing
school buildings, barely budged for over a decade. It hovered around \$1,800 to
\$2,000 per pupil from 2008 through 2022 even as total spending climbed, so its
**share of the budget fell from about 16% to 12%**. That is the kind of deferred
upkeep that eventually shows up as a capital referendum. (TGES tracks operating
cost, not the separate capital bonds districts float to rebuild.)

```{r facilities-data}
facilities <- somsd_series("CSG10", "Per Pupil costs") %>%
  select(end_year, facilities = value) %>%
  inner_join(somsd_series("CSG1", "Per Pupil costs") %>% select(end_year, total = value),
             by = "end_year") %>%
  mutate(share = facilities / total)
stopifnot(nrow(facilities) > 0)
facilities
```

```{r facilities-share-chart}
ggplot(facilities, aes(end_year, share)) +
  geom_line(color = somsd_red, linewidth = 1.2) +
  geom_point(color = somsd_red, size = 2) +
  scale_y_continuous(labels = percent_format(accuracy = 1), limits = c(0, NA)) +
  scale_x_continuous(breaks = seq(2000, 2024, 6)) +
  labs(title = "Facilities' slice of the budget shrank",
       subtitle = "Plant operations & maintenance as a share of budgetary cost per pupil",
       x = NULL, y = "Facilities share of budget",
       caption = "Source: NJ DOE TGES (CSG10 / CSG1)") +
  theme_nj()
```

## 9. What the spending buys: the arts

TGES tells you what a district spends; the School Performance Report tells you
what students actually take. In a town known for its arts scene, that second
question matters. `fetch_arts_enrollment()` pulls visual and performing arts
participation, and SOMSD runs well ahead of the state at every grade level. The
gap is widest in middle school: nearly every 6th-8th grader takes an arts course,
86% take music (vs 64% statewide), and **30% take dance against a state rate of
just 4%**.

```{r fetch-arts, cache = TRUE}
arts_raw <- fetch_arts_enrollment(2024, level = "district") %>%
  filter(district_id == "4900")
stopifnot(nrow(arts_raw) > 0)
```

```{r arts-any-data}
arts_any <- arts_raw %>%
  transmute(grades,
            SOMSD = as.numeric(any_visual_perf_art_district),
            `New Jersey` = as.numeric(any_visual_perf_art_state)) %>%
  tidyr::pivot_longer(c(SOMSD, `New Jersey`), names_to = "geo", values_to = "pct")
stopifnot(nrow(arts_any) > 0)
arts_any
```

```{r arts-grade-chart}
ggplot(arts_any, aes(grades, pct, fill = geo)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.7) +
  geom_text(aes(label = paste0(round(pct), "%")),
            position = position_dodge(width = 0.75), vjust = -0.4, size = 3.3) +
  scale_fill_manual(values = c("SOMSD" = somsd_navy, "New Jersey" = "gray65")) +
  scale_y_continuous(labels = percent_format(scale = 1), limits = c(0, 108)) +
  labs(title = "SOMSD students take the arts at above-state rates",
       subtitle = "Share taking any visual or performing arts course, 2023-24",
       x = NULL, y = "Students taking an arts course", fill = NULL,
       caption = "Source: NJ DOE School Performance Report (Visual & Performing Arts)") +
  theme_nj()
```

```{r arts-disc-data}
arts_disc <- arts_raw %>%
  filter(grades == "Grades 6-8") %>%
  select(music_district, music_state, dance_district, dance_state,
         drama_district, drama_state, vis_arts_district, vis_arts_state) %>%
  mutate(across(everything(), as.numeric)) %>%
  tidyr::pivot_longer(everything(),
                      names_to = c("discipline", "geo"),
                      names_pattern = "(.*)_(district|state)",
                      values_to = "pct") %>%
  mutate(geo = recode(geo, district = "SOMSD", state = "New Jersey"),
         discipline = recode(discipline, music = "Music", dance = "Dance",
                             drama = "Drama/Theater", vis_arts = "Visual arts"))
stopifnot(nrow(arts_disc) > 0)
arts_disc
```

```{r arts-discipline-chart}
ggplot(arts_disc, aes(reorder(discipline, -pct), pct, fill = geo)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.7) +
  geom_text(aes(label = paste0(round(pct), "%")),
            position = position_dodge(width = 0.75), vjust = -0.4, size = 3.3) +
  scale_fill_manual(values = c("SOMSD" = somsd_navy, "New Jersey" = "gray65")) +
  scale_y_continuous(labels = percent_format(scale = 1), limits = c(0, 100)) +
  labs(title = "Middle-school arts: SOMSD vs New Jersey",
       subtitle = "Share of grade 6-8 students taking each art form, 2023-24",
       x = NULL, y = "Students participating", fill = NULL,
       caption = "Source: NJ DOE School Performance Report (Visual & Performing Arts)") +
  theme_nj()
```

## Reproduce this yourself

Every chart above comes from `fetch_tges()`. Once you have a year's tables
(`spending_2025 <- fetch_tges(2025)`), pulling SOMSD's benefits trend is two
lines:

```{r reproduce}
spending_2025[["CSG14"]] %>%
  filter(district_id == "4900", calc_type == "Actuals") %>%
  select(district_name, end_year, `% of Total Salaries`)
```

Swap the district code for any NJ district, or loop `fetch_many_tges(2001:2025)`
for the full history.

```{r session-info}
sessionInfo()
```
