---
title: "SPR Data Dictionary"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SPR Data Dictionary}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 6,
  out.width = "100%",
  message = FALSE,
  warning = FALSE,
  cache = FALSE
)
```

```{r load-packages}
library(njschooldata)
library(dplyr)
library(purrr)
library(ggplot2)
```

## School Performance Reports (SPR) Database

The New Jersey Department of Education's School Performance Reports (SPR) databases contain **63 sheets of data** per year (2017-18 to 2023-24), covering comprehensive school performance metrics.

This document provides a dictionary of available sheets and examples of how to use them.

## Quick Start

### Using the Generic Extractor

```{r generic-extractor, cache = TRUE}
# The generic extractor pulls any sheet by its NJ DOE name.
# level = "district" returns district + state aggregates from one ~30 MB
# workbook; level = "school" returns school-level rows (a much larger download).
ca        <- fetch_spr_data("ChronicAbsenteeism", 2024, level = "district")
teachers  <- fetch_spr_data("TeachersExperience", 2024, level = "district")
grad_dist <- fetch_spr_data("6YrGraduationCohortProfile", 2024, level = "district")
```

```{r generic-extractor-show}
# Every sheet comes back as a tidy data frame with cleaned snake_case columns
dim(ca)
head(names(ca), 9)
```

### Using Convenience Functions

```{r convenience-fns, cache = TRUE}
# Convenience wrappers track the sheet renames NJ DOE makes across years,
# so you do not have to remember which name a given year uses.
ca       <- fetch_chronic_absenteeism(2024, level = "district")
ca_grade <- fetch_absenteeism_by_grade(2024, level = "district")
days     <- fetch_days_absent(2024, level = "district")
```

```{r convenience-fns-show}
# fetch_absenteeism_by_grade() adds a grade_level column (PK, KG, 1-12)
sort(unique(ca_grade$grade_level))
```

## Available SPR Sheets

### Attendance & Discipline

| Sheet Name | Description | Years |
|------------|-------------|-------|
| `ChronicAbsenteeism` | Overall chronic absenteeism rates | 2017-2024 |
| `ChronicAbsenteeismByGrade` | Chronic absenteeism by grade level | 2017-2024 |
| `DaysAbsent` | Days absent statistics (avg, median) | 2017-2024 |
| `ViolenceVandalismHIBSubstanceOf` | Incident counts | 2017-2024 |
| `HIBInvestigations` | HIB investigation details | 2017-2024 |
| `DisciplinaryRemovals` | Suspension/expulsion data | 2017-2024 |

### Staffing & Human Resources

| Sheet Name | Description | Years |
|------------|-------------|-------|
| `TeachersExperience` | Teacher experience levels | 2017-2024 |
| `AdministratorsExperience` | Administrator experience | 2017-2024 |
| `StaffCounts` | Staff counts by category | 2017-2024 |
| `StudentToStaffRatios` | Student-to-staff ratios | 2017-2024 |
| `TeachersAdminsDemographics` | Demographics (race, gender) | 2017-2024 |
| `TeachersAdminsLevelOfEducation` | Educational attainment | 2017-2024 |
| `TeachersAdminsOneYearRetention` | Staff retention rates | 2017-2024 |
| `TeachersBySubjectArea` | Teachers by subject | 2017-2024 |

### College & Career Readiness

| Sheet Name | Description | Years |
|------------|-------------|-------|
| `PSAT-SAT-ACTParticipation` | SAT/ACT/PSAT participation rates | 2017-2024 |
| `PSAT-SAT-ACTPerformance` | SAT/ACT/PSAT score distributions | 2017-2024 |
| `APIBCourseworkPartPerf` | AP/IB performance (detailed) | 2017-2024 |
| `APIBDualEnrPartByStudentGrp` | AP/IB dual enrollment by subgroup | 2017-2024 |
| `APIBCoursesOffered` | AP/IB course offerings | 2017-2024 |
| `CTE_SLEParticipation` | Career/Technical Ed data | 2017-2024 |
| `CTEParticipationByStudentGroup` | CTE by subgroup | 2017-2024 |
| `IndustryValuedCredentialsEarned` | Industry credentials | 2017-2024 |
| `WorkbasedLearningByCareerClust` | Work-based learning | 2017-2024 |
| `Apprenticeship` | Apprenticeship data | 2017-2024 |
| `SealofBiliteracy` | Biliteracy seal earners | 2017-2024 |

### Course Enrollment

| Sheet Name | Description | Years |
|------------|-------------|-------|
| `MathCourseParticipation` | Math course enrollment | 2017-2024 |
| `ScienceCourseParticipation` | Science course enrollment | 2017-2024 |
| `SocStudiesHistoryCourseParticip` | Social studies enrollment | 2017-2024 |
| `WorldLanguagesCourseParticipati` | World language enrollment | 2017-2024 |
| `ComputerScienceCourseParticipat` | CS enrollment | 2017-2024 |
| `VisualAndPerformingArts` | Arts enrollment | 2017-2024 |

### Graduation & Dropout

| Sheet Name | Description | Years |
|------------|-------------|-------|
| `GraduatonRateTrendsProgress` | Graduation rate trends | 2017-2024 |
| `5YrGraduationCohortProfile` | 5-year graduation detailed | 2017-2024 |
| `6YrGraduationCohortProfile` | 6-year graduation detailed | 2021-2024 |
| `FederalGraduationRates` | Federal graduation rates | 2017-2024 |
| `AccountabilityGraduationRates` | ESSA graduation rates | 2017-2024 |
| `GraduationPathways` | Alternate graduation pathways | 2017-2024 |
| `DropoutRateTrends` | Dropout trends | 2017-2024 |

### Accountability

| Sheet Name | Description | Years |
|------------|-------------|-------|
| `ESSAAccountabilityStatus` | ESSA status ratings | 2017-2024 |
| `ESSAAccountabilityProgress` | ESSA progress indicators | 2017-2024 |

### Enrollment & Demographics

| Sheet Name | Description | Years |
|------------|-------------|-------|
| `EnrollmentTrendsbyGrade` | Grade-level enrollment trends | 2017-2024 |
| `EnrollmentTrendsByStudentGroup` | Enrollment by subgroup trends | 2017-2024 |
| `EnrollmentByRacialEthnicGroup` | Detailed racial/ethnic breakdowns | 2017-2024 |
| `PreKAndK-FullDayHalfDay` | Pre-K and K program detail | 2017-2024 |
| `EnrollmentTrendsFullSharedTime` | Full vs shared time | 2017-2024 |
| `EnrollmentByHomeLanguage` | Home language data | 2017-2024 |

### Assessment Performance

| Sheet Name | Description | Years |
|------------|-------------|-------|
| `StudentGrowthTrends` | Student growth percentiles | 2017-2024 |
| `ELAMathPerformanceTrends` | Historical ELA/Math trends | 2017-2024 |
| `NJSLAELAPerformanceTrends` | NJSLA ELA detailed trends | 2017-2024 |
| `NJSLAMathPerformanceTrends` | NJSLA Math detailed trends | 2017-2024 |
| `NJSLAScience` | Science assessment results | 2017-2024 |
| `AlternateAssessmentParticipatio` | Alternate assessment data | 2017-2024 |
| `EnglishLangParticipationPerform` | ELL participation/performance | 2017-2024 |

### Postsecondary

| Sheet Name | Description | Years |
|------------|-------------|-------|
| `PostSecondaryEnrRateSummary` | Postsecondary summary | 2017-2024 |
| `PostsecondaryEnrRatesFall` | Fall enrollment rates | 2017-2024 |
| `PostsecondaryEnrRates16mos` | 16-month enrollment rates | 2017-2024 |

### School Resources

| Sheet Name | Description | Years |
|------------|-------------|-------|
| `SchoolDay` | School day length/schedule | 2017-2024 |
| `DeviceRatios` | Technology/device ratios | 2017-2024 |

### Other

| Sheet Name | Description | Years |
|------------|-------------|-------|
| `Header and Contact` | School contact info | 2017-2024 |
| `Important 2020-2021 Notes` | Data quality notes (2020-21 only) | 2020-2021 |
| `Narrative` | School narrative information | 2017-2024 |

## Usage Examples

### Chronic Absenteeism Analysis

Which districts have the highest chronic absenteeism, and how has the
statewide rate moved since COVID?

```{r ca-analysis-fetch, cache = TRUE}
ca <- fetch_chronic_absenteeism(2024, level = "district")
```

```{r ca-analysis-districts}
# Districts where more than 30% of students are chronically absent
high_absent <- ca %>%
  filter(is_district,
         subgroup == "total population",
         chronically_absent_rate > 30) %>%
  arrange(desc(chronically_absent_rate)) %>%
  select(district_name, chronically_absent_rate)

stopifnot(nrow(high_absent) > 0)
head(high_absent, 8)
```

Compare the statewide rate before and after the pandemic. Wrapping each
year's fetch means a missing year surfaces as a warning, not a silent gap.

```{r ca-trend-fetch, cache = TRUE}
ca_trend <- map_dfr(c(2019, 2022, 2024), function(yr) {
  tryCatch(
    fetch_chronic_absenteeism(yr, level = "district") %>%
      filter(is_state, subgroup == "total population") %>%
      transmute(end_year, chronically_absent_rate),
    error = function(e) {
      warning("chronic absenteeism ", yr, " failed: ", conditionMessage(e))
      NULL
    }
  )
})
```

```{r ca-trend-show}
stopifnot(nrow(ca_trend) == 3)
ca_trend
```

```{r ca-trend-chart}
ggplot(ca_trend, aes(end_year, chronically_absent_rate)) +
  geom_line(color = "#c0392b", linewidth = 1) +
  geom_point(color = "#c0392b", size = 3) +
  geom_text(aes(label = paste0(chronically_absent_rate, "%")), vjust = -1.1) +
  scale_x_continuous(breaks = c(2019, 2022, 2024)) +
  labs(
    title = "NJ chronic absenteeism spiked after COVID and is still elevated",
    subtitle = "Statewide share of students chronically absent (total population)",
    x = NULL, y = "Chronically absent (%)"
  ) +
  theme_minimal()
```

### Teacher Experience Analysis

Which districts rely most on out-of-field teachers? Several SPR numeric
columns arrive as character text, so coerce them before sorting.

```{r teacher-exp-fetch, cache = TRUE}
teachers <- fetch_spr_data("TeachersExperience", 2024, level = "district")
```

```{r teacher-exp-show}
out_of_field <- teachers %>%
  filter(is_district) %>%
  transmute(
    district_name,
    avg_years_exp    = as.numeric(teacher_avg_years_exp_school),
    pct_out_of_field = as.numeric(percentage_of_out_of_field_teachers)
  ) %>%
  arrange(desc(pct_out_of_field))

stopifnot(nrow(out_of_field) > 0)
head(out_of_field, 8)
```

### Course Access Analysis

`APIBCoursesOffered` lists one row per course with enrollment counts. The
counts are stored as character, so coerce before ranking.

```{r course-access-fetch, cache = TRUE}
ap <- fetch_spr_data("APIBCoursesOffered", 2024, level = "district")
```

```{r course-access-show}
# Most-enrolled AP/IB courses statewide
top_ap <- ap %>%
  filter(is_state) %>%
  transmute(course_name, enrolled = as.numeric(student_enroll_count)) %>%
  arrange(desc(enrolled))

stopifnot(nrow(top_ap) > 0)
head(top_ap, 8)
```

### Discipline Analysis

For 2024 the discipline sheet is `DisciplinaryRemovalsByStudgroup`;
`fetch_disciplinary_removals()` picks the right sheet name for each year.
The grouping column is `student_group_grade`, and the sheet reports
suspension and expulsion percentages directly (no enrollment denominator).

```{r discipline-fetch, cache = TRUE}
discipline <- fetch_disciplinary_removals(2024, level = "district")
```

```{r discipline-show}
# Suspension and expulsion rates by student group, statewide
susp_by_group <- discipline %>%
  filter(
    is_state,
    student_group_grade %in% c(
      "Statewide", "Black or African American", "Hispanic", "White",
      "Students with disabilities", "Economically Disadvantaged Students"
    )
  ) %>%
  transmute(
    student_group_grade,
    pct_any_suspension = as.numeric(percent_students_any_suspension),
    pct_expulsions     = as.numeric(percent_students_expulsions)
  ) %>%
  arrange(desc(pct_any_suspension))

stopifnot(nrow(susp_by_group) > 0)
susp_by_group
```

## Data Structure

All SPR data returned by `fetch_spr_data()` includes:

### Standard Identifier Columns
- `end_year` - School year end (e.g., 2024 for 2023-24 school year)
- `county_id`, `county_name` - County code and name
- `district_id`, `district_name` - District code and name
- `school_id`, `school_name` - School code and name (school-level only)

### Aggregation Flags
- `is_state` - State-level aggregation
- `is_county` - County-level aggregation
- `is_district` - District-level aggregation
- `is_school` - School-level aggregation
- `is_charter` - Charter school flag
- `is_charter_sector` - Charter sector aggregation
- `is_allpublic` - All public aggregation

### Sheet-Specific Columns
Each sheet contains additional columns relevant to that data type. Column names are automatically cleaned to snake_case.

## Data Quality Notes

- **2020-2021 Data**: Some data not available due to COVID-19 pandemic disruptions
- **Small Cell Suppression**: Values may be suppressed (NA) for small subgroups to protect privacy
- **Data Validation**: Always check for NA values and data completeness before analysis

## See Also

- [Package documentation](https://almartin82.github.io/njschooldata/)
- [Function reference](https://almartin82.github.io/njschooldata/reference/index.html)
- [Getting started guide](https://almartin82.github.io/njschooldata/articles/getting-started.html)

```{r session-info}
sessionInfo()
```
