---
title: "SPR Data Dictionary"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SPR Data Dictionary}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 6,
  out.width = "100%",
  message = FALSE,
  warning = FALSE
)
```

```{r load-packages}
library(njschooldata)
library(dplyr)
```

## School Performance Reports (SPR) Database

The New Jersey Department of Education's School Performance Reports (SPR) databases contain **63 sheets of data** per year (2017-18 to 2023-24), covering comprehensive school performance metrics.

This document provides a dictionary of available sheets and examples of how to use them.

## Quick Start

### Using the Generic Extractor

```r
# Get chronic absenteeism data
ca <- fetch_spr_data("ChronicAbsenteeism", 2024)

# Get teacher experience data
teachers <- fetch_spr_data("TeachersExperience", 2024)

# Get district-level data
grad_dist <- fetch_spr_data("6YrGraduationCohortProfile", 2024, level = "district")
```

### Using Convenience Functions

```r
# Chronic absenteeism (convenience wrapper)
ca <- fetch_chronic_absenteeism(2024)

# By grade level
ca_grade <- fetch_absenteeism_by_grade(2024)

# Days absent statistics
days <- fetch_days_absent(2024)
```

## Available SPR Sheets

### Attendance & Discipline

| Sheet Name | Description | Years |
|------------|-------------|-------|
| `ChronicAbsenteeism` | Overall chronic absenteeism rates | 2017-2024 |
| `ChronicAbsenteeismByGrade` | Chronic absenteeism by grade level | 2017-2024 |
| `DaysAbsent` | Days absent statistics (avg, median) | 2017-2024 |
| `ViolenceVandalismHIBSubstanceOf` | Incident counts | 2017-2024 |
| `HIBInvestigations` | HIB investigation details | 2017-2024 |
| `DisciplinaryRemovals` | Suspension/expulsion data | 2017-2024 |

### Staffing & Human Resources

| Sheet Name | Description | Years |
|------------|-------------|-------|
| `TeachersExperience` | Teacher experience levels | 2017-2024 |
| `AdministratorsExperience` | Administrator experience | 2017-2024 |
| `StaffCounts` | Staff counts by category | 2017-2024 |
| `StudentToStaffRatios` | Student-to-staff ratios | 2017-2024 |
| `TeachersAdminsDemographics` | Demographics (race, gender) | 2017-2024 |
| `TeachersAdminsLevelOfEducation` | Educational attainment | 2017-2024 |
| `TeachersAdminsOneYearRetention` | Staff retention rates | 2017-2024 |
| `TeachersBySubjectArea` | Teachers by subject | 2017-2024 |

### College & Career Readiness

| Sheet Name | Description | Years |
|------------|-------------|-------|
| `PSAT-SAT-ACTParticipation` | SAT/ACT/PSAT participation rates | 2017-2024 |
| `PSAT-SAT-ACTPerformance` | SAT/ACT/PSAT score distributions | 2017-2024 |
| `APIBCourseworkPartPerf` | AP/IB performance (detailed) | 2017-2024 |
| `APIBDualEnrPartByStudentGrp` | AP/IB dual enrollment by subgroup | 2017-2024 |
| `APIBCoursesOffered` | AP/IB course offerings | 2017-2024 |
| `CTE_SLEParticipation` | Career/Technical Ed data | 2017-2024 |
| `CTEParticipationByStudentGroup` | CTE by subgroup | 2017-2024 |
| `IndustryValuedCredentialsEarned` | Industry credentials | 2017-2024 |
| `WorkbasedLearningByCareerClust` | Work-based learning | 2017-2024 |
| `Apprenticeship` | Apprenticeship data | 2017-2024 |
| `SealofBiliteracy` | Biliteracy seal earners | 2017-2024 |

### Course Enrollment

| Sheet Name | Description | Years |
|------------|-------------|-------|
| `MathCourseParticipation` | Math course enrollment | 2017-2024 |
| `ScienceCourseParticipation` | Science course enrollment | 2017-2024 |
| `SocStudiesHistoryCourseParticip` | Social studies enrollment | 2017-2024 |
| `WorldLanguagesCourseParticipati` | World language enrollment | 2017-2024 |
| `ComputerScienceCourseParticipat` | CS enrollment | 2017-2024 |
| `VisualAndPerformingArts` | Arts enrollment | 2017-2024 |

### Graduation & Dropout

| Sheet Name | Description | Years |
|------------|-------------|-------|
| `GraduatonRateTrendsProgress` | Graduation rate trends | 2017-2024 |
| `5YrGraduationCohortProfile` | 5-year graduation detailed | 2017-2024 |
| `6YrGraduationCohortProfile` | 6-year graduation detailed | 2021-2024 |
| `FederalGraduationRates` | Federal graduation rates | 2017-2024 |
| `AccountabilityGraduationRates` | ESSA graduation rates | 2017-2024 |
| `GraduationPathways` | Alternate graduation pathways | 2017-2024 |
| `DropoutRateTrends` | Dropout trends | 2017-2024 |

### Accountability

| Sheet Name | Description | Years |
|------------|-------------|-------|
| `ESSAAccountabilityStatus` | ESSA status ratings | 2017-2024 |
| `ESSAAccountabilityProgress` | ESSA progress indicators | 2017-2024 |

### Enrollment & Demographics

| Sheet Name | Description | Years |
|------------|-------------|-------|
| `EnrollmentTrendsbyGrade` | Grade-level enrollment trends | 2017-2024 |
| `EnrollmentTrendsByStudentGroup` | Enrollment by subgroup trends | 2017-2024 |
| `EnrollmentByRacialEthnicGroup` | Detailed racial/ethnic breakdowns | 2017-2024 |
| `PreKAndK-FullDayHalfDay` | Pre-K and K program detail | 2017-2024 |
| `EnrollmentTrendsFullSharedTime` | Full vs shared time | 2017-2024 |
| `EnrollmentByHomeLanguage` | Home language data | 2017-2024 |

### Assessment Performance

| Sheet Name | Description | Years |
|------------|-------------|-------|
| `StudentGrowthTrends` | Student growth percentiles | 2017-2024 |
| `ELAMathPerformanceTrends` | Historical ELA/Math trends | 2017-2024 |
| `NJSLAELAPerformanceTrends` | NJSLA ELA detailed trends | 2017-2024 |
| `NJSLAMathPerformanceTrends` | NJSLA Math detailed trends | 2017-2024 |
| `NJSLAScience` | Science assessment results | 2017-2024 |
| `AlternateAssessmentParticipatio` | Alternate assessment data | 2017-2024 |
| `EnglishLangParticipationPerform` | ELL participation/performance | 2017-2024 |

### Postsecondary

| Sheet Name | Description | Years |
|------------|-------------|-------|
| `PostSecondaryEnrRateSummary` | Postsecondary summary | 2017-2024 |
| `PostsecondaryEnrRatesFall` | Fall enrollment rates | 2017-2024 |
| `PostsecondaryEnrRates16mos` | 16-month enrollment rates | 2017-2024 |

### School Resources

| Sheet Name | Description | Years |
|------------|-------------|-------|
| `SchoolDay` | School day length/schedule | 2017-2024 |
| `DeviceRatios` | Technology/device ratios | 2017-2024 |

### Other

| Sheet Name | Description | Years |
|------------|-------------|-------|
| `Header and Contact` | School contact info | 2017-2024 |
| `Important 2020-2021 Notes` | Data quality notes (2020-21 only) | 2020-2021 |
| `Narrative` | School narrative information | 2017-2024 |

## Usage Examples

### Chronic Absenteeism Analysis

```r
# Get school-level chronic absenteeism
ca <- fetch_chronic_absenteeism(2024)

# Filter for high-absenteeism schools
high_absent <- ca %>%
  filter(
    subgroup == "total population",
    chronically_absent_rate > 20
  ) %>%
  arrange(desc(chronically_absent_rate))

# Compare across years
library(purrr)

multi_year <- map_dfr(2018:2024, ~{
  fetch_chronic_absenteeism(.x) %>%
    filter(is_state, subgroup == "total population")
})

# Plot trend
plot(multi_year$end_year, multi_year$chronically_absent_rate,
     type = "b", xlab = "Year", ylab = "Chronic Absenteeism Rate")
```

### Teacher Experience Analysis

```r
# Get teacher experience data
teachers <- fetch_spr_data("TeachersExperience", 2024)

# Analyze new teacher rates
new_teacher_rates <- teachers %>%
  filter(subgroup == "total population") %>%
  mutate(
    new_teacher_rate = as.numeric(`0-3 Years`) /
      as.numeric(`0-3 Years`) + as.numeric(`4-10 Years`) +
      as.numeric(`11-20 Years`) + as.numeric(`20+ Years`)
  ) %>%
  arrange(desc(new_teacher_rate))
```

### Course Access Analysis

```r
# Get math course enrollment
math <- fetch_spr_data("MathCourseParticipation", 2024)

# Check for AP Calculus access
schools_with_calc <- math %>%
  filter(grepl("AP Calculus", math_course, ignore.case = TRUE)) %>%
  pull(school_name) %>%
  unique()

# Compare AP access across districts
ap_access <- fetch_spr_data("APIBCoursesOffered", 2024) %>%
  filter(is_district) %>%
  mutate(
    has_ap_math = grepl("AP Calculus", courses_offered, ignore.case = TRUE),
    has_ap science = grepl("AP.*Science", courses_offered, ignore.case = TRUE)
  )
```

### Discipline Analysis

```r
# Get disciplinary removals data
discipline <- fetch_spr_data("DisciplinaryRemovals", 2024)

# Analyze suspension rates by subgroup
discipline %>%
  filter(subgroup %in% c("black", "white", "hispanic", "total population")) %>%
  group_by(subgroup) %>%
  summarize(
    total_suspended = sum(suspended_count, na.rm = TRUE),
    total_enrolled = sum(enrollment_count, na.rm = TRUE),
    suspension_rate = total_suspended / total_enrolled,
    .groups = "drop"
  )
```

## Data Structure

All SPR data returned by `fetch_spr_data()` includes:

### Standard Identifier Columns
- `end_year` - School year end (e.g., 2024 for 2023-24 school year)
- `county_id`, `county_name` - County code and name
- `district_id`, `district_name` - District code and name
- `school_id`, `school_name` - School code and name (school-level only)

### Aggregation Flags
- `is_state` - State-level aggregation
- `is_county` - County-level aggregation
- `is_district` - District-level aggregation
- `is_school` - School-level aggregation
- `is_charter` - Charter school flag
- `is_charter_sector` - Charter sector aggregation
- `is_allpublic` - All public aggregation

### Sheet-Specific Columns
Each sheet contains additional columns relevant to that data type. Column names are automatically cleaned to snake_case.

## Data Quality Notes

- **2020-2021 Data**: Some data not available due to COVID-19 pandemic disruptions
- **Small Cell Suppression**: Values may be suppressed (NA) for small subgroups to protect privacy
- **Data Validation**: Always check for NA values and data completeness before analysis

## See Also

- [Package documentation](https://almartin82.github.io/njschooldata/)
- [Function reference](https://almartin82.github.io/njschooldata/reference/index.html)
- [Getting started guide](https://almartin82.github.io/njschooldata/articles/getting-started.html)
